### 2017-03-23 VAST_submission/abstract.tex
Added all of the VAST_submission material

\documentclass[journal]{vgtc}                % final (journal style)
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% Please note that the use of figures other than the optional teaser is not permitted on the first page
%% of the journal version.  Figures should begin on the second page and be
%% in CMYK or Grey scale format, otherwise, colour shifting may occur
%% during the printing process.  Papers submitted with figures other than the optional teaser on the
%% first page will be refused. Also, the teaser figure should only have the
%% width of the abstract as the template enforces it.

%% These few lines make a distinction between latex and pdflatex calls and they
%% bring in essential packages for graphics and font handling.
%% Note that due to the \DeclareGraphicsExtensions{} call it is no longer necessary
%% to provide the the path and extension of a graphics file:
%% \includegraphics{diamondrule} is completely sufficient.
%%
\ifpdf%                                % if we use pdflatex
  \pdfoutput=1\relax                   % create PDFs from pdfLaTeX
  \pdfcompresslevel=9                  % PDF Compression
  \pdfoptionpdfminorversion=7          % create PDF 1.7
  \ExecuteOptions{pdftex}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.pdf,.png,.jpg,.jpeg} % for pdflatex we expect .pdf, .png, or .jpg files
\else%                                 % else we use pure latex
  \ExecuteOptions{dvips}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.eps}     % for pure latex we expect eps files
\fi%

%% it is recomended to use ``\autoref{sec:bla}'' instead of ``Fig.~\ref{sec:bla}''
\graphicspath{{figures/}{pictures/}{images/}{./}} % where to search for the images

\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
\usepackage{tabu}                      % only used for the table example
\usepackage{booktabs}                  % only used for the table example
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

%% In preprint mode you may define your own headline.
%\preprinttext{To appear in IEEE Transactions on Visualization and Computer Graphics.}

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{0}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}
%% please declare the paper type of your paper to help reviewers, only shown in review mode
%% choices:
%% * algorithm/technique
%% * application/design study
%% * evaluation
%% * system
%% * theory/model
\vgtcpapertype{please specify}

\title{Interactive Filter and Display of Hillary Clinton's Emails: A Cautionary Tale of Metadata}
\author{Christopher D. Salahub  \& R. Wayne Oldford\\
University of Waterloo\\
Waterloo, Ontario, Canada}
\date{}
\maketitle

\begin{abstract}

We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 Hillary Clinton's emails as provided by Wikileaks.

The visualization focuses on the meta-data of each email, including its senders, receivers, and the timestamp the email appeared on the Clinton server.  An interactive time range slider filters all email and all displays automatically update to changes in the slider.  The main display shows Clinton's most frequent correspondents arranged as nodes of a spoked graph with Clinton at the centre.  Volume determines the thickness of each spoke and high volume determines an inner circle whose spokes are shortened.  Correspondents and their edges are coloured according to whether that email account could be identified as being an approved Federal government account or not.  A second display shows two daily time series: the total number of emails for that day, and the number meeting selection criteria.  A third display shows a scatterplot of the time of day versus the day on which that email appeared.  Scatterplot points are coloured by whether the email was redacted or not.

Other displays add some information beyond metadata.  FOIA exemption codes appear as a selectable list and a barplot shows email counts by FOIA code.  The (stemmed) terms having highest frequency in the displayed email, and those having highest tf-idf are listed in separate displays.  All displays are interactively filtered by time range and selected FOIA codes.

We illustrate how the filtered displays can be used to generate hypotheses and uncover interesting information.  These touch on contentious issues including the handling of classified information, the 2012 attack on the Benghazi U.S. diplomatic compound, and emails apparently missing from those released publicly.

The data are extracted from Wikileaks HTML files, cleaned, and stored in a form useful for interactive exploration.  A local \texttt{R} \texttt{shiny} server provides the interactive displays as a public service online tool to explore and uncover patterns in the meta-data and summary contents of Clinton's email.  Coupled with publicly available sources of information, these interactive tools uncover surprising amounts of information about an individual, especially one holding public office. The ease with which this can be accomplished and shared should serve as a clear warning as to what can be learned about anyone from metadata.

\end{abstract}

The 2016 US Presidential election was one of the most contentious in history.  The existence and possible content of Hillary Clinton's private email server played an significant role in the campaign until the end.

Historically important,  possibly had an impact on the 2016 US presidential election.

Contentious issues:
\begin{itemize}
\item private email server
\item classified documents (outside of state)
\item Sidney Blumenthal
\item Benghazi spin handling of media (Susan Rice)

\item scrubbing of her server (bleachbit)
\item missing email from online email State department
\end{itemize}

Learn:
\begin{itemize}
\item inner circle over time (state or not)
\item spike in email around Libyan revolution
\item gap in the email
\item daily email patterns, server behaviour
\end{itemize}

 Filter:
\begin{itemize}
\item time
\item redacted or not
\item FOIA exemption codes
\end{itemize}

 Content:
\begin{itemize}
\item Term frequency, TFIDF
\end{itemize}

 Tool: 
\begin{itemize}
\item Web-based, interactive filter and display tool
\end{itemize}

Discovery from visualization \& connecting discovery sources

- Nothing on this ... Preparation for Benghazi (security considerations)

- House Oversight and Government Reform Committee (standing committee) 
  Darrell Issa, Chair
     Jason Chaffetz, Chair (Gowdy a member)
  (discovered email server)
    ... 
- House Select Committee on Benghazi (Summer 2014) 
   Trey Gowdy, Chair

\end{document}

### 2017-03-23 VAST_submission/vgtc_journal_latex/paper.tex
Added all of the VAST_submission material

--- 2017-03-23 VAST_submission/abstract.tex
@@
 %% * theory/model
-\vgtcpapertype{please specify}
+\vgtcpapertype{VAST}
 
-\title{Interactive Filter and Display of Hillary Clinton's Emails: A Cautionary Tale of Metadata}
-\author{Christopher D. Salahub  \& R. Wayne Oldford\\
-University of Waterloo\\
-Waterloo, Ontario, Canada}
-\date{}
-\maketitle
+\title{Interactive Filter and Display of Hillary Clinton's Emails: \\A Cautionary Tale of Metadata}
 
-\begin{abstract}
+\author{Christopher D. Salahub and R. Wayne Oldford}
+\authorfooter{
+%% insert punctuation at end of each item
+\item
+ Christopher D. Salahub, University of Waterloo, Canada \\E-mail: csalahub@uwaterloo.ca.
+\item
+ R. Wayne Oldford, University of Waterloo, Canada \\ E-mail: rwoldford@uwaterloo.ca.
+}
 
-We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 Hillary Clinton's emails as provided by Wikileaks.
+%other entries to be set up for journal
+\shortauthortitle{Salahub and Oldford: Filter and Display of Clinton's Emails}
+%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}
+
+\abstract{We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 Hillary Clinton's emails as provided by Wikileaks.
 
@@
 The ease with which this can be accomplished and shared should serve as a clear warning as to what can be learned about anyone from metadata.
+}
 
-\end{abstract}
+%% Keywords that describe your work. Will show as 'Index Terms' in journal
+%% please capitalize first letter and insert punctuation after last keyword
+\keywords{Exploratory data analysis, metadata, text mining, web-scraping, interactive web visualization,  \texttt{R},  \texttt{shiny} }
+
+%% ACM Computing Classification System (CCS). 
+%% See <http://www.acm.org/class/1998/> for details.
+%% The ``\CCScat'' command takes four arguments.
+
+\CCScatlist{ % not used in journal version
+ \CCScat{K.6.1}{Management of Computing and Information Systems}%
+{Project and People Management}{Life Cycle};
+ \CCScat{K.7.m}{The Computing Profession}{Miscellaneous}{Ethics}
+}
+
+%% Uncomment below to include a teaser figure.
+\teaser{
+  \centering
+  \includegraphics[width=\linewidth]{SpiralNetworkOctNov2012}
+  \caption{In the Clouds: Vancouver from Cypress Mountain. Note that the teaser may not be wider than the abstract block.}
+	\label{fig:teaser}
+}
+
+%% Uncomment below to disable the manuscript note
+%\renewcommand{\manuscriptnotetxt}{}
+
+%% Copyright space is enabled by default as required by guidelines.
+%% It is disabled by the 'review' option or via the following command:
+% \nocopyrightspace
+
+\vgtcinsertpkg
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{document}
+%% the only exception to this rule is the \firstsection command
+\firstsection{Introduction}
+
+\maketitle
 
@@
 
+\section{Exposition}
+
+Duis autem vel eum iriure dolor in hendrerit in vulputate velit esse
+molestie consequat, vel illum dolore eu feugiat nulla facilisis at
+vero eros et accumsan et iusto odio dignissim qui blandit praesent
+luptatum zzril delenit augue duis dolore te feugait nulla
+facilisi. Lorem ipsum dolor sit amet, consectetuer adipiscing elit,
+sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna
+aliquam erat volutpat~\cite{Kindlmann:1999:SAG}.
+
+\begin{equation}
+\sum_{j=1}^{z} j = \frac{z(z+1)}{2}
+\end{equation}
+
+Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
+nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
+sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
+rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
+ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
+sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
+dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
+et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
+takimata sanctus est Lorem ipsum dolor sit amet.
+
+\subsection{Lorem ipsum}
+
+Lorem ipsum dolor sit amet (see \autoref{tab:vis_papers}), consetetur sadipscing elitr, sed diam
+nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
+sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
+rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
+ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
+sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
+dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
+et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
+takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
+amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
+invidunt ut labore et dolore magna aliquyam erat, sed diam
+voluptua. At vero eos et accusam et justo duo dolores et ea
+rebum.
+
+\begin{table}[tb]
+  \caption{VIS/VisWeek accepted/presented papers: 1990--2015.}
+  \label{tab:vis_papers}
+  \scriptsize%
+	\centering%
+  \begin{tabu}{%
+	r%
+	*{7}{c}%
+	*{2}{r}%
+	}
+  \toprule
+   year & \rotatebox{90}{Vis/SciVis} &   \rotatebox{90}{SciVis conf} &   \rotatebox{90}{InfoVis} &   \rotatebox{90}{VAST} &   \rotatebox{90}{VAST conf} &   \rotatebox{90}{TVCG @ VIS} &   \rotatebox{90}{CG\&A @ VIS} &   \rotatebox{90}{VIS/VisWeek} \rotatebox{90}{incl.
+TVCG/CG\&A}   &   \rotatebox{90}{VIS/VisWeek} \rotatebox{90}{w/o TVCG/CG\&A}   \\
+  \midrule
+  2015 & 33 & 9 & 38 & 33 & 14 & 17 & 15 & 159 & 127 \\
+  2014 & 34 &   & 45 & 33 & 21 & 20 &   & 153 & 133 \\
+  2013 & 31 &   & 38 & 32 &   & 20 &   & 121 & 101 \\
+  2012 & 42 &   & 44 & 30 &   & 23 &   & 139 & 116 \\
+  2011 & 49 &   & 44 & 26 &   & 20 &   & 139 & 119 \\
+  2010 & 48 &   & 35 & 26 &   &   &   & 109 & 109 \\
+  2009 & 54 &   & 37 & 26 &   &   &   & 117 & 117 \\
+  2008 & 50 &   & 28 & 21 &   &   &   & 99 & 99 \\
+  2007 & 56 &   & 27 & 24 &   &   &   & 107 & 107 \\
+  2006 & 63 &   & 24 & 26 &   &   &   & 113 & 113 \\
+  2005 & 88 &   & 31 &   &   &   &   & 119 & 119 \\
+  2004 & 70 &   & 27 &   &   &   &   & 97 & 97 \\
+  2003 & 74 &   & 29 &   &   &   &   & 103 & 103 \\
+  2002 & 78 &   & 23 &   &   &   &   & 101 & 101 \\
+  2001 & 74 &   & 22 &   &   &   &   & 96 & 96 \\
+  2000 & 73 &   & 20 &   &   &   &   & 93 & 93 \\
+  1999 & 69 &   & 19 &   &   &   &   & 88 & 88 \\
+  1998 & 72 &   & 18 &   &   &   &   & 90 & 90 \\
+  1997 & 72 &   & 16 &   &   &   &   & 88 & 88 \\
+  1996 & 65 &   & 12 &   &   &   &   & 77 & 77 \\
+  1995 & 56 &   & 18 &   &   &   &   & 74 & 74 \\
+  1994 & 53 &   &   &   &   &   &   & 53 & 53 \\
+  1993 & 55 &   &   &   &   &   &   & 55 & 55 \\
+  1992 & 53 &   &   &   &   &   &   & 53 & 53 \\
+  1991 & 50 &   &   &   &   &   &   & 50 & 50 \\
+  1990 & 53 &   &   &   &   &   &   & 53 & 53 \\
+  \midrule
+  \textbf{sum} & \textbf{1515} & \textbf{9} & \textbf{595} & \textbf{277} & \textbf{35} & \textbf{100} & \textbf{15} & \textbf{2546} & \textbf{2431} \\
+  \bottomrule
+  \end{tabu}%
+\end{table}
+
+\subsection{Mezcal Head}
+
+Lorem ipsum dolor sit amet (see \autoref{fig:sample}), consetetur sadipscing elitr, sed diam
+nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
+sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
+rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
+ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
+sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
+dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
+et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
+takimata sanctus est Lorem ipsum dolor sit amet.
+
+\subsubsection{Duis Autem}
+
+Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
+nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
+sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
+rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
+ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
+sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
+dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
+et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
+takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
+amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
+invidunt ut labore et dolore magna aliquyam erat, sed diam
+voluptua. At vero eos et accusam et justo duo dolores et ea
+rebum. Stet clita kasd gubergren, no sea takimata sanctus est. Lorem
+ipsum dolor sit amet.
+
+\begin{figure}[tb]
+ \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
+ \includegraphics[width=\columnwidth]{paper-count-w-2015-new}
+ \caption{A visualization of the data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
+ \label{fig:sample}
+\end{figure}
+
+\subsubsection{Ejector Seat Reservation}
+
+Duis autem~\cite{Lorensen:1987:MCA}\footnote{The algorithm behind
+Marching Cubes \cite{Lorensen:1987:MCA} had already been
+described by Wyvill et al. \cite{Wyvill:1986:DSS} a year
+earlier.} vel eum iriure dolor in hendrerit
+in vulputate velit esse molestie consequat,\footnote{Footnotes
+appear at the bottom of the column.} vel illum dolore eu
+feugiat nulla facilisis at vero eros et accumsan et iusto odio
+dignissim qui blandit praesent luptatum zzril delenit augue duis
+dolore te feugait nulla facilisi. Lorem ipsum dolor sit amet,
+consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt
+ut laoreet dolore magna aliquam erat volutpat.
+
+\paragraph{Confirmed Ejector Seat Reservation}
+
+Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
+suscipit lobortis nisl ut aliquip ex ea commodo
+consequat~\cite{Nielson:1991:TAD}. Duis autem vel eum iriure dolor in
+hendrerit in vulputate velit esse molestie consequat, vel illum dolore
+eu feugiat nulla facilisis at vero eros et accumsan et iusto odio
+dignissim qui blandit praesent luptatum zzril delenit augue duis
+dolore te feugait nulla facilisi.
+
+\paragraph{Rejected Ejector Seat Reservation}
+
+Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
+suscipit lobortis nisl ut aliquip ex ea commodo consequat. Duis autem
+vel eum iriure dolor in hendrerit in vulputate velit esse molestie
+
+\section{Conclusion}
+
+Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
+nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
+sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
+rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
+ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
+sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
+dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
+et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
+takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
+amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
+invidunt ut labore et dolore magna aliquyam erat, sed diam
+voluptua. At vero eos et accusam et justo duo dolores et ea
+rebum.
+
+%% if specified like this the section will be committed in review mode
+\acknowledgments{
+The authors wish to thank A, B, C. This work was supported in part by
+a grant from XYZ.}
+
+%\bibliographystyle{abbrv}
+\bibliographystyle{abbrv-doi}
+%\bibliographystyle{abbrv-doi-narrow}
+%\bibliographystyle{abbrv-doi-hyperref}
+%\bibliographystyle{abbrv-doi-hyperref-narrow}
+
+\bibliography{template}
+
 \end{document}

### 2017-03-23 app.R
Changed Hillary Clinton to Secretary Clinton

@@ -43,11 +43,11 @@ AsSec$Hour6pm <- (AsSec$Hour + 6) %% 24
 
 ### Spiral Network Plot Functions #######################################################
 # filter the data by communications including Clinton
-ClintonCom <- filter(AsSec, 
+ClintonCom <- filter(AsSec,
                      From.name == "Hillary Clinton" | To.name == "Hillary Clinton" |
                        From.name == "H" | To.name == "H")
 # extract names of all the communicated with Clinton
-ClintNet <- sort(table(c(as.character(ClintonCom$To.name), 
+ClintNet <- sort(table(c(as.character(ClintonCom$To.name),
                          as.character(ClintonCom$From.name))),
                  decreasing = TRUE)[-1]
 # extract state mail information, information pertaining to whether an email is .gov or not
@@ -95,7 +95,7 @@ spiralNetPlot <- function(centralNode = "Hillary Clinton", wgtTbl = NA,
     radPoslvl2 <- c(outer(radPoslvl1, shift*(1:numFull), FUN = function(x,y) x + y))
     # add in the rest
     extraPos <- if(numRot-numFull == 0) NULL else(radPoslvl1 +
-                                                    numRot*shift)[1:(length(wgtTbl) - 
+                                                    numRot*shift)[1:(length(wgtTbl) -
                                                                        impDiffs[1]*(numFull+1))]
     # combine
     if (numFull != 0) radPoslvl2 <- c(radPoslvl2, extraPos)
@@ -111,11 +111,11 @@ spiralNetPlot <- function(centralNode = "Hillary Clinton", wgtTbl = NA,
                  gp = gpar(lwd = 1+wgtTbl[ii]/max(wgtTbl)*6,
                            col = adjustcolor(cols[ii]), alpha.f = 0.8))
     }
-    grid.circle(x = c(lvl1x, lvl2x), y = c(lvl1y, lvl2y), r = 0.01, 
+    grid.circle(x = c(lvl1x, lvl2x), y = c(lvl1y, lvl2y), r = 0.01,
                 gp = gpar(col = adjustcolor(cols, alpha.f = 0.8),
                           fill = adjustcolor(cols, alpha.f = 0.8)))
     # place the central node
-    grid.circle(x = 0.5, y = 0.5, r = 0.01, 
+    grid.circle(x = 0.5, y = 0.5, r = 0.01,
                 gp = gpar(col = adjustcolor(col = "firebrick", alpha.f = 1),
[…]

### 2017-03-23 app.R
Removed send labels; Titled Spiral Network Inner Circle by Volume

@@ -295,13 +295,13 @@ server <- function(input, output) {
      if (adjVar()) {
        timevalues <- selDat()$Hour6pm*60 + selDat()$Minutes
        xlab <- "Adjusted Date (dd/mm/yy)"
-       main <- "Adjusted Email Sending Times"
+       main <- "Adjusted Email Times"
        labelset <- c('22:00', '02:00', '06:00', '10:00', '14:00')
      }
      else {
        timevalues <- selDat()$Hour*60 + selDat()$Minutes
        xlab <- "Date (dd/mm/yy)"
-       main <- "Email Sending Times"
+       main <- "Email Sent/Received Times"
        labelset <- c("04:00", "08:00", "12:00", "16:00", "20:00")
      }
      # plot dates based on the dates provided from the slider
@@ -321,8 +321,8 @@ server <- function(input, output) {
    # next display the time series of emails sent by day
    output$DaySum <- renderPlot({
      plot(x = selDays(), y = selCounts(), type = 'l', xlab = 'Date (dd/mm/yy)',
-          xaxt = "n", ylab = 'Number of Emails Sent', pch = 19,
-          main = "Number of Emails Sent by Date",
+          xaxt = "n", ylab = 'Number of Emails', pch = 19,
+          main = "Number of Emails by Date",
           ylim = extendrange(AScounts), col = adjustcolor("black", alpha.f = 0.6))
      # add the schedule if it has been selected
      if (dispSched()) {
@@ -350,7 +350,7 @@ server <- function(input, output) {
      spiralNetPlot2(wgtTbl = sort(table(c(as.character(SpirDat()$To.name),
                                 as.character(SpirDat()$From.name))),
                         decreasing = TRUE)[-1],
-                   title = "Spiral Network Plot of Clinton's Communications")
+                   title = "Inner Circle by Volume of Communication")
    })
    # display the top twenty tfidf terms
    output$tfidf <- renderText(paste(

### 2017-03-23 VAST_submission/vgtc_journal_latex/paper.tex
Updated teaser

@@
   \centering
-  \includegraphics[width=\linewidth]{SpiralNetworkOctNov2012}
-  \caption{In the Clouds: Vancouver from Cypress Mountain. Note that the teaser may not be wider than the abstract block.}
+  \begin{tabular}{cc}
+  \includegraphics[width=0.25\textheight]{DailySept11ToNov152012_All} &
+  \includegraphics[width=0.25\textheight]{InnerCircleSept11ToNov152012_All} \\
+  &\\
+  {\small (a) Missing data in November.} & {\small (b) Inner Circle: Correspondent info by colour; volume by thickness.}
+  \end{tabular}
+  \caption{Metadata: Time, date, correspondents, volume.  Filtered: Sept. 11 to Nov. 15, 2012.  
+  %Source: \href{https://wikileaks.org/clinton-emails/}{\texttt{https://wikileaks.org/clinton-emails/}}
+  }
+
 	\label{fig:teaser}

### 2017-03-24 app.R
Added further email identification functionality to the app, and removed the outdated SpiralNetworkPlot as it is already in the EmailFunctions file and is simply cluttering this app code

@@ -10,6 +10,7 @@ library(shinyBS)
 
 ### Pre-loaded data #####################################################################
 pal <- c("steelblue", "firebrick")
+pal2 <- c("gray50","firebrick", "steelblue")
 EmailData <- read.csv('ClintonEmailData.csv')
 ThompTime <- read.csv('ThompsonTimeline.csv')
 ForSched <- read.csv('ForeignSchedule.csv')
@@ -62,82 +63,9 @@ for (name in allnames) {
   Mail <- filter(ClintonCom, To.name == name | From.name == name)
   state <- any(grepl(".gov", c(as.character(Mail$To.Add), as.character(Mail$To.name),
                                as.character(Mail$From.name), as.character(Mail$From.Add))))
-  stateMail[which(allnames == name)] <- as.numeric(state)
-}
-# write the function to generate the spiral network plot
-# THIS IS AN OUTDATED VERSION, SEE SpiralNetPlot2
-spiralNetPlot <- function(centralNode = "Hillary Clinton", wgtTbl = NA,
-                          levelNum = 2, nodeNum = 20, title = "Title") {
-  # check the wgtTbl status
-  if (!all(is.na(wgtTbl))) {
-    # trim wgtTbl
-    if (length(wgtTbl) > nodeNum) wgtTbl <- wgtTbl[1:nodeNum]
-    # start by generating state mail colouring
-    statemail <- stateMail[names(stateMail) %in% names(wgtTbl)]
-    statemail <- statemail[match(names(wgtTbl), names(statemail))]
-    if (length(statemail) == 0) cols <- "white"
-    else cols <- c("firebrick", "steelblue")[statemail + 1]
-    # generate a new page
-    grid.newpage()
-    # define radial units
-    radunit <- 0.45/levelNum
-    # determine inner level names
-    diffs <- diff(wgtTbl)
-    # select the largest
-    impDiffs <- order(diffs)[1:(levelNum-1)]
-    # get the radial positions of the first level
-    radPoslvl1 <- seq(from = 0, to = 2*pi, length.out = impDiffs[1] + 1)[1:impDiffs[1]]
-    # add the second level, see how much space has to be left in between axes
-    numRot <- ceiling((length(wgtTbl) - impDiffs[1])/impDiffs[1])
-    numFull <- floor((length(wgtTbl) - impDiffs[1])/impDiffs[1])
-    # determine the major positions
[…]

### 2017-03-24 EmailFunctions.R
Changed keyword function to ignore numeric data

@@ -527,6 +527,7 @@ get_keywords <- function(Content, stem = TRUE, table = TRUE) {
   corpus <- str_replace_all(corpus, "(?<=[0-9])\\,(?=[0-9])", "")
   # now safely remove everything else
   corpus <- str_replace_all(corpus, "[^0-9a-zA-Z\\s]", " ")
+  corpus <- str_replace_all(corpus, "[0-9]", "")
   # replace all duplicate spaces
   corpus <- str_replace_all(corpus, "\\s+", " ")
   # split by singular spaces

### 2017-03-24 EmailFunctions.R
Improved the automation of a couple of column insertions

@@ -557,7 +557,16 @@ ASmat <- matrix(unlist(ASmat), ncol = length(relKey), byrow = TRUE)
 ASmat <- read.csv('TermDocumentMatrix.csv')
 # generate a tfidf matrix
 tfIdf <- apply(ASmat, 2, function(col) col/max(col) * log(length(col)/sum(col != 0)))
-colnames(tfIdf) <- relKey
+ASmat <- cbind(AsSec$ID, AsSec$Date, 
+               AsSec[,c("B1","B2","B3","B4","B5","B6","B7","B8","B9","None")],
+               ASmat)
+tfIdf <- cbind(AsSec$ID, AsSec$Date, 
+               AsSec[,c("B1","B2","B3","B4","B5","B6","B7","B8","B9","None")],
+               tfIdf)
+colnames(tfIdf) <- c("ID","Date","B1","B2","B3","B4","B5","B6","B7","B8","B9","None",
+                     relKey)
+colnames(ASmat) <- c("ID","Date","B1","B2","B3","B4","B5","B6","B7","B8","B9","None",
+                     relKey)
 
 ### Frequency and Email Network Plots with Sliders ######################################
 # let's do some processing by days and extract relevant information

### 2017-03-25 EmailFunctions.R
Changed the keyword function to improve its cleaning ability

@@ -520,6 +520,10 @@ get_keywords <- function(Content, stem = TRUE, table = TRUE) {
   # make everything lowercase for reference
   corpus <- tolower(corpus)
   # start by removing a number of problematic terms
+  corpus <- str_replace_all(corpus, "\\&lt;|\\&gt;|</span>|</div>", " ")
+  corpus <- str_replace_all(corpus, "unclassified [a-z0-9][.]*[a-z0-9][.]* department of state case no[.]* [a-z0-9]+ doc no[.]* [a-z0-9]+ date: [0-9/]+ | unclassified [a-z0-9][.]*[a-z0-9][.]* department of state case no[.]* [a-z0-9\\-]+ doc no[.]* [a-z0-9]+",
+                            " ")
+  corpus <- str_replace_all(corpus, "release\\s*in\\s*part|release\\s*in\\s*full", " ")
   corpus <- str_replace_all(corpus, "u.s.|u.s.a.", "usa")
   # remove all non-alphanumeric characters
   # first remove the contractions and commas used in number display
@@ -554,7 +558,7 @@ ASmat <- lapply(lapply(AsSec$Content, function(e) get_keywords(e, table = FALSE)
                 function(email) sapply(relKey, function(kw) sum(kw == email)))
 ASmat <- matrix(unlist(ASmat), ncol = length(relKey), byrow = TRUE)
 # load the term-document matrix
-ASmat <- read.csv('TermDocumentMatrix.csv')
+# ASmat <- read.csv('TermDocumentMatrix.csv')
 # generate a tfidf matrix
 tfIdf <- apply(ASmat, 2, function(col) col/max(col) * log(length(col)/sum(col != 0)))
 ASmat <- cbind(AsSec$ID, AsSec$Date, 

### 2017-03-25 VAST_submission/vgtc_journal_latex/paper.tex
Paper update

@@
 The existence and possible content of Hillary Clinton's private email server played an significant role in the campaign until the end.
+Despite their significant nature, very few of the individuals commenting on the significance of this server had spent any significant time viewing this content.
+This is no fault of theirs, the officially released data has previously only been stored in databases on the United States' State Department Website (see https://foia.state.gov/Learn/New.aspx) and Wikileaks (see https://wikileaks.org/clinton-emails/) in individually searchable form.
+Furthermore, the email data is stored in the inconvenient format of individual PDF documents, and in the case of Wikileaks a slightly more convenient but still cumbersome HTML representation based on the official PDF content.
+This granularity and separation of individual emails on different webpages prevents a great deal interesting analysis of the data, including any aggregate analyses of patterns in content or metadata.
+As such, the emails are incredibly misunderstood documents, and current searches for patterns and points of interest have relied upon searching for terms of interest and sifting through the results of the search.
+Such a method is not only tedious but problematic from a statistical point of view, as approaching any data set with a specific hypothesis before exploring its structure and patterns in abstract leads to biased conclusions and poor analysis.
+In the interest of simplifying the general exploration of this important data set, the Wikileaks HTML email versions were extracted and a series of visualizations were generated around the metadata included in these emails.
+An interactive web application in \texttt{R} \texttt{shiny} was constructed utilizing these visualizations and the extracted data.
+Incorporated in this app was a short article explaining the use of the app, providing links to background information, and outlining some of the findings of the more interesting trends observed.
+This framework provides any individual with interest the ability to generate hypotheses by exploring the data without necessarily searching for any particular conclusion.
+The methods used to generate this
 

### 2017-03-25 VAST_submission/vgtc_journal_latex/paper.tex
Paper update

@@
 This framework provides any individual with interest the ability to generate hypotheses by exploring the data without necessarily searching for any particular conclusion.
-The methods used to generate this
+
+In generating and exploring this data, a considerable amount of respect is gained for the power of the often overlooked metadata of emails, especially with public figures.
+Simply having access to the network of communications, simple indicators of identifiable features of the individuals communicating, and times sent is enough to generate interesting questions which simple Google searches of important events and days can shed a great deal of light on.
+Leveraging data beyond this metadata, including the Freedom of Information Act (FOIA) exemption codes (see https://vault.fbi.gov/explanation-of-exemptions) used to redact the emails, only bolsters this simple metadata driven investigation.
+Metadata is often viewed as less significant or somehow less intrusive data, but the exploration performed in this paper demonstrated, at the very least to the authors, what a powerful tool metadata is for generating hypotheses about data.
+Given the lack of privacy present in the modern internet age and the amount of information most individuals make publicly available, there is a clear warning here about the utility of metadata for bad or for good.
+For a public official like Hillary Clinton, the data become even more interesting due to a number of contentious issues which arose during her tenure as the Secretary of State, independent of and as a result of her use of a private server.
+
+\section{Data Extraction}
+
+Before any of the analysis could begin, the emails had to be converted from the individual PDF and HTML form into a more easily used data structure.
+The choice was made to use the Wikileaks HTML emails as the source for data extraction.
+This choice was made primarily due to the relative ease of loading the raw HTML pages on the Wikileaks database due to their regular method of storage (the email with ID <id> is stored at https://wikileaks.org/clinton-emails/emailid/<id>) and ease of loading.
+These qualities made the HTML emails amenable to programmatic extraction, which was completed using a host of web-scraping and string parsing tools in \texttt{R}.
+In particular the packages \texttt{RCurl} and \texttt{XML} proved invaluable to extract the raw HTML pages for each email, and the packages \texttt{tm}, \texttt{stringr}, and \texttt{SnowballC} were indispensible to process and clean the raw HTML into a useable form.
+After downloading the raw HTML and extracting the data of interest, the resulting data was stored in csv files to provide the flexibility to load the data into any architecture or analysis tool desired.
+
+\section{Shiny App}
 
@@
 \item gap in the email
-\item daily email patterns, server behaviour
+\item daily email patterns, server behaviour (daylight savings time)
 \end{itemize}

### 2017-03-25 VAST_submission/vgtc_journal_latex/paper.tex
Paper update
More paper writing

@@
 
+Unfortunately, this data is not perfect.
+The PDF files released by the state department do not include the typical email server header with information about the time stamp inforamtion, the addresses of all involved, and other useful and regular information.
+Instead, these files appear more similar to screen captures of the emails, providing only the visual display a user would see without the useful information used to generate that display.
+As a result, the email addresses are occasionally not included, and there were initially concerns over the time zones used to generate the time stamp metadata.
+Furthermore, the HTML data on Wikileaks is frequently irregular in form, and although work has been done to make the cleaning and extraction functions as general as possible, there are still certainly fringe cases not considered which occur frequently enough to have an impact.
+It is also important to keep in mind the fact that the data have already been filtered twice.
+First by the emails chosen to be released by Clinton, and next by the emails and email sections which the State Department chose not to redact.
+Indeed, even the metadata can be affected by this redaction, as a number of individuals have their email addresses redacted in all communications, preventing certain analysis from being performed.
+
 \section{Shiny App}
+
+The app (HOST ADDRESS) provides the ability for users to place custom filters on the time frame of server time stamps and classification tags of the emails and view how a series of three displays change.
+The first of these is a spoked network graph of all corresponents with Clinton at the centre.
+The spokes are separated into two groups by the radial length of the spokes, these groups are determined automatically by the largest difference in volume of correspondence present in the data between the ordered counts of number of emails sent between individuals.
+As well, the width of the edges of the graph is determined by the volume of correspondence which occurred between Clinton and the correspondent in either direction.
+Finally, all edges and points are coloured according to a simple scheme.
+If the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, they are coloured blue.
+Individuals with email addresses which are identifiably not hosted at a '.gov' domain are coloured red.
+Finally, individuals which cannot have their email host domain identified are coloured grey.
+It is important to note that for those individuals with grey spokes, the reason the email is not identifiable is the removal of the email by the state department.
+Such a redaction is not a guarantee that the email said individual used is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain.
 

### 2017-03-25 app.R
Reversed the axis directions for the time of day plot as suggested earlier

@@ -240,7 +240,8 @@ server <- function(input, output) {
           ylab = 'Time Sent', yaxt = 'n', xaxt = 'n', pch = 19,
           main = main,
           col = adjustcolor(pal[as.numeric(selDat()$Redacted)+1], alpha.f = 0.5),
-          cex = 0.25, ylim = extendrange(c(0,1440)), xlim = c(input$range[1], input$range[2]))
+          cex = 0.25, ylim = c(1440,0) + c(0.05,-0.05)*1440, 
+          xlim = c(input$range[1], input$range[2]))
      axis(side = 2, at = c(240, 480, 720, 960, 1200),
           labels = labelset)
      axis.Date(side = 1, as.chron(c(input$range[1], input$range[2])), format = "%d/%m/%y")

### 2017-03-25 VAST_submission/vgtc_journal_latex/paper.tex
More paper changes

@@
 Such a redaction is not a guarantee that the email said individual used is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain.
+Still, such speculation cannot be conclusively made, and so these emails are simply reported as they are, unidentifiable.
+
+In addition to the spoked network graph, a time series plot of daily email volume is displayed.
+This plot has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day.
+This allows users to view which days have an exceptionally high or low number of classified conversations.
+Returning to the motivation of hypothesis generation, referencing world events on any days of interest provides a number of interesting insights on how public figures like Clinton react and respond to world events.
+
+For an even more detailed view of exactly when emails are sent and what proportion are classified, a third plot of email sending times by day is included in the web application.
+This plot takes advantage of the time-stamps on every email and then plots the emails as points with times on the vertical axis and the day sent on the horizontal axis.
+The points in this scatterplot are then coloured according to whether the corresponding email was redacted, in which case the point is coloured red, or released in full, in which case the point is coloured blue.
+Alpha blending and transparency of these points was utilized to make the patterns present more obvious and prevent overplotting from obscuring any interesting results.
+
+Finally, a barplot of the FOIA exemption codes used in the selected data is displayed.
+It is important to notice that this barplot is not trivialized by the possibility of selecting individual FOIA exemption codes, as many of the emails in the data set are exempted on a number of grounds, and so have numerous FOIA exemption codes.
+Thus, this barplot is not only interesting when all codes are selected, but can be used to see which codes commonly co-occur in documents.
 

### 2017-03-25 app.R
Fixing a labelling mistake

@@ -237,7 +237,7 @@ server <- function(input, output) {
      # plot dates based on the dates provided from the slider
      plot(x = as.chron(floor(selDat()$Date)),
           y = timevalues, xlab = xlab,
-          ylab = 'Time Sent', yaxt = 'n', xaxt = 'n', pch = 19,
+          ylab = 'Time Sent/Received', yaxt = 'n', xaxt = 'n', pch = 19,
           main = main,
           col = adjustcolor(pal[as.numeric(selDat()$Redacted)+1], alpha.f = 0.5),
           cex = 0.25, ylim = c(1440,0) + c(0.05,-0.05)*1440, 

### 2017-03-25 app.R
Fixed a section of code which was too general and would lead to improper specification

@@ -60,11 +60,12 @@ stateMail <- rep(0, length(allnames))
 names(stateMail) <- allnames
 # for each name identify if it is associated with a .gov email anywhere in the data
 for (name in allnames) {
-  Mail <- filter(ClintonCom, To.name == name | From.name == name)
-  state <- any(grepl(".gov", c(as.character(Mail$To.Add), as.character(Mail$To.name),
-                               as.character(Mail$From.name), as.character(Mail$From.Add))))
-  notState <- any(grepl("[@.]", c(as.character(Mail$To.Add), as.character(Mail$To.name),
-                                   as.character(Mail$From.name), as.character(Mail$From.Add))))
+  toMail <- filter(ClintonCom, To.name == name)
+  fromMail <- filter(ClintonCom, From.name == name)
+  state <- any(grepl(".gov", c(as.character(toMail$To.Add), as.character(toMail$To.name),
+                               as.character(fromMail$From.name), as.character(fromMail$From.Add))))
+  notState <- any(grepl("[@.]", c(as.character(toMail$To.Add), as.character(toMail$To.name),
+                                  as.character(fromMail$From.name), as.character(fromMail$From.Add))))
   stateMail[which(allnames == name)] <- as.numeric(state) + as.numeric(notState)
 }
 

### 2017-03-25 EmailFunctions.R
Fixed a section of code which was too general and would lead to improper specification

@@ -845,9 +845,10 @@ allnames <- levels(as.factor(c(levels(ClintonCom$To.name),
 stateMail <- rep(0, length(allnames))
 names(stateMail) <- allnames
 for (name in allnames) {
-  Mail <- filter(ClintonCom, To.name == name | From.name == name)
-  state <- any(grepl(".gov", c(as.character(Mail$To.Add), as.character(Mail$To.name),
-                               as.character(Mail$From.name), as.character(Mail$From.Add))))
+  toMail <- filter(ClintonCom, To.name == name)
+  fromMail <- filter(ClintonCom, From.name == name)
+  state <- any(grepl(".gov", c(as.character(toMail$To.Add), as.character(toMail$To.name),
+                               as.character(fromMail$From.name), as.character(fromMail$From.Add))))
   stateMail[which(allnames == name)] <- as.numeric(state)
 }
 # this is my double sided slider hack

### 2017-03-25 VAST_submission/vgtc_journal_latex/paper.tex
More paper updates

@@
 Returning to the motivation of hypothesis generation, referencing world events on any days of interest provides a number of interesting insights on how public figures like Clinton react and respond to world events.
+To augment this graph, Clinton's official foreign visit schedule (https://history.state.gov/departmenthistory/travels/secretary/clinton-hillary-rodham) was extracted, cleaned, and incorporated into the application.
+This coloured time segments during which Clinton was travelling with blue rectangles, highlighting her pattern of foreign visits and allowing a comparison between these and the daily email sending patterns.
 
@@
 Thus, this barplot is not only interesting when all codes are selected, but can be used to see which codes commonly co-occur in documents.
+
+\section{Observations}
+
+In designing, creating, testing, and playing with the app, a number of interesting observations were made.
+These are summarized below, and some address contentious issues in regards to Clinton's emails and a number of significant events which occurred during her tenure as the United States Secretary of State.
+
+\subsection{Daily Email Patterns}
+
+\includegraphics{SentReceivedEntireTenure}
+
+The most striking thing noticed about the email sending times was the regularity of the pattern.
+Across the entirety of Clinton's four year tenure as the United States Secretary of State, email communication is densest between midnight and four in the morning, and there is a conspicuous gap in communication between roughly four in the afternoon and ten at night.
+This pattern is quite surprising, as it suggests that Clinton and her staff are most actively communicating late at night, rather than at the start of the regular work day around seven or eight in the morning.
+It would be expected that individuals would be sleeping late at night, and as a result that the density of emails would be relatively low.
+Rather, the data suggests that the only real break Clinton and her staff take from email communication is in the early evening, when many of them doubtless have official functions to attend.
+Regardless of the individual daily variation, the consistency of this pattern suggests that Clinton and her staff slept at strange times, and perhaps very irregularly, or were at least frequently interrupted by emails while sleeping.
+It is also possible these late night emails are more common because they are the ones Clinton chose to release, after all, it has been observed Clinton used the service Bleachbit to delete many old emails from her server (http://www.politico.com/story/2016/08/hillary-clinton-emails-bleachbit-227425, http://money.cnn.com/2016/08/26/technology/hillary-clinton-bleachbit/) and the organizational emails sent late at night to coordinate the following day may have simply been overlooked or maintained deliberately.
+Either way, the presence of such high volumes late at night tells a great deal about the irregularity of sleep that must have occurred among Clinton's staff during her tenure.
+
+Another interesting pattern in the data is that of the two alternating modes at two and three in the morning.
+Immediately, hypothesizing that the cause of these modes is some regular server function is obvious due to the precision and consistency of the pattern.
+The alternation in the pattern was initially vexing, but comparing the shifts in the mode to the dates of daylight savings time in North America for the time period found a perfect match.
+It would seem that the alternation, then, is not deliberate, but rather a consequence of the one hour time shift that occurs in March and November every year in North America.
+Still, this observation does not reveal much about the content of the emails sent at this time, or why they are so regular.
+To better explore these emails, the app was run on a new set of data which was filtered to only contain the emails sent at two and three in the morning.
+Of the 32,359 emails sent during Clinton's tenure in this data set, 4,020 or roughly one eighth were sent at exactly two or three in the morning.
 

### 2017-03-25 app.R
Changed colouring to make it more accessible

@@ -10,7 +10,7 @@ library(shinyBS)
 
 ### Pre-loaded data #####################################################################
 pal <- c("steelblue", "firebrick")
-pal2 <- c("gray50","firebrick", "steelblue")
+pal2 <- c("darkorange","firebrick", "steelblue")
 EmailData <- read.csv('ClintonEmailData.csv')
 ThompTime <- read.csv('ThompsonTimeline.csv')
 ForSched <- read.csv('ForeignSchedule.csv')
@@ -206,19 +206,19 @@ server <- function(input, output) {
   DateRange <- reactive(paste(chron(c(input$range[1], input$range[2]),
                                     format = "day mon year"),
                               collapse = " - "))
-  selDat <- reactive(filter(AsSec[apply(AsSec[,input$ClassFilter],1,any),],
+  selDat <- reactive(filter(AsSec[apply(AsSec[,input$ClassFilter, drop = FALSE],1,any),],
                             Date < input$range[2] + 1 & Date >= input$range[1]))
   selDays <- reactive(ASDays[ASDays < input$range[2] + 1 & ASDays >= input$range[1]])
   selCounts <- reactive(AScounts[ASDays < input$range[2] + 1 & ASDays >= input$range[1]])
-  SpirDat <- reactive(filter(ClintonCom[apply(ClintonCom[,input$ClassFilter],1,any),],
+  SpirDat <- reactive(filter(ClintonCom[apply(ClintonCom[,input$ClassFilter, drop = FALSE],1,any),],
                               Date < input$range[2] + 1 & Date >= input$range[1]))
   Sched <- reactive(filter(ForSched, (StartDate >= input$range[1] &
                                         StartDate < input$range[2]) |
                              (EndDate < input$range[2] & EndDate >= input$range[1])))
   dispSched <- reactive(input$Schedule)
-  tfIdf <- reactive(filter(TfIdf[apply(TfIdf[,input$ClassFilter],1,any),],
+  tfIdf <- reactive(filter(TfIdf[apply(TfIdf[,input$ClassFilter, drop = FALSE],1,any),],
                            Date < input$range[2] + 1 & Date >= input$range[1])[,-(1:12)])
-  freq <- reactive(filter(Freq[apply(Freq[,input$ClassFilter],1,any),],
+  freq <- reactive(filter(Freq[apply(Freq[,input$ClassFilter, drop = FALSE],1,any),],
[…]

### 2017-03-25 EmailFunctions.R
Improved cleanup to fix wild &lt; etc.

@@ -20,10 +20,16 @@ infoExtractor <- function(emails, ids, includeRaw = FALSE) {
   # extract the recipient information from the Wikileaks header
   to <- str_extract(emails, "(?<=(To\\: <span title=\"Original: )).+(?=</span>)")
   nameTo <- str_extract(to, "(?<=(\">)).+")
+  nameTo <- str_replace_all(nameTo, "&lt.", "\\(")
+  nameTo <- str_replace_all(nameTo, "&gt.", "\\)")
+  nameTo <- str_replace_all(nameTo, "&quot.", "")
   addressTo <- str_extract(to, "[\\.a-zA-Z0-9\\_]+@[\\.a-zA-Z0-9]+")
   # extract the sender information from the Wikileaks header
   from <- str_extract(emails, "(?<=(From\\: <span title=\"Original: )).+(?=</span>)")
   nameFrom <- str_extract(from, "(?<=(\">)).+")
+  nameFrom <- str_replace_all(nameFrom, "&lt.", "\\(")
+  nameFrom <- str_replace_all(nameFrom, "&gt.", "\\)")
+  nameFrom <- str_replace_all(nameFrom, "&quot.", "")
   addressFrom <- str_extract(from, "[\\.a-zA-Z0-9\\_]+@[\\.a-zA-Z0-9]+")
   # also extract the forwarding contact chain using generic email matching on "@"
   forwards <- str_extract_all(emails, "[a-zA-Z0-9]+\\@[a-zA-Z0-9\\.]+|To:|From:")

### 2017-03-26 app.R
Made filtering more efficient and added the ability to filter by emails sent to and from Clinton

@@ -16,9 +16,6 @@ ThompTime <- read.csv('ThompsonTimeline.csv')
 ForSched <- read.csv('ForeignSchedule.csv')
 Freq <- read.csv("TermDocumentMatrix.csv")
 TfIdf <- read.csv("TFIDF.csv")
-# commented out daily reads, the most recent version has much greater selection power
-#DailyFreq <- as.matrix(read.csv('DailyFrequency.csv'))
-#DailyTfIdf <- as.matrix(read.csv('TFIDFbyDay.csv'))
 
 ### Filter and prepare the data from time as secretary of state #########################
 # select all emails sent while Secretary
@@ -125,7 +122,7 @@ spiralNetPlot2 <- function(centralNode = "Hillary Clinton", wgtTbl = NA,
               just = "left", gp = gpar(face = 2))
     grid.text("Red - Identifiably Not .gov", x = 0.98, y = 0.02,
               just = "right", gp = gpar(face = 2))
-    grid.text("Grey - Unidentifiable", x = 0.5, y = 0.02,
+    grid.text("Orange - Unidentifiable", x = 0.5, y = 0.02,
               just = "centre", gp = gpar(face = 2))
   }
   else  {
@@ -177,6 +174,9 @@ ui <- fluidPage(
             fluidRow(selectInput("AdjVar", "Daily Reference Point: ", c("0:00", "18:00"),
                                  selected = "0:00")),
             fluidRow(checkboxInput("Schedule", "Display Foreign Travel Schedule")),
+            fluidRow(selectInput("ToFromFilter", "Show Emails: ", 
+                                 c("From Clinton", "To Clinton", "All Emails"),
+                                 selected = "All Emails", multiple = FALSE)),
             fluidRow(selectInput("ClassFilter", "FOIA Codes: ",
                                  c("None", "B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", "B9"),
                                  selected = c("None", "B1", "B2", "B3", "B4", "B5", "B6", "B7",
@@ -206,41 +206,55 @@ server <- function(input, output) {
   DateRange <- reactive(paste(chron(c(input$range[1], input$range[2]),
                                     format = "day mon year"),
                               collapse = " - "))
-  selDat <- reactive(filter(AsSec[apply(AsSec[,input$ClassFilter, drop = FALSE],1,any),],
[…]

### 2017-03-26 VAST_submission/vgtc_journal_latex/paper.tex
First cut; sectioned

--- 2017-03-25 VAST_submission/vgtc_journal_latex/paper.tex
@@
 %\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
+
+% Colors
+\usepackage[
+hyperref,% for working together with hyperref
+table% for coloring matrix entries
+]{xcolor}% color package; load before tocstyle
 
@@
 
+% Commands with arguments
+\newcommand*{\TODO}[1]{\textcolor{red}{TODO #1}}
+% See http://tex.stackexchange.com/questions/251460/how-to-put-symbols-of-equal-size-on-top-of-each-other?noredirect=1#comment599930_251460
+
 \begin{document}
@@
 \firstsection{Introduction}
+\maketitle
+The 2016 U.S.
+Presidential election was one of the most contentious in history.
+The existence and possible content of Hillary Clinton's private email server dogged former Secretary Clinton's bid for the U.S.
+presidency and was doubtless a contributing factor to her surprising defeat by Donald Trump.
 
-\maketitle
+On March 16, 2016 Wikileaks  published a searchable archive \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton's private server.
+The documents were provided as pdfs by the U.S.
+Department of State in response to Freedom of Information Act (FOIA) \cite{FOIA} requests.
+The State Dept.
+also provided a searchable web archive of the pdf documents, released in several installments from May 2015 to March 2017\cite{StateDeptFOIA}.
+Both sites provide a  useful tool for anyone searching for particular terms in the documents.
+The Wikileaks site was put to much use by investigative reporters, and others, to search for topical news items.
+
+What is critically missing from either site, however, is the ability to easily conduct statistical analyses of their contents.  Only then can patterns be easily discerned.
+
+In this paper, we describe a complementary website service (\texttt{rshiny.math.uwaterloo.ca/clinton}).
+This service does not provide the search facilities found on Wikileaks, but rather provides the user with interactive visualizations that allow them to uncover patterns in the emails and, through a variety of filters on the emails,  to explore these patterns at some depth.
+The service is intended to be used in conjunction with either the Wikileaks service or the State Dept.
+FOIA service.
+The latter two provide the search mechanisms as well as the actual contents of individual emails.
+One imagines having three browsers open, one for the data base search (e.g.
+Wikileaks), one for statistical visualization and exploration (our service), and one for internet search (to provide important context for the other two services).
+Each one fills a need not met by the other two; the synergy of all three provide the user a powerful set of investigative tools.
+
+In providing the service, we have purposely restrict consideration (for the most part) to displaying patterns of simple characteristics of the emails, displaying simple characteristics of the emails,  the so-called metadata.
+
+\TODO{MORE TO COME HERE}
+
+Meta data ...  Cautionary
+
+\subsection{A brief timeline on the private email server}
+While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to the private email server and its contents.
+It also introduces some of the principal characters in the whole affair.
+More complete and in depth timelines are readily available on the internet (e.g.
+\cite{attkissonTimeline, thompsonTimeline, WashPostTimeline, clintonWikipedia, TimeMagEverything}).
+
+On November 21, 2008, the New York Times reported that Hillary Clinton had decided to accept the position of U.S.
+Secretary of State.
+On January 13, 2009 the internet domain name \texttt{clintonemail.com} was registered \cite{whoisClintonserver}; eight days later Senator Clinton was confirmed as Secretary of State.
+
+Public knowledge that a private email server being used by Secretary Clinton and others for State Department and personal communications did not surface until March 2015 \cite{NewYorkTimes2015} during the course of a U.S.
+Congressional investigation \cite{BenghaziReport} of the September 11, 2012 attack by militants on U.S.
+compounds in Benghazi Libya.
+
+The State Department had difficulty fulfilling public FOIA and House Benghazi Committee requests \cite{TakingRootWashPost} for Secretary Clinton's government emails because she had exclusively used the private server for all her email.
+On March 10, 2015,  Clinton told reporters that she turned over 30,490 emails to the State Department and deleted 31,830 emails deemed to be personal \cite{WashPostTimeline}.
+Clinton had tasked three lawyers Cheryl Mills (Clinton's former chief of staff),  David Kendall (Clinton's personal lawyer), and Heather Samuelson (a State Department staffer during Clinton's tenure) to make the determinations as to which emails were work related and which were not \cite{emailVetting, thompsonTimeline}.
+
+On March 10, 2015 the House Benghazi Committee requested that the private email server be turned over to a neutral third party to determine which emails are personal and which are government records \cite{serverRequest}, but was informed March 27 by David Kendall that no emails remained on the private server for any kind of review \cite{severScrubbedLawyer}.
+Between March 25 and 31, 2015, Paul Combetta (then the server's system administrator), erased all backup copies using \texttt{BleachBit} (see \texttt{www.bleachbit.org}).
+
+Combetta, Mills, and Samuelson will later be granted partial immunity by the Justice Department during the FBI investigations into the private email server, as were two others: Bryan Pagliano (original server manager) and  John Bentel (former director of Information Resources Management for the State Department's Executive Secretariat) \cite{immunityPolitico, immunityDailyCaller, immunityIT}.
+
+On April 12, 2015 Clinton announces that she is running for the U.S.
+Presidency.
+On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce finding classified information in the emails and that the information they found was classified at the time sent \cite{serverClasssified}, though her campaign declared that they must have been classified after the fact.
+On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.
+
+Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton on use of private email server \cite{nochargeFBI}.
+On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin have been found on his laptop.
+Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.
+In both cases, Comey and the FBI are criticized by pundits from different political parties.
+
+\section{Old stuff}
+\TODO{This material is to be reused wherever}
+All work related emails were eventually released to the State Department who subsequently published more than 30,000 on the U.S.
+To add to the confusion, only those work government emails, as vetted Secretary Clinton's lawyers
+
+The emails and the server became a source for criticism of Secretary Clinton, and others, and came to dog her bid for the presidency in 2016, in what may be one of the most contentious U.S.
+elections in history.
+Part of this criticism was levelled  More complete timelines are publicly available on the internet (e.g.
+\cite{attkissonTimeline, TimeMagEverything, clintonWikipedia}).
+For our purposes, it is enoug
+
+ 33,000 supposedly missing emails
+ Encouraged 
+ and three other American citizens.  Public knowledge of the private email server surfaces State business being conducted on a private server surfaced only after the a
 
 The 2016 US Presidential election was one of the most contentious in history.
-The existence and possible content of Hillary Clinton's private email server played an significant role in the campaign until the end.
+The existence and possible content of Hillary Clinton's private email server played a significant role in the campaign until the end.
+
+The timeline of events related to use of this private server may be found elsewhere (e.g. 
 Despite their significant nature, very few of the individuals commenting on the significance of this server had spent any significant time viewing this content.
@@
 
-\section{Data Extraction}
+\section{Data}
 
+\TODO{This needs to be short and to the point.  Just say what is done.}
 Before any of the analysis could begin, the emails had to be converted from the individual PDF and HTML form into a more easily used data structure.
@@
 
-\section{Shiny App}
-
-The app (HOST ADDRESS) provides the ability for users to place custom filters on the time frame of server time stamps and classification tags of the emails and view how a series of three displays change.
-The first of these is a spoked network graph of all corresponents with Clinton at the centre.
+\section{Displays}
+\TODO{Each of these four displays is described and commented on in context of whole time line}
+\subsection{Inner circle} 
 The spokes are separated into two groups by the radial length of the spokes, these groups are determined automatically by the largest difference in volume of correspondence present in the data between the ordered counts of number of emails sent between individuals.
@@
 
-In addition to the spoked network graph, a time series plot of daily email volume is displayed.
+\subsection{Email volume}
 This plot has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day.
@@
 
-For an even more detailed view of exactly when emails are sent and what proportion are classified, a third plot of email sending times by day is included in the web application.
+\subsection{Email times}
 This plot takes advantage of the time-stamps on every email and then plots the emails as points with times on the vertical axis and the day sent on the horizontal axis.
@@
 
-Finally, a barplot of the FOIA exemption codes used in the selected data is displayed.
+\subsection{FOIA exemptions}.
  It is important to notice that this barplot is not trivialized by the possibility of selecting individual FOIA exemption codes, as many of the emails in the data set are exempted on a number of grounds, and so have numerous FOIA exemption codes.
 Thus, this barplot is not only interesting when all codes are selected, but can be used to see which codes commonly co-occur in documents.
+
+\subsection{Term frequency and TFIDF}
+
+\section{Filters}
+\TODO{Each filter is described and displayed }
+\subsection{Time filtering}
+\subsection{Sent or received}
+\subsection{FOIA exemptions}
+\subsection{Auxiliary information}
+\TODO{e.g. Clinton's foreign schedule}
+\section{Some further exploration}
+\TODO{Some of our findings.  Unless done in the previous section}
+\section{Shiny App}
+\TODO{Again short and to the point}
+The app (HOST ADDRESS) provides the ability for users to place custom filters on the time frame of server time stamps and classification tags of the emails and view how a series of three displays change.
+
+Finally, a barplot of the FOIA exemption codes used in the selected data is displayed
 
@@
 
-Duis autem vel eum iriure dolor in hendrerit in vulputate velit esse
-molestie consequat, vel illum dolore eu feugiat nulla facilisis at
-vero eros et accumsan et iusto odio dignissim qui blandit praesent
-luptatum zzril delenit augue duis dolore te feugait nulla
-facilisi. Lorem ipsum dolor sit amet, consectetuer adipiscing elit,
-sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna
-aliquam erat volutpat~\cite{Kindlmann:1999:SAG}.
-
-\begin{equation}
-\sum_{j=1}^{z} j = \frac{z(z+1)}{2}
-\end{equation}
-
-Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
-nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
-sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
-ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
-sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
-dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
-et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
-takimata sanctus est Lorem ipsum dolor sit amet.
-
-\subsection{Lorem ipsum}
-
-Lorem ipsum dolor sit amet (see \autoref{tab:vis_papers}), consetetur sadipscing elitr, sed diam
-nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
-sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
-ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
-sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
-dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
-et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
-takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
-amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
-invidunt ut labore et dolore magna aliquyam erat, sed diam
-voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum.
-
-\begin{table}[tb]
-  \caption{VIS/VisWeek accepted/presented papers: 1990--2015.}
-  \label{tab:vis_papers}
-  \scriptsize%
-	\centering%
-  \begin{tabu}{%
-	r%
-	*{7}{c}%
-	*{2}{r}%
-	}
-  \toprule
-   year & \rotatebox{90}{Vis/SciVis} &   \rotatebox{90}{SciVis conf} &   \rotatebox{90}{InfoVis} &   \rotatebox{90}{VAST} &   \rotatebox{90}{VAST conf} &   \rotatebox{90}{TVCG @ VIS} &   \rotatebox{90}{CG\&A @ VIS} &   \rotatebox{90}{VIS/VisWeek} \rotatebox{90}{incl.
-TVCG/CG\&A}   &   \rotatebox{90}{VIS/VisWeek} \rotatebox{90}{w/o TVCG/CG\&A}   \\
-  \midrule
-  2015 & 33 & 9 & 38 & 33 & 14 & 17 & 15 & 159 & 127 \\
-  2014 & 34 &   & 45 & 33 & 21 & 20 &   & 153 & 133 \\
-  2013 & 31 &   & 38 & 32 &   & 20 &   & 121 & 101 \\
-  2012 & 42 &   & 44 & 30 &   & 23 &   & 139 & 116 \\
-  2011 & 49 &   & 44 & 26 &   & 20 &   & 139 & 119 \\
-  2010 & 48 &   & 35 & 26 &   &   &   & 109 & 109 \\
-  2009 & 54 &   & 37 & 26 &   &   &   & 117 & 117 \\
-  2008 & 50 &   & 28 & 21 &   &   &   & 99 & 99 \\
-  2007 & 56 &   & 27 & 24 &   &   &   & 107 & 107 \\
-  2006 & 63 &   & 24 & 26 &   &   &   & 113 & 113 \\
-  2005 & 88 &   & 31 &   &   &   &   & 119 & 119 \\
-  2004 & 70 &   & 27 &   &   &   &   & 97 & 97 \\
-  2003 & 74 &   & 29 &   &   &   &   & 103 & 103 \\
-  2002 & 78 &   & 23 &   &   &   &   & 101 & 101 \\
-  2001 & 74 &   & 22 &   &   &   &   & 96 & 96 \\
-  2000 & 73 &   & 20 &   &   &   &   & 93 & 93 \\
-  1999 & 69 &   & 19 &   &   &   &   & 88 & 88 \\
-  1998 & 72 &   & 18 &   &   &   &   & 90 & 90 \\
-  1997 & 72 &   & 16 &   &   &   &   & 88 & 88 \\
-  1996 & 65 &   & 12 &   &   &   &   & 77 & 77 \\
-  1995 & 56 &   & 18 &   &   &   &   & 74 & 74 \\
-  1994 & 53 &   &   &   &   &   &   & 53 & 53 \\
-  1993 & 55 &   &   &   &   &   &   & 55 & 55 \\
-  1992 & 53 &   &   &   &   &   &   & 53 & 53 \\
-  1991 & 50 &   &   &   &   &   &   & 50 & 50 \\
-  1990 & 53 &   &   &   &   &   &   & 53 & 53 \\
-  \midrule
-  \textbf{sum} & \textbf{1515} & \textbf{9} & \textbf{595} & \textbf{277} & \textbf{35} & \textbf{100} & \textbf{15} & \textbf{2546} & \textbf{2431} \\
-  \bottomrule
-  \end{tabu}%
-\end{table}
-
-\subsection{Mezcal Head}
-
-Lorem ipsum dolor sit amet (see \autoref{fig:sample}), consetetur sadipscing elitr, sed diam
-nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
-sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
-ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
-sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
-dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
-et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
-takimata sanctus est Lorem ipsum dolor sit amet.
-
-\subsubsection{Duis Autem}
-
-Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
-nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
-sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
-ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
-sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
-dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
-et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
-takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
-amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
-invidunt ut labore et dolore magna aliquyam erat, sed diam
-voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum. Stet clita kasd gubergren, no sea takimata sanctus est. Lorem
-ipsum dolor sit amet.
-
-\begin{figure}[tb]
- \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
- \includegraphics[width=\columnwidth]{paper-count-w-2015-new}
- \caption{A visualization of the data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
- \label{fig:sample}
-\end{figure}
-
-\subsubsection{Ejector Seat Reservation}
-
-Duis autem~\cite{Lorensen:1987:MCA}\footnote{The algorithm behind
-Marching Cubes \cite{Lorensen:1987:MCA} had already been
-described by Wyvill et al. \cite{Wyvill:1986:DSS} a year
-earlier.} vel eum iriure dolor in hendrerit
-in vulputate velit esse molestie consequat,\footnote{Footnotes
-appear at the bottom of the column.} vel illum dolore eu
-feugiat nulla facilisis at vero eros et accumsan et iusto odio
-dignissim qui blandit praesent luptatum zzril delenit augue duis
-dolore te feugait nulla facilisi. Lorem ipsum dolor sit amet,
-consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt
-ut laoreet dolore magna aliquam erat volutpat.
-
-\paragraph{Confirmed Ejector Seat Reservation}
-
-Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
-suscipit lobortis nisl ut aliquip ex ea commodo
-consequat~\cite{Nielson:1991:TAD}. Duis autem vel eum iriure dolor in
-hendrerit in vulputate velit esse molestie consequat, vel illum dolore
-eu feugiat nulla facilisis at vero eros et accumsan et iusto odio
-dignissim qui blandit praesent luptatum zzril delenit augue duis
-dolore te feugait nulla facilisi.
-
-\paragraph{Rejected Ejector Seat Reservation}
-
-Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
-suscipit lobortis nisl ut aliquip ex ea commodo consequat. Duis autem
-vel eum iriure dolor in hendrerit in vulputate velit esse molestie
-
-\section{Conclusion}
-
-Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
-nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
-sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
-ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
-sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
-dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
-et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
-takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
-amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
-invidunt ut labore et dolore magna aliquyam erat, sed diam
-voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum.
-
 %% if specified like this the section will be committed in review mode
 \acknowledgments{
-The authors wish to thank A, B, C. This work was supported in part by
-a grant from XYZ.}
+This work was supported in part by
+a Discovery grant from the Natural Sciences and Engineering Research Council of Canada..}
 
-%\bibliographystyle{abbrv}
-\bibliographystyle{abbrv-doi}
+\newpage
+\bibliographystyle{abbrv}
+%\bibliographystyle{abbrv-doi}
 %\bibliographystyle{abbrv-doi-narrow}
@@
 
-\bibliography{template}
+\bibliography{clinton}
 

### 2017-03-27 EmailFunctions.R
Improved the stopword coverage

@@ -543,11 +543,17 @@ get_keywords <- function(Content, stem = TRUE, table = TRUE) {
   # split by singular spaces
   corpus <- unlist(str_split(corpus, " "))
   # remove common stopwords
-  corpus <- removeWords(corpus, c(stopwords(), "s"))
+  Cust_stop <- c(stopwords(), "will", "state", "pm", "said", "secretari", "can", "call",
+                 "depart", "time", "presid", "govern", "offic", "work", "usa", "also", "meet",
+                 "new", tolower(LETTERS))
+  corpus <- removeWords(corpus, stopwords())
   # remove empty strings which may have occurred
   corpus <- corpus[corpus != ""]
   # stem the remaining terms
-  if (stem) corpus <- stemDocument(corpus)
+  if (stem) {
+    corpus <- removeWords(stemDocument(corpus), Cust_stop)
+    corpus <- corpus[corpus != ""]
+  }
   # now to extract the unique words used
   if (table) corpus <- sort(table(corpus), decreasing = TRUE)
   corpus

### 2017-03-27 VAST_submission/vgtc_journal_latex/paper.tex
fixed citation typos

@@
 
-On March 10, 2015 the House Benghazi Committee requested that the private email server be turned over to a neutral third party to determine which emails are personal and which are government records \cite{serverRequest}, but was informed March 27 by David Kendall that no emails remained on the private server for any kind of review \cite{severScrubbedLawyer}.
+On March 10, 2015 the House Benghazi Committee requested that the private email server be turned over to a neutral third party to determine which emails are personal and which are government records \cite{serverRequest}, but was informed March 27 by David Kendall that no emails remained on the private server for any kind of review \cite{serverScrubbedLawyer}.
 Between March 25 and 31, 2015, Paul Combetta (then the server's system administrator), erased all backup copies using \texttt{BleachBit} (see \texttt{www.bleachbit.org}).
@@
 Presidency.
-On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce finding classified information in the emails and that the information they found was classified at the time sent \cite{serverClasssified}, though her campaign declared that they must have been classified after the fact.
+On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce finding classified information in the emails and that the information they found was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact.
 On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.

### 2017-03-27 app.R
Added .mil colouring and modified the legend below the spiral plot

@@ -10,7 +10,7 @@ library(shinyBS)
 
 ### Pre-loaded data #####################################################################
 pal <- c("steelblue", "firebrick")
-pal2 <- c("darkorange","firebrick", "steelblue")
+pal2 <- c("darkorange","firebrick", "steelblue", "black")
 EmailData <- read.csv('ClintonEmailData.csv')
 ThompTime <- read.csv('ThompsonTimeline.csv')
 ForSched <- read.csv('ForeignSchedule.csv')
@@ -59,11 +59,14 @@ names(stateMail) <- allnames
 for (name in allnames) {
   toMail <- filter(ClintonCom, To.name == name)
   fromMail <- filter(ClintonCom, From.name == name)
-  state <- any(grepl(".gov", c(as.character(toMail$To.Add), as.character(toMail$To.name),
+  state <- any(grepl("[\\.]gov", c(as.character(toMail$To.Add), as.character(toMail$To.name),
                                as.character(fromMail$From.name), as.character(fromMail$From.Add))))
-  notState <- any(grepl("[@.]", c(as.character(toMail$To.Add), as.character(toMail$To.name),
+  mil <- any(grepl("[\\.]mil$", c(as.character(toMail$To.Add), as.character(toMail$To.name),
+                             as.character(fromMail$From.name), as.character(fromMail$From.Add))))
+  notState <- any(grepl("[@\\.]", c(as.character(toMail$To.Add), as.character(toMail$To.name),
                                   as.character(fromMail$From.name), as.character(fromMail$From.Add))))
-  stateMail[which(allnames == name)] <- as.numeric(state) + as.numeric(notState)
+  stateMail[which(allnames == name)] <- as.numeric(state) + as.numeric(notState) + 
+    2*as.numeric(mil)
 }
 
 # modify the spiral network plot function, improve it
@@ -118,12 +121,14 @@ spiralNetPlot2 <- function(centralNode = "Hillary Clinton", wgtTbl = NA,
     grid.text(title, x = 0, y = 0.98, just = "left",
               gp = gpar(face = 2))
     # add a legend
-    grid.text("Blue - Identifiably .gov", x = 0, y = 0.02,
-              just = "left", gp = gpar(face = 2))
-    grid.text("Red - Identifiably Not .gov", x = 0.98, y = 0.02,
-              just = "right", gp = gpar(face = 2))
[…]

### 2017-03-28 app.R
Modified the server logic

@@ -222,50 +222,61 @@ ui <- fluidPage(
 
 # Define server logic required to draw all displays
 server <- function(input, output) {
-  # select data using the input date and classification filters
-  DateRange <- reactive(paste(chron(c(input$range[1], input$range[2]),
-                                    format = "day mon year"),
-                              collapse = " - "))
-  selIDs <- reactive(
-    switch(input$ToFromFilter,
-           "All Emails" = filter(AsSec[apply(AsSec[,input$ClassFilter, drop = FALSE],1,any),],
-                                Date < input$range[2] + 1 & Date >= input$range[1])$ID,
-           "From Clinton" = filter(AsSec[apply(AsSec[,input$ClassFilter, drop = FALSE],1,any),],
-                                 Date < input$range[2] + 1 & Date >= input$range[1] &
-                                   (From.name == "Hillary Clinton" | From.name == "H"))$ID,
-           "To Clinton" = filter(AsSec[apply(AsSec[,input$ClassFilter, drop = FALSE],1,any),],
-                                 Date < input$range[2] + 1 & Date >= input$range[1] &
-                                   (To.name == "Hillary Clinton" | To.name == "H"))$ID)
-  )
-  selDays <- reactive(ASDays[ASDays < input$range[2] + 1 & ASDays >= input$range[1]])
-  selCounts <- reactive(AScounts[ASDays < input$range[2] + 1 & ASDays >= input$range[1]])
-  Sched <- reactive(filter(ForSched, (StartDate >= input$range[1] &
-                                        StartDate < input$range[2]) |
-                             (EndDate < input$range[2] & EndDate >= input$range[1])))
-  dispSched <- reactive(input$Schedule)
-  adjVar <- reactive(input$AdjVar == "18:00")
+  # insert a caching function
+  inputProc <- function(input) reactive({
+    # generate a storage item
+    intermed <- list()
+    # select data using the input date and classification filters
+    intermed$DateRange <- paste(chron(c(input$range[1], input$range[2]),
+                                      format = "day mon year"),
+                                collapse = " - ")
+    intermed$selIDs <- 
+      switch(input$ToFromFilter,
[…]

### 2017-03-28 VAST_submission/vgtc_journal_latex/paper.tex
Added some references, updated and added to the paper

@@
 
-In providing the service, we have purposely restrict consideration (for the most part) to displaying patterns of simple characteristics of the emails, displaying simple characteristics of the emails,  the so-called metadata.
+In providing the service, we have purposely restrict consideration (for the most part) to displaying patterns of simple characteristics of the emails, the so-called metadata.
+The reasons for this are two-fold.
+The first is that for this data the metadata is simply more reliable and regular.
+The emails are often heavily redacted, and this introduces both gaps in content and irregularities.
+The metadata, however, remains somewhat informative and regular in format despite redactions, as the only redacted metadata is that of individual email addresses, and noting these redactions is in itself useful.
+The second reason is demonstrate the power of supported metadata in the exploration of communication data.
+For public figures in particular, a communication network plot and time-stamps can be very easily leveraged for insightful data exploration if outside sources are consulted.
+The individual's communication habits can be extended to gain insight into their general habits in life, while the individuals they communicate with most frequently can establish interesting connections which might otherwise be missed, such as that between Clinton and Tony Blair.
 
@@
 \TODO{This needs to be short and to the point.  Just say what is done.}
-Before any of the analysis could begin, the emails had to be converted from the individual PDF and HTML form into a more easily used data structure.
-The choice was made to use the Wikileaks HTML emails as the source for data extraction.
-This choice was made primarily due to the relative ease of loading the raw HTML pages on the Wikileaks database due to their regular method of storage (the email with ID <id> is stored at https://wikileaks.org/clinton-emails/emailid/<id>) and ease of loading.
-These qualities made the HTML emails amenable to programmatic extraction, which was completed using a host of web-scraping and string parsing tools in \texttt{R}.
-In particular the packages \texttt{RCurl} and \texttt{XML} proved invaluable to extract the raw HTML pages for each email, and the packages \texttt{tm}, \texttt{stringr}, and \texttt{SnowballC} were indispensible to process and clean the raw HTML into a useable form.
+HTML versions of the emails were pulled programmatically from the Wikileaks database, as the storage structure supported programmatic extraction using web-scraping and string parsing tools in \texttt{R}.
+In particular, the packages \texttt{RCurl} and \texttt{XML} proved invaluable to extract the raw HTML pages for each email, and the packages \texttt{tm}, \texttt{stringr}, and \texttt{SnowballC} were indispensible to process and clean the raw HTML into a useable form.
 After downloading the raw HTML and extracting the data of interest, the resulting data was stored in csv files to provide the flexibility to load the data into any architecture or analysis tool desired.
 
-Unfortunately, this data is not perfect.
-The PDF files released by the state department do not include the typical email server header with information about the time stamp inforamtion, the addresses of all involved, and other useful and regular information.
-Instead, these files appear more similar to screen captures of the emails, providing only the visual display a user would see without the useful information used to generate that display.
-As a result, the email addresses are occasionally not included, and there were initially concerns over the time zones used to generate the time stamp metadata.
-Furthermore, the HTML data on Wikileaks is frequently irregular in form, and although work has been done to make the cleaning and extraction functions as general as possible, there are still certainly fringe cases not considered which occur frequently enough to have an impact.
-It is also important to keep in mind the fact that the data have already been filtered twice.
-First by the emails chosen to be released by Clinton, and next by the emails and email sections which the State Department chose not to redact.
-Indeed, even the metadata can be affected by this redaction, as a number of individuals have their email addresses redacted in all communications, preventing certain analysis from being performed.
+The PDF files released by the state department used to generate the Wikileaks HTML emails do not include the typical email server header with time stamp information, email addresses, and other regular data.
+Instead, these files are more similar to screen captures of the emails, providing only the visual display a user would see without the useful information used to generate that display.
+As a result, the data is very irregular: email addresses are occasionally not included, there were initially concerns over the time zones used to generate the time stamp metadata, and there are doubtless still many fringe cases of irregular spacing or placement of text within the HTML code not considered in cleaning.
+While the metadata is relatively consistent in spite of these irregularities, there are still impure and messy addresses and contact names arising from these irregularities.
+
+After the emails were pulled, stopwords defined by the \texttt{stopwords} function from \texttt{tm} were removed and the remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.
+Both of these seemed imperfect and ill-suited to the data, failing to remove words with little informative power and creating often cryptic stems.
 
@@
 As well, the width of the edges of the graph is determined by the volume of correspondence which occurred between Clinton and the correspondent in either direction.
-Finally, all edges and points are coloured according to a simple scheme.
-If the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, they are coloured blue.
-Individuals with email addresses which are identifiably not hosted at a '.gov' domain are coloured red.
-Finally, individuals which cannot have their email host domain identified are coloured grey.
-It is important to note that for those individuals with grey spokes, the reason the email is not identifiable is the removal of the email by the state department.
-Such a redaction is not a guarantee that the email said individual used is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain.
-Still, such speculation cannot be conclusively made, and so these emails are simply reported as they are, unidentifiable.
+Finally, all edges and points are coloured according to a simple scheme: if the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, their node is coloured blue; nodes with email addresses which are identifiably not hosted at a '.gov' domain are coloured red; nodes with '.mil' accounts are coloured black; and nodes unidentifiable hosts are coloured orange.
+It is important to note that for those individuals with orange spokes, the most likely reason the email is not identifiable is the redaction of the email by the state department.
+Such a redaction is not a guarantee that the email of the individual is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain.
 
@@
 
+One such result which is immediately obvious is the strange pattern of sending times.
+Email communications occur with the highest frequency between midnight and 3 am, and with the lowest frequency between 4 pm and 10 pm.
+This pattern, which is persistent through filters of emails sent to Clinton and by Clinton, is surprising as it indicates that the highest level of communication was occurring at a time typically devoted to sleep.
+Another pattern immediately obvious is the alternating modes at 2 and 3 am, with alternations occurring on the dates of daylight savings in North America.
+The regularity of this pattern in the emails suggests some server function performed early in the morning, perhaps some maintenance operation performed every 24 hours.
+
 \subsection{FOIA exemptions}.
- It is important to notice that this barplot is not trivialized by the possibility of selecting individual FOIA exemption codes, as many of the emails in the data set are exempted on a number of grounds, and so have numerous FOIA exemption codes.
-Thus, this barplot is not only interesting when all codes are selected, but can be used to see which codes commonly co-occur in documents.
+ The United States Freedom of Information Act (FOIA) exemption codes \cite{FOIA} are used by the United States Federal Government to remove sensitive information from released documents accessed using a Freedom of Information Act request.
+The sensitive matters being addressed by the codes are national security and foreign policy matters for B1, personnel practices for B2, exempted by statute for B3, trade secrets or financial information obtained in confidence for B4, inter- or intra-agency memorandums for B5, personal privacy for B6, records compiled for law enforcement for B7, prepared in relation to financial monitoring institutions for B8, and geophysical and geological information concerning wells for B9 exemptions.
+The frequencies of each of these codes within filtered emails are shown in the final display.
+Note that as emails can have numerous redaction codes, this barplot can be useful to identify which codes co-occur over a selection.
 
 \subsection{Term frequency and TFIDF}
+The term frequency and tf-idf displays for this data are perhaps the least directly useful.
+As mentioned above, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} proved inadequate for the messy and non-standard data.
+In particular, the omission of many common and uninformative action verbs such as 'will' and 'can' from the stopwords used led to very unsatisfactory results.
+A set of custom stopwords was therefore applied in an attempt to remove some of these uninformative words.
+The tf-idf was more informative, but the stemming algorithm, when coupled with the typos and abbreviations which are incredibly common in the emails, meant that the results are still incredibly unclear.
+The terms with the highest scores over any interval do not seem to communicate anything particularly useful about the content of emails sent in that range, with one notable exception in the case of the term 'windrush.' This term appeared with a high tf-idf in a number of selections which contained communications with Tony Blair (email contact "aclb" in the spiral plot), and seems to refer to Windrush Ventures, a company owned by Blair which appears vital in managing Blair's complicated and opaque finances since he left the office of the Prime Minister of the United Kingdom.
+\cite{windrushTelegraph, windrushGuardian}
 
@@
 \subsection{Time filtering}
+Time filtering within the app can be completed at the level of individual days.
+The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days.
+As well, clicking and dragging the blue bar in between the boundaries allows the user to sweep the data with an interval of a fixed width.
+Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
 \subsection{Sent or received}
+A drop down menu allows users to select emails by those sent to Clinton, sent by Clinton, or by the entire email database.
 \subsection{FOIA exemptions}
+Multiple filtering by FOIA exemption codes is supported in the app with a box selection area.
+Users can add FOIA exemption codes currently not included to the data by clicking the box with their mouse and selecting a code from the menu which is then displayed below the box.
+To remove codes, the user simply clicks the code within the display box and presses delete, which moves the code to the selection menu.
+The box therefore displays all codes currently included in the data displayed.
 \subsection{Auxiliary information}
 \TODO{e.g. Clinton's foreign schedule}
+A number of cosmetic adjustments using auxiliary information can also be made.
+Users can choose to adjust the scatterplot of time by day, which re-defines 6 pm as the time at which the day changes.
+This simply groups the emails best, placing all emails into one mostly uninterrupted region by redefining the day about the midsection of the region of lowest email traffic.
+Another cosmetic adjustment is the inclusion of Clinton's foreign trip schedule as blue regions on the email volume plot.
+This schedule was taken from the official State Department website \cite{ForeignSched} and allows users to distinguish periods spent out of the United States (coloured in blue) from those spent within the United States (uncoloured).
 \section{Some further exploration}
@@
 \TODO{Again short and to the point}
-The app (HOST ADDRESS) provides the ability for users to place custom filters on the time frame of server time stamps and classification tags of the emails and view how a series of three displays change.
-
-Finally, a barplot of the FOIA exemption codes used in the selected data is displayed
+The \texttt{R} \texttt{shiny} app at (HOST ADDRESS) provides the ability for users to interactively filter the data on all features described above and view the resulting displays.
+Alongside this display is a short article explaining the use of the app, providing links to background information, and outlining some of the more interesting trends observed.
+It is hosted on the Waterloo \texttt{rshiny} server for public access, allowing for any individual to make use of this tool to generate hypotheses and investigate this intriguing dataset.
 

### 2017-03-28 app.R
Changed labelling

@@ -123,11 +123,11 @@ spiralNetPlot2 <- function(centralNode = "Hillary Clinton", wgtTbl = NA,
     # add a legend
     grid.text(".gov", x = 0, y = 0.02, just = "left", 
               gp = gpar(face = 2, col = "steelblue"))
-    grid.text(".mil", x = 0.1, y = 0.02, just = "left",
+    grid.text(".mil", x = 0.33, y = 0.02, just = "center",
               gp = gpar(face = 2, col = "black"))
-    grid.text("Not .gov", x = 0.2, y = 0.02, just = "left", 
+    grid.text("Not .gov", x = 0.66, y = 0.02, just = "center", 
               gp = gpar(face = 2, col = "firebrick"))
-    grid.text("Unidentifiable", x = 0.3, y = 0.02, just = "left", 
+    grid.text("Unidentifiable", x = 1, y = 0.02, just = "right", 
               gp = gpar(face = 2, col = "darkorange"))
   }
   else  {

### 2017-03-28 VAST_submission/vgtc_journal_latex/paper.tex
Added images

@@
   \begin{tabular}{cc}
-  \includegraphics[width=0.25\textheight]{DailySept11ToNov152012_All} &
-  \includegraphics[width=0.25\textheight]{InnerCircleSept11ToNov152012_All} \\
+  \includegraphics[height=0.22\textheight]{EmailSentRecdSept11toNov23} &
+  \includegraphics[height=0.22\textheight]{InnerCircleSept11ToNov23} \\
   &\\
-  {\small (a) Missing data in November.} & {\small (b) Inner Circle: Correspondent info by colour; volume by thickness.}
+  {\small (a) Missing block of email: October 30 - November 9} & {\small (b) Inner Circle: Correspondent info by colour; volume by thickness.}
   \end{tabular}
-  \caption{Metadata: Time, date, correspondents, volume.  Filtered: Sept. 11 to Nov. 15, 2012.  
+  \caption{Metadata: Time, date, redaction, correspondents, volume, source type.  Filtered: Sept. 11 to Nov. 23, 2012.  
   %Source: \href{https://wikileaks.org/clinton-emails/}{\texttt{https://wikileaks.org/clinton-emails/}}
@@
 
-On March 16, 2016 Wikileaks  published a searchable archive \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton's private server.
+On March 16, 2016 Wikileaks  published a searchable archive \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton on her private server.
 The documents were provided as pdfs by the U.S.
@@
 The State Dept.
-also provided a searchable web archive of the pdf documents, released in several installments from May 2015 to March 2017\cite{StateDeptFOIA}.
+also provided a searchable web archive of the documents, released in several instalments from May 2015 to March 2017\cite{StateDeptFOIA}.
 Both sites provide a  useful tool for anyone searching for particular terms in the documents.
@@
 
-What is critically missing from either site, however, is the ability to easily conduct statistical analyses of their contents.  Only then can patterns be easily discerned.
+What is critically missing from either site is a facility to learn summary, or statistical, features across all, or groups of,  emails.
+A separate site where visual displays of summary features of the emails, which can be filtered by user selection, will go a long way towards filling this need.
+The site would be \emph{complementary} to the more typical ``content search and record display'' site as provided by either online archive.
+Add general internet search in a third browser, and the three combined would provide the means to actively explore the contents of any email archive.
+One imagines, for example, having three browsers open simultaneously, one for the data base search (e.g.
+Wikileaks), one for statistical visualization and exploration (proposed here), and one for internet search (to provide important context for the other two services).
+Each fills a need not met by the other two; the synergy of all three provide the user a powerful set of investigative tools.
 
-In this paper, we describe a complementary website service (\texttt{rshiny.math.uwaterloo.ca/clinton}).
-This service does not provide the search facilities found on Wikileaks, but rather provides the user with interactive visualizations that allow them to uncover patterns in the emails and, through a variety of filters on the emails,  to explore these patterns at some depth.
-The service is intended to be used in conjunction with either the Wikileaks service or the State Dept.
-FOIA service.
-The latter two provide the search mechanisms as well as the actual contents of individual emails.
-One imagines having three browsers open, one for the data base search (e.g.
-Wikileaks), one for statistical visualization and exploration (our service), and one for internet search (to provide important context for the other two services).
-Each one fills a need not met by the other two; the synergy of all three provide the user a powerful set of investigative tools.
+In this paper, we describe an implementation of the missing visual analytic service and illustrate it on Secretary Clinton's e-mails.
+The implementation is located at  \texttt{rshiny.math.uwaterloo.ca/clinton} where any visitor may interactively explore summary features  of the entire corpus of emails.
+Different visual displays present different salient features from all selected emails; emails can be selected via a variety of interactive filters.
+All displays are reactive and update simultaneously in response to user interaction with the data filters.
+Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in the emails.
 
-In providing the service, we have purposely restrict consideration (for the most part) to displaying patterns of simple characteristics of the emails, the so-called metadata.
-The reasons for this are two-fold.
-The first is that for this data the metadata is simply more reliable and regular.
-The emails are often heavily redacted, and this introduces both gaps in content and irregularities.
-The metadata, however, remains somewhat informative and regular in format despite redactions, as the only redacted metadata is that of individual email addresses, and noting these redactions is in itself useful.
-The second reason is demonstrate the power of supported metadata in the exploration of communication data.
-For public figures in particular, a communication network plot and time-stamps can be very easily leveraged for insightful data exploration if outside sources are consulted.
-The individual's communication habits can be extended to gain insight into their general habits in life, while the individuals they communicate with most frequently can establish interesting connections which might otherwise be missed, such as that between Clinton and Tony Blair.
+The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} ... 
+\TODO{paper overview now}
 
-\TODO{MORE TO COME HERE}
+\section{The data}
+\label{sect:data}
+As of March 3, 2017, a total of 32,795 available emails, either to or from Hillary Clinton, have been made publicly available in pdf form as a searchable database \cite{StateDeptFOIA}.
+Many of these have been redacted according to the FOIA exemption codes \cite{FOIA}.
 
-Meta data ...  Cautionary
+Wikileaks \cite{Wikileaks} has provided the same redacted pdfs and, more usefully for analysis purposes, an HTML version of each pdf.
+Consequently, have used the Wikileaks database as our data source.
+All data extraction, cleaning, analysis, and presentation is done using the open source statistical programming language \texttt{R}~\cite{Rsystem}.
 
-\subsection{A brief timeline on the private email server}
+\subsection{Data extraction and cleaning}
+The raw HTML of each message was programmatically downloaded from the Wikileaks archive using \texttt{R} packages  \texttt{RCurl}~\cite{RCurl2016package} and \texttt{XML}~\cite{XML2016package}.
+This took several hours and required \TODO{Chris: mentioned etiquette concerns here?
+Timeout?}
+Once downloaded, the HTML data were processed using the \texttt{R} packages \texttt{tm}~\cite{tm2008paper, tm2017package}, \texttt{stringr}~\cite{stringr2016package}, and \texttt{SnowballC}~\cite{snowballc2014package}.
+
+After downloading the raw HTML and extracting the data of interest, the resulting data was stored in \texttt{csv} files to provide the flexibility to load the data into any architecture or analysis tool desired.
+These will be transferred to a relational data base should our server traffic warrant it.
+
+\subsection{Metadata}
+For each email, from the HTML we extract as best we can the identity of the persons sending and receiving the email as well as the date and time at which the email was processed by the server.
+\TODO{Chris:  Can you mention which fields we did or did not actually use?
+i.e.
+To: From" CC: BCC: etc.?}
+
+The HTML was constructed from email printed out, redacted, and then provided as pdfs by the State Department.
+Consequently, detailed email header information as would normally be available electronically is mostly missing.
+All time stamps appear to be the local date and time at which the server sent or received that email (e.g.
+no time zone or other source time or IP chain information is available).
+Moreover, because of redaction (typically FOIA exemption B6 ~\cite{FOIA}),  sender and receiver emails may have truncated domains,  contain only the person's name, or be missing altogether.
+In cases where both \texttt{From} and \texttt{To} are entirely missing, but there is an email chain within the message, we  \TODO{Chris?
+What happens?}  (e.g.
+see \texttt{https://wikileaks.org/clinton-emails/emailid/31599}).
+
+Using regular expressions to extract the metadata was occasionally challenging and it is always possible that some fringe cases have been mishandled.
+On the whole the metadata is fairly consistent and only rarely do some impure and messy addresses and contact names arise  due to irregular spacing or placement of text within the HTML code.
+One such fringe case is Huma Abedin's \texttt{clintonemail.com} email account which will show in the displays as an overly long string.
+We have chosen not to special case this but leave it as is.
+
+Where possible, the email addresses have also been parsed so that they may be classified into one of four categories: those which are \texttt{.gov},  those which are \texttt{.mil},  those which are identifiable as coming from a domain that is neither  \texttt{.gov} nor  \texttt{.mil}, and those whose domain was not identifiable from the data.
+\subsection{Redaction information}
+Emails that are redacted are marked as such by the string ``RELEASE IN PART'' and by the presence of one or more of nine FOIA exemption codes B1, B2, \ldots, B9  marking the place where text is missing (redacted).
+This provides two further pieces of quasi-metadata (since it is not actual email content) that can be used in analysis.
+\subsection{Content information}
+The entire (redacted) content of each email is available on the State Department and Wikileaks sites and the user is encouraged to view it there (the pdf forms are more informative, especially in appreciating the extent of the redactions).
+To provide some coarse statistical summaries of the content, all word tokens are extracted and partially processed to reduce their number.
+For example,  ``stopwords'', as identified by the\texttt{stopwords} function from  were removed.
+\TODO{Chris: I see from EmailFunctions.R that you only use the default stopwords() and an additional list.
+Maybe you should use union(stopwords(), stopwords(kind="SMART"), yourList) ?}  The remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.
+While neither the stopword lists nor the stemming tools are particularly well tuned to this corpus of emails, they nevertheless provide some insight as to the topics covered.
+
+\section{Displays}
+\label{sect:Displays}
+\TODO{Each of these four displays is described and commented on in context of whole time line}
+Four displays of the metadata and quasi-metadata
+
+\subsection{Inner circle} 
+The spokes are separated into two groups by the radial length of the spokes, these groups are determined automatically by the largest difference in volume of correspondence present in the data between the ordered counts of number of emails sent between individuals.
+As well, the width of the edges of the graph is determined by the volume of correspondence which occurred between Clinton and the correspondent in either direction.
+Finally, all edges and points are coloured according to a simple scheme: if the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, their node is coloured blue; nodes with email addresses which are identifiably not hosted at a '.gov' domain are coloured red; nodes with '.mil' accounts are coloured black; and nodes unidentifiable hosts are coloured orange.
+It is important to note that for those individuals with orange spokes, the most likely reason the email is not identifiable is the redaction of the email by the state department.
+Such a redaction is not a guarantee that the email of the individual is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain.
+
+\begin{center}
+\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
+\end{center}
+
+\subsection{Email volume}
+This plot has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day.
+This allows users to view which days have an exceptionally high or low number of classified conversations.
+Returning to the motivation of hypothesis generation, referencing world events on any days of interest provides a number of interesting insights on how public figures like Clinton react and respond to world events.
+
+\begin{center}
+\includegraphics[width=0.95\linewidth]{VolumeFullTime}
+\end{center}
+
+\subsection{Email times}
+This plot takes advantage of the time-stamps on every email and then plots the emails as points with times on the vertical axis and the day sent on the horizontal axis.
+The points in this scatterplot are then coloured according to whether the corresponding email was redacted, in which case the point is coloured red, or released in full, in which case the point is coloured blue.
+Alpha blending and transparency of these points was utilized to make the patterns present more obvious and prevent overplotting from obscuring any interesting results.
+
+\begin{center}
+\includegraphics[width=0.95\linewidth]{DailyFullTime}
+\end{center}
+
+One such result which is immediately obvious is the strange pattern of sending times.
+Email communications occur with the highest frequency between midnight and 3 am, and with the lowest frequency between 4 pm and 10 pm.
+This pattern, which is persistent through filters of emails sent to Clinton and by Clinton, is surprising as it indicates that the highest level of communication was occurring at a time typically devoted to sleep.
+Another pattern immediately obvious is the alternating modes at 2 and 3 am, with alternations occurring on the dates of daylight savings in North America.
+The regularity of this pattern in the emails suggests some server function performed early in the morning, perhaps some maintenance operation performed every 24 hours.
+
+\subsection{FOIA exemptions}
+The United States Freedom of Information Act (FOIA) exemption codes \cite{FOIA} are used by the United States Federal Government to remove sensitive information from released documents accessed using a Freedom of Information Act request.
+The sensitive matters being addressed by the codes are national security and foreign policy matters for B1, personnel practices for B2, exempted by statute for B3, trade secrets or financial information obtained in confidence for B4, inter- or intra-agency memorandums for B5, personal privacy for B6, records compiled for law enforcement for B7, prepared in relation to financial monitoring institutions for B8, and geophysical and geological information concerning wells for B9 exemptions.
+The frequencies of each of these codes within filtered emails are shown in the final display.
+Note that as emails can have numerous redaction codes, this barplot can be useful to identify which codes co-occur over a selection.
+\begin{center}
+\includegraphics[width=0.95\linewidth]{ExemptionsFullTime}
+\end{center}
+
+\subsection{Term frequency and TFIDF}
+The term frequency and tf-idf displays for this data are perhaps the least directly useful.
+As mentioned above, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} proved inadequate for the messy and non-standard data.
+In particular, the omission of many common and uninformative action verbs such as 'will' and 'can' from the stopwords used led to very unsatisfactory results.
+A set of custom stopwords was therefore applied in an attempt to remove some of these uninformative words.
+The tf-idf was more informative, but the stemming algorithm, when coupled with the typos and abbreviations which are incredibly common in the emails, meant that the results are still incredibly unclear.
+The terms with the highest scores over any interval do not seem to communicate anything particularly useful about the content of emails sent in that range, with one notable exception in the case of the term 'windrush.' This term appeared with a high tf-idf in a number of selections which contained communications with Tony Blair (email contact "aclb" in the spiral plot), and seems to refer to Windrush Ventures, a company owned by Blair which appears vital in managing Blair's complicated and opaque finances since he left the office of the Prime Minister of the United Kingdom.
+\cite{windrushTelegraph, windrushGuardian}
+
+\section{Filters}
+\label{sect:Filters}
+\TODO{Each filter is described and displayed }
+\subsection{Time filtering}
+Time filtering within the app can be completed at the level of individual days.
+The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days.
+As well, clicking and dragging the blue bar in between the boundaries allows the user to sweep the data with an interval of a fixed width.
+Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
+\subsection{Sent or received}
+A drop down menu allows users to select emails by those sent to Clinton, sent by Clinton, or by the entire email database.
+\subsection{FOIA exemptions}
+Multiple filtering by FOIA exemption codes is supported in the app with a box selection area.
+Users can add FOIA exemption codes currently not included to the data by clicking the box with their mouse and selecting a code from the menu which is then displayed below the box.
+To remove codes, the user simply clicks the code within the display box and presses delete, which moves the code to the selection menu.
+The box therefore displays all codes currently included in the data displayed.
+\subsection{Auxiliary information}
+\TODO{e.g. Clinton's foreign schedule}
+A number of cosmetic adjustments using auxiliary information can also be made.
+Users can choose to adjust the scatterplot of time by day, which re-defines 6 pm as the time at which the day changes.
+This simply groups the emails best, placing all emails into one mostly uninterrupted region by redefining the day about the midsection of the region of lowest email traffic.
+Another cosmetic adjustment is the inclusion of Clinton's foreign trip schedule as blue regions on the email volume plot.
+This schedule was taken from the official State Department website \cite{ForeignSched} and allows users to distinguish periods spent out of the United States (coloured in blue) from those spent within the United States (uncoloured).
+
+\section{Analysis}
+\label{sect:Analysis}
+
+\subsection{Context: A brief timeline on the private email server}
+\label{sect:Analysis:Timeline}
 While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to the private email server and its contents.
@@
 
-\section{Old stuff}
-\TODO{This material is to be reused wherever}
-All work related emails were eventually released to the State Department who subsequently published more than 30,000 on the U.S.
-To add to the confusion, only those work government emails, as vetted Secretary Clinton's lawyers
-
-The emails and the server became a source for criticism of Secretary Clinton, and others, and came to dog her bid for the presidency in 2016, in what may be one of the most contentious U.S.
-elections in history.
-Part of this criticism was levelled  More complete timelines are publicly available on the internet (e.g.
-\cite{attkissonTimeline, TimeMagEverything, clintonWikipedia}).
-For our purposes, it is enoug
-
- 33,000 supposedly missing emails
- Encouraged 
- and three other American citizens.  Public knowledge of the private email server surfaces State business being conducted on a private server surfaced only after the a
-
-The 2016 US Presidential election was one of the most contentious in history.
-The existence and possible content of Hillary Clinton's private email server played a significant role in the campaign until the end.
-
-The timeline of events related to use of this private server may be found elsewhere (e.g. 
-Despite their significant nature, very few of the individuals commenting on the significance of this server had spent any significant time viewing this content.
-This is no fault of theirs, the officially released data has previously only been stored in databases on the United States' State Department Website (see https://foia.state.gov/Learn/New.aspx) and Wikileaks (see https://wikileaks.org/clinton-emails/) in individually searchable form.
-Furthermore, the email data is stored in the inconvenient format of individual PDF documents, and in the case of Wikileaks a slightly more convenient but still cumbersome HTML representation based on the official PDF content.
-This granularity and separation of individual emails on different webpages prevents a great deal interesting analysis of the data, including any aggregate analyses of patterns in content or metadata.
-As such, the emails are incredibly misunderstood documents, and current searches for patterns and points of interest have relied upon searching for terms of interest and sifting through the results of the search.
-Such a method is not only tedious but problematic from a statistical point of view, as approaching any data set with a specific hypothesis before exploring its structure and patterns in abstract leads to biased conclusions and poor analysis.
-In the interest of simplifying the general exploration of this important data set, the Wikileaks HTML email versions were extracted and a series of visualizations were generated around the metadata included in these emails.
-An interactive web application in \texttt{R} \texttt{shiny} was constructed utilizing these visualizations and the extracted data.
-Incorporated in this app was a short article explaining the use of the app, providing links to background information, and outlining some of the findings of the more interesting trends observed.
-This framework provides any individual with interest the ability to generate hypotheses by exploring the data without necessarily searching for any particular conclusion.
-
-In generating and exploring this data, a considerable amount of respect is gained for the power of the often overlooked metadata of emails, especially with public figures.
-Simply having access to the network of communications, simple indicators of identifiable features of the individuals communicating, and times sent is enough to generate interesting questions which simple Google searches of important events and days can shed a great deal of light on.
-Leveraging data beyond this metadata, including the Freedom of Information Act (FOIA) exemption codes (see https://vault.fbi.gov/explanation-of-exemptions) used to redact the emails, only bolsters this simple metadata driven investigation.
-Metadata is often viewed as less significant or somehow less intrusive data, but the exploration performed in this paper demonstrated, at the very least to the authors, what a powerful tool metadata is for generating hypotheses about data.
-Given the lack of privacy present in the modern internet age and the amount of information most individuals make publicly available, there is a clear warning here about the utility of metadata for bad or for good.
-For a public official like Hillary Clinton, the data become even more interesting due to a number of contentious issues which arose during her tenure as the Secretary of State, independent of and as a result of her use of a private server.
-
-\section{Data}
-
-\TODO{This needs to be short and to the point.  Just say what is done.}
-HTML versions of the emails were pulled programmatically from the Wikileaks database, as the storage structure supported programmatic extraction using web-scraping and string parsing tools in \texttt{R}.
-In particular, the packages \texttt{RCurl} and \texttt{XML} proved invaluable to extract the raw HTML pages for each email, and the packages \texttt{tm}, \texttt{stringr}, and \texttt{SnowballC} were indispensible to process and clean the raw HTML into a useable form.
-After downloading the raw HTML and extracting the data of interest, the resulting data was stored in csv files to provide the flexibility to load the data into any architecture or analysis tool desired.
-
-The PDF files released by the state department used to generate the Wikileaks HTML emails do not include the typical email server header with time stamp information, email addresses, and other regular data.
-Instead, these files are more similar to screen captures of the emails, providing only the visual display a user would see without the useful information used to generate that display.
-As a result, the data is very irregular: email addresses are occasionally not included, there were initially concerns over the time zones used to generate the time stamp metadata, and there are doubtless still many fringe cases of irregular spacing or placement of text within the HTML code not considered in cleaning.
-While the metadata is relatively consistent in spite of these irregularities, there are still impure and messy addresses and contact names arising from these irregularities.
-
-After the emails were pulled, stopwords defined by the \texttt{stopwords} function from \texttt{tm} were removed and the remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.
-Both of these seemed imperfect and ill-suited to the data, failing to remove words with little informative power and creating often cryptic stems.
-
-\section{Displays}
-\TODO{Each of these four displays is described and commented on in context of whole time line}
-\subsection{Inner circle} 
-The spokes are separated into two groups by the radial length of the spokes, these groups are determined automatically by the largest difference in volume of correspondence present in the data between the ordered counts of number of emails sent between individuals.
-As well, the width of the edges of the graph is determined by the volume of correspondence which occurred between Clinton and the correspondent in either direction.
-Finally, all edges and points are coloured according to a simple scheme: if the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, their node is coloured blue; nodes with email addresses which are identifiably not hosted at a '.gov' domain are coloured red; nodes with '.mil' accounts are coloured black; and nodes unidentifiable hosts are coloured orange.
-It is important to note that for those individuals with orange spokes, the most likely reason the email is not identifiable is the redaction of the email by the state department.
-Such a redaction is not a guarantee that the email of the individual is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain.
-
-\subsection{Email volume}
-This plot has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day.
-This allows users to view which days have an exceptionally high or low number of classified conversations.
-Returning to the motivation of hypothesis generation, referencing world events on any days of interest provides a number of interesting insights on how public figures like Clinton react and respond to world events.
-
-\subsection{Email times}
-This plot takes advantage of the time-stamps on every email and then plots the emails as points with times on the vertical axis and the day sent on the horizontal axis.
-The points in this scatterplot are then coloured according to whether the corresponding email was redacted, in which case the point is coloured red, or released in full, in which case the point is coloured blue.
-Alpha blending and transparency of these points was utilized to make the patterns present more obvious and prevent overplotting from obscuring any interesting results.
-
-One such result which is immediately obvious is the strange pattern of sending times.
-Email communications occur with the highest frequency between midnight and 3 am, and with the lowest frequency between 4 pm and 10 pm.
-This pattern, which is persistent through filters of emails sent to Clinton and by Clinton, is surprising as it indicates that the highest level of communication was occurring at a time typically devoted to sleep.
-Another pattern immediately obvious is the alternating modes at 2 and 3 am, with alternations occurring on the dates of daylight savings in North America.
-The regularity of this pattern in the emails suggests some server function performed early in the morning, perhaps some maintenance operation performed every 24 hours.
-
-\subsection{FOIA exemptions}.
- The United States Freedom of Information Act (FOIA) exemption codes \cite{FOIA} are used by the United States Federal Government to remove sensitive information from released documents accessed using a Freedom of Information Act request.
-The sensitive matters being addressed by the codes are national security and foreign policy matters for B1, personnel practices for B2, exempted by statute for B3, trade secrets or financial information obtained in confidence for B4, inter- or intra-agency memorandums for B5, personal privacy for B6, records compiled for law enforcement for B7, prepared in relation to financial monitoring institutions for B8, and geophysical and geological information concerning wells for B9 exemptions.
-The frequencies of each of these codes within filtered emails are shown in the final display.
-Note that as emails can have numerous redaction codes, this barplot can be useful to identify which codes co-occur over a selection.
-
-\subsection{Term frequency and TFIDF}
-The term frequency and tf-idf displays for this data are perhaps the least directly useful.
-As mentioned above, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} proved inadequate for the messy and non-standard data.
-In particular, the omission of many common and uninformative action verbs such as 'will' and 'can' from the stopwords used led to very unsatisfactory results.
-A set of custom stopwords was therefore applied in an attempt to remove some of these uninformative words.
-The tf-idf was more informative, but the stemming algorithm, when coupled with the typos and abbreviations which are incredibly common in the emails, meant that the results are still incredibly unclear.
-The terms with the highest scores over any interval do not seem to communicate anything particularly useful about the content of emails sent in that range, with one notable exception in the case of the term 'windrush.' This term appeared with a high tf-idf in a number of selections which contained communications with Tony Blair (email contact "aclb" in the spiral plot), and seems to refer to Windrush Ventures, a company owned by Blair which appears vital in managing Blair's complicated and opaque finances since he left the office of the Prime Minister of the United Kingdom.
-\cite{windrushTelegraph, windrushGuardian}
-
-\section{Filters}
-\TODO{Each filter is described and displayed }
-\subsection{Time filtering}
-Time filtering within the app can be completed at the level of individual days.
-The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days.
-As well, clicking and dragging the blue bar in between the boundaries allows the user to sweep the data with an interval of a fixed width.
-Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
-\subsection{Sent or received}
-A drop down menu allows users to select emails by those sent to Clinton, sent by Clinton, or by the entire email database.
-\subsection{FOIA exemptions}
-Multiple filtering by FOIA exemption codes is supported in the app with a box selection area.
-Users can add FOIA exemption codes currently not included to the data by clicking the box with their mouse and selecting a code from the menu which is then displayed below the box.
-To remove codes, the user simply clicks the code within the display box and presses delete, which moves the code to the selection menu.
-The box therefore displays all codes currently included in the data displayed.
-\subsection{Auxiliary information}
-\TODO{e.g. Clinton's foreign schedule}
-A number of cosmetic adjustments using auxiliary information can also be made.
-Users can choose to adjust the scatterplot of time by day, which re-defines 6 pm as the time at which the day changes.
-This simply groups the emails best, placing all emails into one mostly uninterrupted region by redefining the day about the midsection of the region of lowest email traffic.
-Another cosmetic adjustment is the inclusion of Clinton's foreign trip schedule as blue regions on the email volume plot.
-This schedule was taken from the official State Department website \cite{ForeignSched} and allows users to distinguish periods spent out of the United States (coloured in blue) from those spent within the United States (uncoloured).
-\section{Some further exploration}
-\TODO{Some of our findings.  Unless done in the previous section}
-\section{Shiny App}
-\TODO{Again short and to the point}
-The \texttt{R} \texttt{shiny} app at (HOST ADDRESS) provides the ability for users to interactively filter the data on all features described above and view the resulting displays.
-Alongside this display is a short article explaining the use of the app, providing links to background information, and outlining some of the more interesting trends observed.
-It is hosted on the Waterloo \texttt{rshiny} server for public access, allowing for any individual to make use of this tool to generate hypotheses and investigate this intriguing dataset.
-
+\subsection{Highest peak}
 Historically important,  possibly had an impact on the 2016 US presidential election.
@@
 
-\section{Exposition}
+\section{Old stuff}
+\TODO{This material is to be reused wherever}
+An interactive web application in \texttt{R} \texttt{shiny} was constructed utilizing these visualizations and the extracted data.
+Incorporated in this app was a short article explaining the use of the app, providing links to background information, and outlining some of the findings of the more interesting trends observed.
+This framework provides any individual with interest the ability to generate hypotheses by exploring the data without necessarily searching for any particular conclusion.
+
+In generating and exploring this data, a considerable amount of respect is gained for the power of the often overlooked metadata of emails, especially with public figures.
+Simply having access to the network of communications, simple indicators of identifiable features of the individuals communicating, and times sent is enough to generate interesting questions which simple Google searches of important events and days can shed a great deal of light on.
+Leveraging data beyond this metadata, including the Freedom of Information Act (FOIA) exemption codes (see https://vault.fbi.gov/explanation-of-exemptions) used to redact the emails, only bolsters this simple metadata driven investigation.
+Metadata is often viewed as less significant or somehow less intrusive data, but the exploration performed in this paper demonstrated, at the very least to the authors, what a powerful tool metadata is for generating hypotheses about data.
+Given the lack of privacy present in the modern internet age and the amount of information most individuals make publicly available, there is a clear warning here about the utility of metadata for bad or for good.
+For a public official like Hillary Clinton, the data become even more interesting due to a number of contentious issues which arose during her tenure as the Secretary of State, independent of and as a result of her use of a private server.
+\section{Concluding Remarks}
+In providing the service, we purposely focused the displays (for the most part) on simple characteristics of the emails, the so-called metadata.  There are several related reasons for this focus.
+
+First, metadata is often more reliable, regular, and, being easily collected, more generally available.
+For Secretary Clinton's emails,  some metadata is lost, either because their source is a printed form or because it has been redacted.
+Examining metadata also arguably intrudes least upon the privacy of the individual correspondents, compared at least to the email's content.
+
+Second, since at least the Snowden revelations (e.g.
+\cite{NYRsnowdenLeaks}), public discourse has grown on the potential value of metadata to those who have it.
+By providing a web service where anyone can see for themselves what might be learned from metadata, we hope to contribute positively to this important discussion.
+Moreover, the work-related only metadata of a public figure as senior as Secretary Clinton will hopefully resonate more strongly with other public figures  engaged in the debate (e.g.
+\cite{NYRmetadata, ObamaMetadata, JebBushMetadata2015, PompeoMetadata, TrumpMetadata}) than might that of an ordinary citizen.
+
+Finally, it is important that users realize that the value of metadata is not just in itself, where one might expect to easily discern general day-to-day habits such as one's circle of correspondents.
+Rather it is that when coupled with other sources, which are abundantly and publicly available for Senator Clinton, much more can be learned than from any one source alone.
+Those who have knowledge of, or access to, other sources may filter metadata to test previously held hypothesis; conversely, exploration of metadata could uncover patterns that generated hypotheses to be tested elsewhere.
 

### 2017-03-28 VAST_submission/vgtc_journal_latex/paper.tex
Changed Chris-specific TODOs

@@
 The raw HTML of each message was programmatically downloaded from the Wikileaks archive using \texttt{R} packages  \texttt{RCurl}~\cite{RCurl2016package} and \texttt{XML}~\cite{XML2016package}.
-This took several hours and required \TODO{Chris: mentioned etiquette concerns here?
-Timeout?}
+This took several hours and required the use of manually inserted system pauses to prevent timeouts in the connection, most likely due to Wikileaks DDoS protection software.
+Besides avoiding such timeouts, these pauses are considered web-scraping etiquette and best practice.
 Once downloaded, the HTML data were processed using the \texttt{R} packages \texttt{tm}~\cite{tm2008paper, tm2017package}, \texttt{stringr}~\cite{stringr2016package}, and \texttt{SnowballC}~\cite{snowballc2014package}.
@@
 For each email, from the HTML we extract as best we can the identity of the persons sending and receiving the email as well as the date and time at which the email was processed by the server.
-\TODO{Chris:  Can you mention which fields we did or did not actually use?
-i.e.
-To: From" CC: BCC: etc.?}
+The fields used were the address to, address from, contact name to, contact name from, subject line, and time.
+As well, forwarding chains of email addresses were captured through the identification of any to or from fields followed by emails within the text.
+In cases where no contact name was present the address was substituted.
+When the address was missing no imputation was completed.
+Carbon copy information was not extracted.
 
@@
 Moreover, because of redaction (typically FOIA exemption B6 ~\cite{FOIA}),  sender and receiver emails may have truncated domains,  contain only the person's name, or be missing altogether.
-In cases where both \texttt{From} and \texttt{To} are entirely missing, but there is an email chain within the message, we  \TODO{Chris?
-What happens?}  (e.g.
+In cases where both \texttt{From} and \texttt{To} are entirely missing, but there is an email chain within the message, we do not impute values.
+\TODO{I could implement the earlier mentioned forward chain extraction, I thought I had already implemented it but it seems I did not}  (e.g.
 see \texttt{https://wikileaks.org/clinton-emails/emailid/31599}).
@@
 To provide some coarse statistical summaries of the content, all word tokens are extracted and partially processed to reduce their number.
-For example,  ``stopwords'', as identified by the\texttt{stopwords} function from  were removed.
-\TODO{Chris: I see from EmailFunctions.R that you only use the default stopwords() and an additional list.
-Maybe you should use union(stopwords(), stopwords(kind="SMART"), yourList) ?}  The remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.
+For example,  ``stopwords'', as identified by the\texttt{stopwords} function under either the "en" and "SMART" settings, or a set of custom stopwords, were removed.
+The remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.
 While neither the stopword lists nor the stemming tools are particularly well tuned to this corpus of emails, they nevertheless provide some insight as to the topics covered.

### 2017-03-28 VAST_submission/vgtc_journal_latex/paper.tex
Fleshed out the general outline of the paper, added some subsection labels and references

@@
 
-The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} ... 
-\TODO{paper overview now}
+The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} which discusses the source of the data, the extraction and cleaning process in Subsection \ref{sect:data:extrclean}, the details of the metadata in Subsection \ref{sect:data:metadata}, the use of redaction codes in Subsection \ref{sect:data:redact} and content data used in Subsection \ref{sect:data:contentinfo}.
+The displays of the data and their patterns over the entire unfiltered data are then described in Section \ref{sect:Displays}.
+Within this section, Subsection \ref{sect:Displays:circle} describes the spoked network plot, Subsection \ref{sect:Displays:volume} addresses the email volume time series plot, Subsection \ref{sect:Displays:times} discusses the scatterplot of daily sending times, Subsection \ref{sect:Displays:FOIA} outlines the exemption code barplot, and Subsection \ref{sect:Displays:TFIDF} explains the content tf-idf and frequency information.
+Interactive filters which are implemented in the application are discussed in Section \ref{sect:Filters}.
+Finally, brief analysis of some interesting findings is included in Section \ref{sect:Analysis}, alongside a summary of the timeline of the Clinton email imbroglio in Subsection \ref{sect:Analysis:Timeline}.
 
@@
 \subsection{Data extraction and cleaning}
+\label{sect:data:extrclean}
 The raw HTML of each message was programmatically downloaded from the Wikileaks archive using \texttt{R} packages  \texttt{RCurl}~\cite{RCurl2016package} and \texttt{XML}~\cite{XML2016package}.
-This took several hours and required the use of manually inserted system pauses to prevent timeouts in the connection, most likely due to Wikileaks DDoS protection software.
-Besides avoiding such timeouts, these pauses are considered web-scraping etiquette and best practice.
+This took several hours and required the use of manually inserted system pauses to prevent time-outs in the connection, most likely due to Wikileaks DDoS (distributed denial of service) protection software.
+Besides avoiding such DDoS protection, these pauses are considered web-scraping etiquette and best practice.
 Once downloaded, the HTML data were processed using the \texttt{R} packages \texttt{tm}~\cite{tm2008paper, tm2017package}, \texttt{stringr}~\cite{stringr2016package}, and \texttt{SnowballC}~\cite{snowballc2014package}.
@@
 \subsection{Metadata}
+\label{sect:data:metadata}
 For each email, from the HTML we extract as best we can the identity of the persons sending and receiving the email as well as the date and time at which the email was processed by the server.
@@
 \subsection{Redaction information}
+\label{sect:data:redact}
 Emails that are redacted are marked as such by the string ``RELEASE IN PART'' and by the presence of one or more of nine FOIA exemption codes B1, B2, \ldots, B9  marking the place where text is missing (redacted).
@@
 \subsection{Content information}
+\label{sect:data:contentinfo}
 The entire (redacted) content of each email is available on the State Department and Wikileaks sites and the user is encouraged to view it there (the pdf forms are more informative, especially in appreciating the extent of the redactions).
@@
 \subsection{Inner circle}
+\label{sect:Displays:circle} 
 The spokes are separated into two groups by the radial length of the spokes, these groups are determined automatically by the largest difference in volume of correspondence present in the data between the ordered counts of number of emails sent between individuals.
@@
 \subsection{Email volume}
+\label{sect:Displays:volume}
 This plot has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day.
@@
 \subsection{Email times}
+\label{sect:Displays:times}
 This plot takes advantage of the time-stamps on every email and then plots the emails as points with times on the vertical axis and the day sent on the horizontal axis.
@@
 \subsection{FOIA exemptions}
+\label{sect:Displays:FOIA}
 The United States Freedom of Information Act (FOIA) exemption codes \cite{FOIA} are used by the United States Federal Government to remove sensitive information from released documents accessed using a Freedom of Information Act request.
@@
 \subsection{Term frequency and TFIDF}
+\label{sect:Displays:TFIDF}
 The term frequency and tf-idf displays for this data are perhaps the least directly useful.

### 2017-03-29 EmailFunctions.R
Added stopwords(kind = SMART) as suggested

@@ -519,6 +519,9 @@ l_layer_line(mail, x =rep(as.numeric(ASDays[ASDays == "09/11/12"]), 2),
 # alternatively, we can use a double-sided slider and basic R, as shown below
 
 ### Keyword Analysis ####################################################################
+ExtraWords <- c("will", "state", "pm", "said", "secretari", "can", "call",
+                "depart", "time", "presid", "govern", "offic", "work", "usa", "also", "meet",
+                "new", tolower(LETTERS))
 # define a function which can be used to clean and process email content
 get_keywords <- function(Content, stem = TRUE, table = TRUE) {
   # first combine all the emails into one string
@@ -543,10 +546,8 @@ get_keywords <- function(Content, stem = TRUE, table = TRUE) {
   # split by singular spaces
   corpus <- unlist(str_split(corpus, " "))
   # remove common stopwords
-  Cust_stop <- c(stopwords(), "will", "state", "pm", "said", "secretari", "can", "call",
-                 "depart", "time", "presid", "govern", "offic", "work", "usa", "also", "meet",
-                 "new", tolower(LETTERS))
-  corpus <- removeWords(corpus, stopwords())
+  Cust_stop <- c(stopwords(), stopwords(kind = "SMART"), ExtraWords)
+  corpus <- removeWords(corpus, Cust_stop)
   # remove empty strings which may have occurred
   corpus <- corpus[corpus != ""]
   # stem the remaining terms

### 2017-03-29 VAST_submission/vgtc_journal_latex/paper.tex
Added explanations below images in the 'displays' section, added images to the 'filter' section

@@
 As well, the width of the edges of the graph is determined by the volume of correspondence which occurred between Clinton and the correspondent in either direction.
+For brevity, the plot displays nodes for only the top 20 correspondents at most.
 Finally, all edges and points are coloured according to a simple scheme: if the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, their node is coloured blue; nodes with email addresses which are identifiably not hosted at a '.gov' domain are coloured red; nodes with '.mil' accounts are coloured black; and nodes unidentifiable hosts are coloured orange.
@@
 \end{center}
+
+While most of Clinton's top 20 correspondents have identifiable state accounts, three do not.
+One of these, Sidney Blumenthal, is of particular interest, as the advisory nature of his communications with Clinton despite lacking official state department clearance has already generated significant controversy \cite{BlumenthalControversy}.
 
@@
 \end{center}
+
+This plot shows highly variable nature of the daily email volumes, with a number of peaks and troughs suggesting regions in which further investigation might be insightful.
 
@@
 
+It is clear that redactions made for national security and foreign policy reasons account for only a small portion of redactions made.
+The most frequent redactions are those made to protect personal privacy, which occur in almost half of all emails, and those made with the vague B5 exemption code, which occur in roughly a quarter of all emails.
+The other codes appear in very few emails and just over a third of the emails are not redacted at all.
+
 \subsection{Term frequency and TFIDF}
@@
 Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
+\includegraphics{DateSliderImage}
 \subsection{Sent or received}
 A drop down menu allows users to select emails by those sent to Clinton, sent by Clinton, or by the entire email database.
+\includegraphics{ToFromImage}
 \subsection{FOIA exemptions}
 Multiple filtering by FOIA exemption codes is supported in the app with a box selection area.
-Users can add FOIA exemption codes currently not included to the data by clicking the box with their mouse and selecting a code from the menu which is then displayed below the box.
+Users can add FOIA exemption codes currently not included to the data by clicking the box and selecting a code from the menu which is then displayed below the box.
 To remove codes, the user simply clicks the code within the display box and presses delete, which moves the code to the selection menu.
-The box therefore displays all codes currently included in the data displayed.
+The box therefore displays all codes currently included in the filter.
+\includegraphics{FOIABoxImage}
 \subsection{Auxiliary information}

### 2017-03-29 VAST_submission/vgtc_journal_latex/paper.tex
Small sentence change

@@
 \label{sect:Filters}
-\TODO{Each filter is described and displayed }
 \subsection{Time filtering}
@@
 \subsection{Highest peak}
+Within the plot of email volume over time, one very large peak stands out.
 Historically important,  possibly had an impact on the 2016 US presidential election.

### 2017-03-29 EmailFunctions.R
Minor tidying up after experiments with faster filter. Determined that load time was the real time issue with the app

@@ -9,7 +9,7 @@ library(dplyr)
 library(loon)
 
 # load the raw HTML data if desired
-fullmail <- readRDS('ClintonsEmails')
+fullmail <- readRDS('fullmail_full')
 
 ### Email Extraction Functions #########################################################
 # generate a function to clean up emails using the library chron

### 2017-03-29 app.R

@@ -50,8 +50,8 @@ ClintNet <- sort(table(c(as.character(ClintonCom$To.name),
                  decreasing = TRUE)[-1]
 # extract state mail information, information pertaining to whether an email is .gov or not
 # first get all names from above network
-allnames <- levels(as.factor(c(levels(ClintonCom$To.name),
-                               levels(ClintonCom$From.name))))
+allnames <- levels(as.factor(c(levels(as.factor(ClintonCom$To.name)),
+                               levels(as.factor(ClintonCom$From.name)))))
 # generate storage in advance
 stateMail <- rep(0, length(allnames))
 names(stateMail) <- allnames
@@ -197,9 +197,9 @@ ui <- fluidPage(
                                  c("From Clinton", "To Clinton", "All Emails"),
                                  selected = "All Emails", multiple = FALSE)),
             fluidRow(selectInput("ClassFilter", "FOIA Codes: ",
-                                 c("None", "B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", "B9"),
-                                 selected = c("None", "B1", "B2", "B3", "B4", "B5", "B6", "B7",
-                                              "B8", "B9"),
+                                 c("B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", "B9", "None"),
+                                 selected = c("B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", 
+                                              "B9", "None"),
                                  multiple = TRUE)),
             fluidRow(h4("20 Highest TFIDF Terms")),
             fluidRow(textOutput("tfidf")),
@@ -255,6 +255,7 @@ server <- function(input, output) {
   })
   # memoise the above function for caching
   inter <- inputProc(input)
+  # now define all outputs
    output$Times <- renderPlot({
      # first determine whether adjusted or unadjusted times are to be used
      if (inter()$adjVar) {
@@ -263,8 +264,7 @@ server <- function(input, output) {
        xlab <- "Adjusted Date (dd/mm/yy)"
        main <- paste("Adjusted Email", inter()$toFromLab, "Times")
        labelset <- c('22:00', '02:00', '06:00', '10:00', '14:00')
-     }
-     else {
+     } else {
        timevalues <- AsSec$Hour[AsSec$ID %in% inter()$selIDs]*60 + 
          AsSec$Minutes[AsSec$ID %in% inter()$selIDs]
        xlab <- "Date (dd/mm/yy)"

### 2017-03-30 VAST_submission/vgtc_journal_latex/paper.tex
Still writing

@@
   \caption{Metadata: Time, date, redaction, correspondents, volume, source type.  Filtered: Sept. 11 to Nov. 23, 2012.  
-  %Source: \href{https://wikileaks.org/clinton-emails/}{\texttt{https://wikileaks.org/clinton-emails/}}
   }
@@
 Moreover, because of redaction (typically FOIA exemption B6 ~\cite{FOIA}),  sender and receiver emails may have truncated domains,  contain only the person's name, or be missing altogether.
-In cases where both \texttt{From} and \texttt{To} are entirely missing, but there is an email chain within the message, we do not impute values.
-\TODO{I could implement the earlier mentioned forward chain extraction, I thought I had already implemented it but it seems I did not}  (e.g.
+In cases where both \texttt{From} and \texttt{To} are entirely missing, but there is an email chain within the message, we do not impute values  (e.g.
 see \texttt{https://wikileaks.org/clinton-emails/emailid/31599}).
@@
 \label{sect:Displays:circle} 
-The spokes are separated into two groups by the radial length of the spokes, these groups are determined automatically by the largest difference in volume of correspondence present in the data between the ordered counts of number of emails sent between individuals.
-As well, the width of the edges of the graph is determined by the volume of correspondence which occurred between Clinton and the correspondent in either direction.
-For brevity, the plot displays nodes for only the top 20 correspondents at most.
-Finally, all edges and points are coloured according to a simple scheme: if the name associated with a node in the graph is at any point associated with an email address hosted at some '.gov' domain in the correspondences, their node is coloured blue; nodes with email addresses which are identifiably not hosted at a '.gov' domain are coloured red; nodes with '.mil' accounts are coloured black; and nodes unidentifiable hosts are coloured orange.
-It is important to note that for those individuals with orange spokes, the most likely reason the email is not identifiable is the redaction of the email by the state department.
-Such a redaction is not a guarantee that the email of the individual is not an official state email, but it is an indication that the email address is viewed as personal or private information, which suggests some email hosted outside of the '.gov' domain.
+This display takes up to the 20 most frequent correspondents in the selected emails and arranges their email addresses  at the ends of  equiangular spokes around the hub that is Secretary Clinton's email.
+The hub is coloured red to show that this account is known not to be a government sponsored account (either \texttt{.gov} of \texttt{.mil}).
+Every account that is identifiably not a government sponsored account is coloured red.
+Those that are identifiably government are coloured either blue (for \texttt{.gov}) or black (for \texttt{.mil}, if any appear).
+Those which can not be determined as either government or not, are coloured orange.
 
+\begin{figure}[h]
 \begin{center}
 \includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
+\caption{Inner Circle over all correspondence}
+\label{fig:InnerCircleAll}
 \end{center}
+\end{figure}
+The figure shows the top 20 (or ``Inner Circle'')  hereover all emails in the collection.
+Note that Clinton aide Huma Abedin is seen to be using both a \texttt{.gov} account for some email and a \texttt{clintonemail} account for others and so appears as two different colour spokes.
+The other identfiably non-government account is that of Sidney Blumenthal.
+This is of particular interest, as his closeness to Clinton in providing advice on all kinds of matter has been the source of controversy \cite{BlumenthalControversy} given that he had been refused a State department position and so lacked  State Department clearance.
+Wendy Sherman, though appointed by Clinton to  Under Secretary of State for Political Affairs in 2011, her email address could not be identified as either definitively government or definitively non-government and so is shown in orange.
+It is important to note, however, that the most likely reason an email is not identifiable is that it has been redacted (likely B6).
+While no guarantee that the email is not a government email, it is more likely that it was viewed as personal or private information and so probably not government.
 
-While most of Clinton's top 20 correspondents have identifiable state accounts, three do not.
-One of these, Sidney Blumenthal, is of particular interest, as the advisory nature of his communications with Clinton despite lacking official state department clearance has already generated significant controversy \cite{BlumenthalControversy}.
-
+The greater the number of emails between Clinton and the correspondent, the wider is the spoke.
+Whenever the difference in volume is great enough the correspondents will also be separated into two groups.
+Those with the greatest correspondence will have shorter spokes and hence be visually ``closer'' to Clinton.
+As the display shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
+\TODO{Chris, can you be more precise about the algorithm?}
 \subsection{Email volume}
 \label{sect:Displays:volume}
-This plot has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day.
+Figure~\ref{fig:VolumeAll} has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day.
 This allows users to view which days have an exceptionally high or low number of classified conversations.
@@
 
+\begin{figure}[h]
 \begin{center}
 \includegraphics[width=0.95\linewidth]{VolumeFullTime}
+\caption{Email Volume: All emails (top) ; Sent by Clinton (bottom)}
+\label{fig:VolumeAll}
 \end{center}
+\end{figure}
 
@@
 \label{sect:Displays:times}
-This plot takes advantage of the time-stamps on every email and then plots the emails as points with times on the vertical axis and the day sent on the horizontal axis.
-The points in this scatterplot are then coloured according to whether the corresponding email was redacted, in which case the point is coloured red, or released in full, in which case the point is coloured blue.
-Alpha blending and transparency of these points was utilized to make the patterns present more obvious and prevent overplotting from obscuring any interesting results.
-
+As shown in Figure~\ref{fig:TimesAll}, this display shows the time stamp of each email separated into its calendar date on the horizontal axis and its (24hr) time of day on the vertical.  
+\begin{figure}[h]
 \begin{center}
 \includegraphics[width=0.95\linewidth]{DailyFullTime}
+\caption{Email time patterns: all emails}
+\label{fig:TimesAll}
 \end{center}
+\end{figure}
+Each email is a point in the plot and is coloured red if any part of that email was redacted, and blue if it was released in full.   Alpha blending prevents over-plotting from obscuring patterns.
 
-One such result which is immediately obvious is the strange pattern of sending times.
-Email communications occur with the highest frequency between midnight and 3 am, and with the lowest frequency between 4 pm and 10 pm.
-This pattern, which is persistent through filters of emails sent to Clinton and by Clinton, is surprising as it indicates that the highest level of communication was occurring at a time typically devoted to sleep.
-Another pattern immediately obvious is the alternating modes at 2 and 3 am, with alternations occurring on the dates of daylight savings in North America.
-The regularity of this pattern in the emails suggests some server function performed early in the morning, perhaps some maintenance operation performed every 24 hours.
+A few patterns are easily discerned.
+For example, the least amount of email appears for a few hours around  20:00 hours, or 8 PM.
+Evenings appear to be when email traffic is lightest.
+Otherwise, for the most part, it seems fairly uniformly distributed throughout.
+Midnight is at the top (and bottom) of the plot, so it would seem that email will occur mainly from midnight until about 4 or 5 PM.
+Note also the surprisingly regular horizontal lines that appear across the top of the plot.
+This regularity is suggestive of some automated schedule for the server which causes the email to stack up before being recorded at either 2 or 3 AM.
+The location of these lines switches exactly whenever daylight savings time switches in North America.
 
@@
 \label{sect:Displays:TFIDF}
-The term frequency and tf-idf displays for this data are perhaps the least directly useful.
-As mentioned above, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} proved inadequate for the messy and non-standard data.
-In particular, the omission of many common and uninformative action verbs such as 'will' and 'can' from the stopwords used led to very unsatisfactory results.
-A set of custom stopwords was therefore applied in an attempt to remove some of these uninformative words.
-The tf-idf was more informative, but the stemming algorithm, when coupled with the typos and abbreviations which are incredibly common in the emails, meant that the results are still incredibly unclear.
-The terms with the highest scores over any interval do not seem to communicate anything particularly useful about the content of emails sent in that range, with one notable exception in the case of the term 'windrush.' This term appeared with a high tf-idf in a number of selections which contained communications with Tony Blair (email contact "aclb" in the spiral plot), and seems to refer to Windrush Ventures, a company owned by Blair which appears vital in managing Blair's complicated and opaque finances since he left the office of the Prime Minister of the United Kingdom.
-\cite{windrushTelegraph, windrushGuardian}
+The 20 terms (stemmed words excluding stopwords) which appear in the greatest number of the emails selected are displayed as those having ``Highest Frequency''.  
+\begin{figure}[h]
+\begin{center}
+\begin{tabular}{ccc}
+\includegraphics[width=0.4\linewidth]{HighFreqAll} &
+$~~~~$ &
+\includegraphics[width=0.4\linewidth]{HighTFIDFAll} 
+\end{tabular}
+\end{center}
+\end{figure}
+Those terms which appear frequently within some emails but less frequently across emails will have a high tf-idf (term frequency - inverse document frequency) score.
+The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display.
+\TODO{Chris, check that I have described these two measures correctly}
+The top 20 shown here are for all emails in the corpus; these will change depending on the filtering.
+
+The terms give some limited insight into the contents of the email.
+As seen here, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.
+One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies which could magnify the within email frequency for some terms.
+
+As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is also challenged by this messy and non-standard data.
+The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'.
+The stemming algorithm was also challenged by typographical errors in emails and the proliferation of acronyms (e.g.
+for individuals and government abbreviations).
+Even so,  the terms nevertheless occasionally turn up something of interest as in the case of ``windrush'',  which turned out to refer to ``Windrush Ventures'' (e.g.
+see  \cite{windrushTelegraph, windrushGuardian}),  a company owned by former U.K.
+Prime Minister Tony Blair .
+``Windrush'' appears at the end of emails as part of Mr.
+Blair's electronic signature for his email id ``aclb''  (e.g.
+see the ``Inner Circle''  of Figure~\ref{fig:teaser}, b).
 
@@
 \label{sect:Filters}
+There are only three filters but each one affects all data displays.
+As each filter is changed, every display redraws itself on the filtered data, immediately in reaction to the change.
+In this way, the filters can be used together to focus on particular subsets of the emails or simply to observe how the display patterns change with the filter being applied.
 \subsection{Time filtering}
-Time filtering within the app can be completed at the level of individual days.
+A sliding time window 
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.35\linewidth]{DateSliderImage}
+\end{center}
+\end{figure}
+filters the emails displayed to those whose date lies between the two end points.
+The size of the time window is changed by  moving either end, either by mouse click and drag or by selecting an end to move with the arrow keys.
+The whole time window is moved by selecting the middle bar and either dragging it or moving it with the arrow keys.
+
+This filter is simple but powerful.
+All emails are displayed when the range covers all dates.
+Moving the end points towards one another allows the user to focus the displays on any particular range, down to as fine as all emails on a single day.
+With a fixed range of days, for example a two week period,  dragging the middle bar from left to right will have each display smoothly update over time.
+%
+%Time filtering within the app can be completed at the level of individual days.
 The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days.
@@
 Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
-\includegraphics{DateSliderImage}
 \subsection{Sent or received}
-A drop down menu allows users to select emails by those sent to Clinton, sent by Clinton, or by the entire email database.
-\includegraphics{ToFromImage}
+A drop down menu 
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.35\linewidth]{ToFromImage}
+\end{center}
+\end{figure}
+is used to select those emails which Clinton sent, or received, or either sent or received.
+The default is ``All Emails'', which is the same as no filter being applied for sent or received.
+Choosing ``From Clinton'' selects for display only those mails sent by Secretary Clinton, choosing ``To  Clinton'' selects only those emails which she received.
+This might be used, for example, to explore whether the inner circle of correspondents changes depending on whether Secretary Clinton is emailing them, or they are emailing her.
+
 \subsection{FOIA exemptions}
 Multiple filtering by FOIA exemption codes is supported in the app with a box selection area. 
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.35\linewidth]{FOIABoxImage}
+\end{center}
+\end{figure}
 Users can add FOIA exemption codes currently not included to the data by clicking the box and selecting a code from the menu which is then displayed below the box.
@@
 The box therefore displays all codes currently included in the filter.
-\includegraphics{FOIABoxImage}
 \subsection{Auxiliary information}
@@
 \newpage
+$~~~$
+\newpage
+
 \bibliographystyle{abbrv}

### 2017-03-31 VAST_submission/vgtc_journal_latex/paper.tex
Added zero dates
Added my edits

@@
 The greater the number of emails between Clinton and the correspondent, the wider is the spoke.
-Whenever the difference in volume is great enough the correspondents will also be separated into two groups.
-Those with the greatest correspondence will have shorter spokes and hence be visually ``closer'' to Clinton.
+Spoke width is scaled linearly relative to the greatest number of emails for all correspondents in the selection, with a constant minimum width for visibility.
+The correspondents are also separated into two groups by the greatest difference in ordered email volumes.
+Those which appear before this difference when all are ordered by decreasing email volume have nodes placed at half the distance from the centre.
+will have shorter spokes and hence be visually ``closer'' to Clinton.
 As the display shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
@@
 The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display.
-\TODO{Chris, check that I have described these two measures correctly}
 The top 20 shown here are for all emails in the corpus; these will change depending on the filtering.
@@
 \subsection{Highest peak}
-Within the plot of email volume over time, one very large peak stands out.
-Historically important,  possibly had an impact on the 2016 US presidential election.
+Within the plot of email volume over time, one very large peak stands out in the later half of 2011.
+Using the time filtering function the date of this peak can be identified as August 21, 2011.
+A quick internet search of this date reveals that it was the start of the Battle of Tripoli, the final battle in the 2011 Libyan Civil War in which the NATO-backed National Transitional Council \cite{NATONTCSupport,LibyaContactGroup} toppled the regime of Muammar Gaddafi.
+\cite{BattleforTripoli,AlJazeeraTripoli} This suggests that the spike in emails was a response by the State Department as a whole to the news of the National Transitional Council's final attacks on Tripoli.
+The tf-idf and term frequencies over this time frame support the suggestion that this event was the focus of most of the communication.
+
+Terms included in this selection such as "tnc" (the abbreviation for the National Transitional Council rebels used by Clinton), "jalil" (the head of the National Transitional Council \cite{Jalil}), "libya", "tripoli", and "qadhafi" clearly refer to the Libyan Civil War.
+Of greater interest are some of the less clearly connected terms, such as "davutoglu," "tarhouni," and "jupp." Searching for these terms gives some immediate and obvious connections.
+"Davutoglu" appears to refer to Ahmet Davutoglu, the Turkish Minister of Foreign Affairs at the time.
+"Jupp" seems to be a reference to Alain Juppe, the French Minister of Foreign Affairs.
+Finally, "tarhouni" likely refers to Ali Tarhouni, the Minister for Oil and Finance within the National Transitional Council.
+The mention of these particular names gives insight into the individuals the State Department likely viewed as most significant during the Battle of Tripoli and transition of the Libyan government.
+
+\subsection{No email sent by Clinton}
+
+Using 1 month window slider and looking for flat spots at zero
+
+Dates:
+\begin{itemize}
+\item None until mid April 2009
+\item June 8 or 9, June 15 to about the 20th, 2009
+\item July 17-20,  around 27, 2009 
+\item Day or 2 around Oct 13, 14, 2009
+\item April 10-15 or so? 2011
+\item one or two days late August, 2011. (no mail)
+\item end of march first half of April 2012
+\item August 27-30  2012 
+\item low first week of September, 2012
+\item 2 weeks end of october, up to Nov 10 approx 2012
+\item Dec 7- 17? 2012,  Again about Dec 25 2012 to Jan 1, 2013
+\end{itemize}
 

### 2017-03-31 VAST_submission/vgtc_journal_latex/paper.tex

--- 2017-03-30 VAST_submission/vgtc_journal_latex/paper.tex
@@
   &\\
-  {\small (a) Missing block of email: October 30 - November 9} & {\small (b) Inner Circle: Correspondent info by colour; volume by thickness.}
+  {\small (a) An empty email block: October 30 - November 9} & {\small (b) Inner Circle: Correspondent info by colour; volume by thickness.}
   \end{tabular}
@@
 
-In this paper, we describe an implementation of the missing visual analytic service and illustrate it on Secretary Clinton's e-mails.
+In this paper, we describe an implementation of such a visual analytic service and illustrate it on the Secretary Clinton's e-mail archive.
 The implementation is located at  \texttt{rshiny.math.uwaterloo.ca/clinton} where any visitor may interactively explore summary features  of the entire corpus of emails.
@@
 Interactive filters which are implemented in the application are discussed in Section \ref{sect:Filters}.
-Finally, brief analysis of some interesting findings is included in Section \ref{sect:Analysis}, alongside a summary of the timeline of the Clinton email imbroglio in Subsection \ref{sect:Analysis:Timeline}.
+Finally, brief analysis of some interesting findings is included in Section \ref{sect:Analysis}, alongside a summary of the timeline of the Clinton email imbroglio in Section \ref{sect:Timeline}.
+
+%\section{Background}
+%\label{sect:Background}
+%
+%\subsection{A brief timeline on the private email server}
+%\label{sect:Background:Timeline}
+\section{A brief timeline on the private email server}
+\label{sect:Timeline}
+While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to the private email server and its contents.
+It also introduces some of the principal characters involved.
+More complete and in depth timelines are readily available elsewhere (e.g.
+\cite{attkissonTimeline, thompsonTimeline, WashPostTimeline, clintonWikipedia, TimeMagEverything}).
+
+On November 21, 2008, the New York Times reported that Hillary Clinton had decided to accept the position of U.S.
+Secretary of State.
+On January 13, 2009 the internet domain name \texttt{clintonemail.com} was registered \cite{whoisClintonserver}; eight days later Senator Clinton was confirmed as Secretary of State.
+
+Public knowledge that a private email server being used by Secretary Clinton and others for State Department and personal communications did not surface until March 2015 \cite{NewYorkTimes2015} during the course of a U.S.
+Congressional investigation \cite{BenghaziReport} of the September 11, 2012 attack by militants on U.S.
+compounds in Benghazi Libya.
+
+The State Department had difficulty fulfilling public FOIA and House Benghazi Committee requests \cite{TakingRootWashPost} for Secretary Clinton's government emails because she had exclusively used the private server for all her email.
+On March 10, 2015,  Clinton told reporters that she turned over 30,490 emails to the State Department and deleted 31,830 emails deemed to be personal \cite{WashPostTimeline}.
+Clinton had tasked three lawyers Cheryl Mills (Clinton's former chief of staff),  David Kendall (Clinton's personal lawyer), and Heather Samuelson (a State Department staffer during Clinton's tenure) to make the determinations as to which emails were work related and which were not \cite{emailVetting, thompsonTimeline}.
+
+On March 10, 2015 the House Benghazi Committee requested that the private email server be turned over to a neutral third party to determine which emails are personal and which are government records \cite{serverRequest}, but was informed March 27 by David Kendall that no emails remained on the private server for any kind of review \cite{serverScrubbedLawyer}.
+Between March 25 and 31, 2015, Paul Combetta (then the server's system administrator), erased all backup copies using \texttt{BleachBit} (see \texttt{www.bleachbit.org}).
+
+Combetta, Mills, and Samuelson will later be granted partial immunity by the Justice Department during the FBI investigations into the private email server, as were two others: Bryan Pagliano (original server manager) and  John Bentel (former director of Information Resources Management for the State Department's Executive Secretariat) \cite{immunityPolitico, immunityDailyCaller, immunityIT}.
+
+On April 12, 2015 Clinton announces that she is running for the U.S.
+Presidency.
+On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce finding classified information in the emails and that the information they found was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact.
+On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.
+
+Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton on use of private email server \cite{nochargeFBI}.
+On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin have been found on his laptop.
+Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.
+In both cases, Comey and the FBI are criticized by pundits from different political parties.
 
@@
 To provide some coarse statistical summaries of the content, all word tokens are extracted and partially processed to reduce their number.
-For example,  ``stopwords'', as identified by the\texttt{stopwords} function under either the "en" and "SMART" settings, or a set of custom stopwords, were removed.
-The remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.
-While neither the stopword lists nor the stemming tools are particularly well tuned to this corpus of emails, they nevertheless provide some insight as to the topics covered.
+For example,  ``stopwords'', as identified by the \texttt{stopwords} function (with both  \texttt{kind = "en"} and  \texttt{kind = "SMART"}), or as from a set of custom stopwords, were removed.
+Remaining words were stemmed by the \texttt{stemDocument} function from \texttt{tm} and \texttt{SnowballC}.
+While neither the stopword lists nor the stemming tools are particularly well tuned to this corpus of emails, they nevertheless provide some hints about the topics covered.
 
@@
 \label{sect:Displays:circle} 
-This display takes up to the 20 most frequent correspondents in the selected emails and arranges their email addresses  at the ends of  equiangular spokes around the hub that is Secretary Clinton's email.
+This display takes the most frequent correspondents (max.
+20) in the selected emails and arranges their email addresses  at the ends of  equiangular spokes having Secretary Clinton as hub.
 The hub is coloured red to show that this account is known not to be a government sponsored account (either \texttt{.gov} of \texttt{.mil}).
@@
 \end{figure}
-The figure shows the top 20 (or ``Inner Circle'')  hereover all emails in the collection.
-Note that Clinton aide Huma Abedin is seen to be using both a \texttt{.gov} account for some email and a \texttt{clintonemail} account for others and so appears as two different colour spokes.
-The other identfiably non-government account is that of Sidney Blumenthal.
-This is of particular interest, as his closeness to Clinton in providing advice on all kinds of matter has been the source of controversy \cite{BlumenthalControversy} given that he had been refused a State department position and so lacked  State Department clearance.
-Wendy Sherman, though appointed by Clinton to  Under Secretary of State for Political Affairs in 2011, her email address could not be identified as either definitively government or definitively non-government and so is shown in orange.
-It is important to note, however, that the most likely reason an email is not identifiable is that it has been redacted (likely B6).
-While no guarantee that the email is not a government email, it is more likely that it was viewed as personal or private information and so probably not government.
+Figure~\ref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
+Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.
+The other red account is that of Sidney Blumenthal.
+This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to provide Clinton advice throughout her tenure.
+Wendy Sherman appears as orange.
+Sherman was appointed by Clinton to be Under Secretary of State for Political Affairs in 2011 and the high frequency of Sherman emails dates from this time.
+Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.
+Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided.
 
-The greater the number of emails between Clinton and the correspondent, the wider is the spoke.
-Whenever the difference in volume is great enough the correspondents will also be separated into two groups.
-Those with the greatest correspondence will have shorter spokes and hence be visually ``closer'' to Clinton.
-As the display shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
+The greater the number of emails between correspondents, the wider is their connecting spoke.
+Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.
+As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
 \TODO{Chris, can you be more precise about the algorithm?}
@@
 \label{sect:Displays:volume}
-Figure~\ref{fig:VolumeAll} has two lines, the first is a line showing the total volume of emails for each day, and the second shows the number of emails satisfying the classification filter criteria for a given day.
-This allows users to view which days have an exceptionally high or low number of classified conversations.
-Returning to the motivation of hypothesis generation, referencing world events on any days of interest provides a number of interesting insights on how public figures like Clinton react and respond to world events.
-
+Figure~\ref{fig:VolumeAll} 
 \begin{figure}[h]
@@
 \includegraphics[width=0.95\linewidth]{VolumeFullTime}
-\caption{Email Volume: All emails (top) ; Sent by Clinton (bottom)}
+\caption{Email Volume: All emails (top) ; Those sent by Clinton (bottom)}
 \label{fig:VolumeAll}
@@
 \end{figure}
+presents two time series showing the volume of emails for each day.
+In grey at top is the daily total; in magenta at bottom is the total for the subset chosen by the filters.
+In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails sent from Clinton to others.
 
-This plot shows highly variable nature of the daily email volumes, with a number of peaks and troughs suggesting regions in which further investigation might be insightful.
+Over the four year scale shown in Figure~\ref{fig:VolumeAll}, the display is quite busy.
+Even so, a few features stand out.
+For example, there is a notable lack of email from the first few months of Clinton's tenure;  there is essentially no email from Clinton at the beginning and very little to her.
+
+One also notices the spikes of email activity.
+These can be checked against world events to generate hypotheses about what might be occurring to cause such spikes.
+For example, the largest spike occurs on August 21, 2011, the beginning day of the battle for Tripoli and on which it was reported that two of Qaddafi's sons had been captured \cite{battleTripoli}.
+It is also the day on which the famous ``tick tock Libya'' email is composed by Jake Sullivan providing a timeline crediting Secretary Clinton with leading the U.S.
+policy on Libya ``from start to finish'' \cite{tickTockLibya}.
+Much of the email that day is redacted (B5 primarily).
+Other peaks (and valleys) of activity could be similarly investigated.
 
@@
 \label{sect:Displays:times}
-As shown in Figure~\ref{fig:TimesAll}, this display shows the time stamp of each email separated into its calendar date on the horizontal axis and its (24hr) time of day on the vertical.  
+Figure~\ref{fig:TimesAll}  
 \begin{figure}[h]
 \begin{center}
-\includegraphics[width=0.95\linewidth]{DailyFullTime}
-\caption{Email time patterns: all emails}
+\includegraphics[width=0.95\linewidth]{DailyFullTimeFromClinton}
+\caption{Email time patterns: all emails sent from Clinton}
 \label{fig:TimesAll}
@@
 \end{figure}
-Each email is a point in the plot and is coloured red if any part of that email was redacted, and blue if it was released in full.   Alpha blending prevents over-plotting from obscuring patterns.
+shows all emails sent by Clinton over the four year period.
+Each email appears as a point determined by the time shown that it was sent.
+The calendar date determines the position on the horizontal axis,  the (24hr) time of day determines the vertical.
+Points are coloured red if any part of that email was redacted, and blue if it was released in full.
+Alpha blending is used to minimize the effect of over-plotting.
 
@@
 Midnight is at the top (and bottom) of the plot, so it would seem that email will occur mainly from midnight until about 4 or 5 PM.
+
+If these are the actual times Secretary Clinton composed and sent email, it suggests that much of that activity occurred in the middle of the night.
+To reflect the low email activity period as actual ``downtime'' between ``days'' (now interpreted as separable 24 hour email cycles),  we offer the possibility of choosing 6 PM as the ``end'' of a day rather than midnight.
+This places the least active email period as a ``quiet'' boundary between days.
+Unfortunately, though it better represents the rhythm of correspondence, it can be confusing since it means that the ``next day''  begins just after 6 PM.
+
 Note also the surprisingly regular horizontal lines that appear across the top of the plot.
@@
 
+When the received emails are added to the plot, the pattern is essentially the same (though much denser) and the horizontal lines where the server has scheduled something are even clearer.
+
 \subsection{FOIA exemptions}
 \label{sect:Displays:FOIA}
-The United States Freedom of Information Act (FOIA) exemption codes \cite{FOIA} are used by the United States Federal Government to remove sensitive information from released documents accessed using a Freedom of Information Act request.
-The sensitive matters being addressed by the codes are national security and foreign policy matters for B1, personnel practices for B2, exempted by statute for B3, trade secrets or financial information obtained in confidence for B4, inter- or intra-agency memorandums for B5, personal privacy for B6, records compiled for law enforcement for B7, prepared in relation to financial monitoring institutions for B8, and geophysical and geological information concerning wells for B9 exemptions.
-The frequencies of each of these codes within filtered emails are shown in the final display.
-Note that as emails can have numerous redaction codes, this barplot can be useful to identify which codes co-occur over a selection.
+U.S.
+Freedom of Information Act (FOIA) exemption codes B1 through B9 are used in each place in an email where information has been redacted.
+For the authoritative definition the act itself should be consulted \cite{FOIA}.
+Briefly, they mark each redaction as B1 for national security and foreign policy matters,  B2 for personnel practices, B3 for statutory exemptions,  B4 for trade secrets or financial information obtained in confidence,  B5 for inter- or intra-agency memorandums,  B6 for personal privacy,  B7 for records compiled for law enforcement,  B8 for records prepared in relation to financial monitoring institutions, and B9 for geophysical and geological information concerning oil and gas wells.
+Each email can contain redactions for any number of exemption codes.
+
+Figure \ref{fig:FOIAcodes}
+\begin{figure}[h]
 \begin{center}
 \includegraphics[width=0.95\linewidth]{ExemptionsFullTime}
+\caption{FOIA barplot: Exemption codes from all email}
+\label{fig:FOIAcodes}
 \end{center}
+\end{figure}
+is a barplot of all FOIA exemption codes found in all of the emails.
+As can be seen, B1 redactions made for national security and foreign policy reasons account for a small portion of all redactions made though affect more than 2,000 emails in the collection.
+Nearly half of the emails contain B6 exemptions, those made to protect personal privacy.
+The inter-agency, intra-agency exemption B5 is used in more than 8,000 emails.
+This code has come under some criticism from advocates for transparent government as being overly applied \cite{unredactedB5}.
+The remaining codes appear in very few emails and about 12,000 emails, or about a third of them, are not redacted at all.
 
-It is clear that redactions made for national security and foreign policy reasons account for only a small portion of redactions made.
-The most frequent redactions are those made to protect personal privacy, which occur in almost half of all emails, and those made with the vague B5 exemption code, which occur in roughly a quarter of all emails.
-The other codes appear in very few emails and just over a third of the emails are not redacted at all.
+As with all other displays, the FOIA barplot will update in reaction to all filters making it a useful tool to identify how codes co-occur over any selection of emails.
+Conversely, filtering on FOIA codes can be interesting to see who sees correspondence of varying sensitivities, especially B1 and B5.
 
@@
 \end{tabular}
+\caption{Top 20 terms appearing in all emails}
+\label{fig:top20all}
 \end{center}
@@
 \TODO{Chris, check that I have described these two measures correctly}
-The top 20 shown here are for all emails in the corpus; these will change depending on the filtering.
+The top 20 shown in Figure~\ref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.
 
 The terms give some limited insight into the contents of the email.
-As seen here, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.
+As seen in Figure~\ref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.
 One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies which could magnify the within email frequency for some terms.
@@
 for individuals and government abbreviations).
-Even so,  the terms nevertheless occasionally turn up something of interest as in the case of ``windrush'',  which turned out to refer to ``Windrush Ventures'' (e.g.
-see  \cite{windrushTelegraph, windrushGuardian}),  a company owned by former U.K.
-Prime Minister Tony Blair .
-``Windrush'' appears at the end of emails as part of Mr.
-Blair's electronic signature for his email id ``aclb''  (e.g.
-see the ``Inner Circle''  of Figure~\ref{fig:teaser}, b).
 
@@
 \includegraphics[width=0.35\linewidth]{DateSliderImage}
+\caption{Time slider: move either end, or move the interval}
 \end{center}
+\label{fig:timeSlider}
 \end{figure}
@@
 \includegraphics[width=0.35\linewidth]{ToFromImage}
+\caption{Filtering on whether Clinton sent or received the email, or either}
+\label{fig:toFromFilter}
 \end{center}
@@
 \subsection{FOIA exemptions}
-Multiple filtering by FOIA exemption codes is supported in the app with a box selection area. 
+Multiple filtering by FOIA exemption codes is supported using a box selection area. 
 \begin{figure}[h]
 \begin{center}
-\includegraphics[width=0.35\linewidth]{FOIABoxImage}
+\includegraphics[width=0.55\linewidth]{FOIABoxImage}
+\caption{FOIA codes:  Emails containing any of these codes appear}
+\label{fig:selectFOIA}
 \end{center}
 \end{figure}
-Users can add FOIA exemption codes currently not included to the data by clicking the box and selecting a code from the menu which is then displayed below the box.
-To remove codes, the user simply clicks the code within the display box and presses delete, which moves the code to the selection menu.
-The box therefore displays all codes currently included in the filter.
+Figure~\ref{fig:selectFOIA} displays the codes that are selected for email display.
+Any email having one or more of these codes will appear in the displays.
+To not have emails display with a B1 exemption code, for example, then the user selects and deletes the B1 from the list.
+Or to have only the emails that  contain B1 redactions displayed, all other codes can deleted from the list.
+Codes are returned to active display at any time by clicking in the display box and selecting them again from a drop down menu of the excluded codes.
+Any number of codes  (one or more) can be selected to be included in the displays.
+
 \subsection{Auxiliary information}
-\TODO{e.g. Clinton's foreign schedule}
-A number of cosmetic adjustments using auxiliary information can also be made.
-Users can choose to adjust the scatterplot of time by day, which re-defines 6 pm as the time at which the day changes.
-This simply groups the emails best, placing all emails into one mostly uninterrupted region by redefining the day about the midsection of the region of lowest email traffic.
-Another cosmetic adjustment is the inclusion of Clinton's foreign trip schedule as blue regions on the email volume plot.
-This schedule was taken from the official State Department website \cite{ForeignSched} and allows users to distinguish periods spent out of the United States (coloured in blue) from those spent within the United States (uncoloured).
+To demonstrate how other sources of information can easily be used to supplement the metadata found in the email archive, we consulted the State Department website and retrieved Secretary Clinton's official foreign travel schedule  \cite{ForeignSched}.
+This information was added to the display as a simple check box ``Display Foreign Travel Schedule''.
+When checked, light blue vertical bars appear on the  email volume time series display.
+Appearing underneath the series, these mark time periods when Secretary Clinton was travelling outside the U.S.
+on State Department business.
 
-\section{Analysis}
+\section{Some interactive analyses}
 \label{sect:Analysis}
+Each of the displays and filters described in the previous sections gives a slightly different and informative view of summary features of the email archive.
+However, their analytic power is considerably amplified when used together.
+Rather complex queries can be formed involving date, classification code, and whether Clinton is sending or receiving email.
+Moreover, given a particular pattern in one display, its relation to that in another can be of considerable interest.
+Together with access to the archive contents and an internet search engine, an enormous amount can be learned from these visual analytic tools.
 
-\subsection{Context: A brief timeline on the private email server}
-\label{sect:Analysis:Timeline}
-While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to the private email server and its contents.
-It also introduces some of the principal characters in the whole affair.
-More complete and in depth timelines are readily available on the internet (e.g.
-\cite{attkissonTimeline, thompsonTimeline, WashPostTimeline, clintonWikipedia, TimeMagEverything}).
+In this section, we explore a few avenues of enquiry that might naturally arise.
+At best this only hints at the analytic power that interactive filtering and display can bring to bear on an email archive.
+To get a better appreciation, the user is encouraged to try it themselves at \texttt{rshiny.math.uwaterloo.ca/clinton}.
+The tool is useful both for generating and for testing hypotheses.
 
-On November 21, 2008, the New York Times reported that Hillary Clinton had decided to accept the position of U.S.
-Secretary of State.
-On January 13, 2009 the internet domain name \texttt{clintonemail.com} was registered \cite{whoisClintonserver}; eight days later Senator Clinton was confirmed as Secretary of State.
+\subsection{The last two years}
+One might begin an interactive analysis by looking at the email patterns in the last two years of Clinton's tenure as Secretary of State.
+This is done by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.
+Figure~\ref{fig:lastTwoYears}
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.95\linewidth]{EmailVolumeB1LastTwoYears}
+\caption{Last two years:  Selected Emails are those that are B1 redacted}
+\label{fig:lastTwoYears}
+\end{center}
+\end{figure}
+shows the resulting email volume time series over this period.
+Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.
+The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.
+As can be seen, there are relatively few of these.
 
-Public knowledge that a private email server being used by Secretary Clinton and others for State Department and personal communications did not surface until March 2015 \cite{NewYorkTimes2015} during the course of a U.S.
-Congressional investigation \cite{BenghaziReport} of the September 11, 2012 attack by militants on U.S.
-compounds in Benghazi Libya.
+Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).
+Another series of the next largest spikes appear at the far left of this time window, at the beginning of 2011.
+We might choose to compress the time window more, to focus on this section of the emails.
 
-The State Department had difficulty fulfilling public FOIA and House Benghazi Committee requests \cite{TakingRootWashPost} for Secretary Clinton's government emails because she had exclusively used the private server for all her email.
-On March 10, 2015,  Clinton told reporters that she turned over 30,490 emails to the State Department and deleted 31,830 emails deemed to be personal \cite{WashPostTimeline}.
-Clinton had tasked three lawyers Cheryl Mills (Clinton's former chief of staff),  David Kendall (Clinton's personal lawyer), and Heather Samuelson (a State Department staffer during Clinton's tenure) to make the determinations as to which emails were work related and which were not \cite{emailVetting, thompsonTimeline}.
+Bringing the right end of the time slider in to April 15, 2011 produces the series of Figure~\ref{fig:early2011B1}.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.95\linewidth]{EmailVolumeB1JanApril2011}
+\caption{Jan. 1 -- Apr 15, 2011:  Selected Emails are those that are B1 redacted}
+\label{fig:early2011B1}
+\end{center}
+\end{figure}
+As can be seen, there continue to be highs and lows of email activity.
+The highest peak now is at the far left.
+This is January 29, 2011, the day after President Mubarak of Egypt has ordered the army into the streets of Cairo to quell protests \cite{cairoTanks2011, cairoResponse2011}.
+Dipping into the contents of the Wikileaks email archive, much of the email on this day is seen to be about managing the Egyptian Crisis.
 
-On March 10, 2015 the House Benghazi Committee requested that the private email server be turned over to a neutral third party to determine which emails are personal and which are government records \cite{serverRequest}, but was informed March 27 by David Kendall that no emails remained on the private server for any kind of review \cite{serverScrubbedLawyer}.
-Between March 25 and 31, 2015, Paul Combetta (then the server's system administrator), erased all backup copies using \texttt{BleachBit} (see \texttt{www.bleachbit.org}).
+Slightly to the right of this peak, there are several more.
+At this time, international forces are gathering to consider Libya and its leader Muammar Gaddafi.
+Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on this, we move the sliders in to focus on the time range from February 25 when Secretary Clinton announced the suspension of the Libyan embassy in Washington, to March 18, the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes.
+1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}.
+On March 19, 2011 NATO military operations in Libya began \cite{LibyaTimelineWiki}.
 
-Combetta, Mills, and Samuelson will later be granted partial immunity by the Justice Department during the FBI investigations into the private email server, as were two others: Bryan Pagliano (original server manager) and  John Bentel (former director of Information Resources Management for the State Department's Executive Secretariat) \cite{immunityPolitico, immunityDailyCaller, immunityIT}.
+The readjusted time series are shown in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.95\linewidth]{EmailVolumeB1LibyaBuildup}
+\caption{Feb. 25 -- March 15, 2011:  Selected Emails are those that are B1 redacted}
+\label{fig:EmailVolumeB1LibyaBuildup}
+\end{center}
+\end{figure}
+The left most spike is the rightmost spike of approximately 80 emails from Figure~{EmailVolumeB1JanApril2011}.   
+Of course all other displays have been updating every time the time slider is adjusted.
 
-On April 12, 2015 Clinton announces that she is running for the U.S.
-Presidency.
-On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce finding classified information in the emails and that the information they found was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact.
-On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.
+Some sense of the content of the emails can be had from the updated term frequency displays shown in Figure~\ref{fig:TermFreqB1LibyaBuildup}.
+\begin{figure}[h]
+\begin{center}
+\begin{tabular}{ccc}
+\includegraphics[width=0.4\linewidth]{TFIDFB1LibyaBuildup} &
+$~~~~$ &
+\includegraphics[width=0.4\linewidth]{TermFreqB1LibyaBuildup} 
+\end{tabular}
+\caption{Frequent terms: B1 only emails, Feb. 25 -- March 18, 2011}
+\label{fig:TermFreqB1LibyaBuildup}
+\end{center}
+\end{figure}
+Not surprisingly,  Libya and its leader figure prominently in this restricted set of emails.
 
-Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton on use of private email server \cite{nochargeFBI}.
-On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin have been found on his laptop.
-Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.
-In both cases, Comey and the FBI are criticized by pundits from different political parties.
+Perhaps more surprising is a term like ``windrush''.
+A little investigation shows that this is a reference to ``Windrush Ventures'' (e.g.
+see  \cite{windrushTelegraph, windrushGuardian}),  a company owned by former U.K.
+Prime Minister Tony Blair .
+``Windrush'' appears at the end of emails as part of Mr.
+Blair's electronic signature.
+Mr.
+Blair is part of this B1 selected correspondence.
 
-\subsection{Highest peak}
-Within the plot of email volume over time, one very large peak stands out.
-Historically important,  possibly had an impact on the 2016 US presidential election.
+What other restrictions, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.95\linewidth]{FOIAB12011LibyaBuildup}
+\caption{FOIA code distribution: All  B1 redacted, Feb. 25 -- March 18, 2011}
+\label{fig:FOIAB12011LibyaBuildup}
+\end{center}
+\end{figure}
+There appear to be about 100 emails in the selected set.  In addition to having B1 redactions, many of them also have B5 (inter- or intra-agency) and B6 (personal) redactions.
+
+We might now consider the composition of the inner circle for this set of emails.  Figure~\ref{fig:InnerCircle2011B1LibyaBuildup} 
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.95\linewidth]{InnerCircle2011B1LibyaBuildup}
+\caption{Inner Circle: All  B1 redacted, Jan. 1 -- Apr 15, 2011}
+\label{fig:InnerCircle2011B1LibyaBuildup}
+\end{center}
+\end{figure}
+shows the set of correspondents.
+Not surprisingly, Jake Sullivan is the principal correspondent.
+Huma Abedin is next, though she is using two emails, one a government account the other a personal \texttt{clintonemail.com} account.
+Melanie Verveer has the least amount of traffic (judging by the lightness of her spoke colour).
+Sidney Blumenthal, a non-government employee, is in this inner circle of B1 exempted emails.
+So too is former U.K.
+Prime Minister Tony Blair; this is the unidentifiable domain account \texttt{aclb}.
+It should be noted that there could be other accounts which do not appear in the inner circle because their emails were not present in the email record.  See Section~\ref{sect:data:metadata}.
+
+We could continue to drill down into this information, checking the (unredacted) contents of the emails themselves from either the Wikileaks or State Department sources.
+We might also more closely connect email dates and correspondents with world events.
+Instead, we now turn our attention to another feature which appears in Figure~\ref{fig:lastTwoYears}.
+
+\subsection{No email sent by Clinton}
+Towards the very right of Figure~\ref{fig:lastTwoYears} there appears to be an anomaly in both time series.
+There is a noticeable flat area at zero in both series.
+If we move the leftmost time sliders to the June 1, 2012, and the rightmost to the end, February 1, 2013, we have the email volumes for the last 8 months of the archive.
+The resulting series are shown in Figure~\ref{fig:EmailVolumeB1Jun1On}.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.95\linewidth]{EmailVolumeB1Jun1On}
+\caption{June 1, 2012 -- February 1, 2013:  Selected Emails are those that are B1 redacted}
+\label{fig:EmailVolumeB1Jun1On}
+\end{center}
+\end{figure}
+There appears to be no emails for a fairly large chunk of time.
+In fact, there are few or no emails in the archive from October 30 to November 9, 2012.
+This seems curious, given that the archive is supposed to be complete in Secretary Clinton's work-related email during this time period.
+The block of no email appears as an even more unusual pattern in Figure~\ref{fig:teaser}(a).
+There the block is conspicuous by the absence amongst the emails from September 11 to November 23, 2012.
+The email on which the displays of Figure~\ref{fig:teaser} are based, is no longer filtered by FOIA code.
+It contains all email in the archive in this time period.
+Direct examination of the archives at Wikileaks and the State Department show no mail in this empty block beyond that seen in Figure~\ref{fig:teaser}(a).
+
+The inner circle for all emails from September 11 to November 23, 2012 is shown in Figure~\ref{fig:teaser}(b).
+Those closest to Clinton in terms of volume of correspondence are Clinton aides Jake Sullivan, Cheryl Mills, and Monica Hanley.
+Monica Hanley is also an account whose identity as government or not could not be determined.
+Tony Blair and Sidney Blumenthal both appear in the inner circle, as does Wendy Sherman (though the account could not be identified as government or not).
+Oscar Flores is an assistant to the Clintons, often working at their home.
+The rest are identified as corresponding through government accounts.
+
+One might ask what is happening around this time period?
+By far and away the most significant event to the U.S.
+administration was the attack on the U.S.
+diplomatic compound in Benghazi on September 11, 2012.
+The aftermath of this, particularly in a U.S.
+Presidential Election year (November, 2012), consumed much of the energy of the U.S.
+administration and Congress, not to mention the ensuing media frenzy.
+\subsubsection{Are there other zeros?}
+Using 1 month window slider and looking for flat spots at zero
+Dates:
+\begin{itemize}
+\item None until mid April 2009
+\item June 8 or 9, June 15 to about the 20th, 2009
+\item July 17-20,  around 27, 2009 
+\item Day or 2 around Oct 13, 14, 2009
+\item April 10-15 or so? 2011
+\item one or two days late August, 2011. (no mail)
+\item end of march first half of April 2012
+\item August 27-30  2012 
+\item low first week of September, 2012
+\item 2 weeks end of october, up to Nov 10 approx 2012
+\item Dec 7- 17? 2012,  Again about Dec 25 2012 to Jan 1, 2013
+\end{itemize}
 
@@
    Trey Gowdy, Chair
-
-\section{Old stuff}
-\TODO{This material is to be reused wherever}
-An interactive web application in \texttt{R} \texttt{shiny} was constructed utilizing these visualizations and the extracted data.
-Incorporated in this app was a short article explaining the use of the app, providing links to background information, and outlining some of the findings of the more interesting trends observed.
-This framework provides any individual with interest the ability to generate hypotheses by exploring the data without necessarily searching for any particular conclusion.
-
-In generating and exploring this data, a considerable amount of respect is gained for the power of the often overlooked metadata of emails, especially with public figures.
-Simply having access to the network of communications, simple indicators of identifiable features of the individuals communicating, and times sent is enough to generate interesting questions which simple Google searches of important events and days can shed a great deal of light on.
-Leveraging data beyond this metadata, including the Freedom of Information Act (FOIA) exemption codes (see https://vault.fbi.gov/explanation-of-exemptions) used to redact the emails, only bolsters this simple metadata driven investigation.
-Metadata is often viewed as less significant or somehow less intrusive data, but the exploration performed in this paper demonstrated, at the very least to the authors, what a powerful tool metadata is for generating hypotheses about data.
-Given the lack of privacy present in the modern internet age and the amount of information most individuals make publicly available, there is a clear warning here about the utility of metadata for bad or for good.
-For a public official like Hillary Clinton, the data become even more interesting due to a number of contentious issues which arose during her tenure as the Secretary of State, independent of and as a result of her use of a private server.
 \section{Concluding Remarks}

### 2017-03-31 VAST_submission/vgtc_journal_latex/paper.tex
Small typo fixes, TODOs addressed, and minor addition of content and references following the discussion of the gap in Clinton's email

@@
 
-\abstract{We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 Hillary Clinton's emails as provided by Wikileaks.
+\abstract{We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 of Hillary Clinton's emails as provided by Wikileaks.
 
@@
 
-The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} which discusses the source of the data, the extraction and cleaning process in Subsection \ref{sect:data:extrclean}, the details of the metadata in Subsection \ref{sect:data:metadata}, the use of redaction codes in Subsection \ref{sect:data:redact} and content data used in Subsection \ref{sect:data:contentinfo}.
+The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} which addresses the source of the data, the extraction and cleaning process in Subsection \ref{sect:data:extrclean}, the details of the metadata in Subsection \ref{sect:data:metadata}, the use of redaction codes in Subsection \ref{sect:data:redact} and content data used in Subsection \ref{sect:data:contentinfo}.
 The displays of the data and their patterns over the entire unfiltered data are then described in Section \ref{sect:Displays}.
@@
 On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin have been found on his laptop.
-Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.
+Since Abedin was a close aide to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.
 In both cases, Comey and the FBI are criticized by pundits from different political parties.
@@
 Wikileaks \cite{Wikileaks} has provided the same redacted pdfs and, more usefully for analysis purposes, an HTML version of each pdf.
-Consequently, have used the Wikileaks database as our data source.
+Consequently, we have used the Wikileaks database as our data source.
 All data extraction, cleaning, analysis, and presentation is done using the open source statistical programming language \texttt{R}~\cite{Rsystem}.
@@
 One such fringe case is Huma Abedin's \texttt{clintonemail.com} email account which will show in the displays as an overly long string.
-We have chosen not to special case this but leave it as is.
+We have chosen not to edit this special case but leave it as is.
 
@@
 \label{sect:Displays}
-\TODO{Each of these four displays is described and commented on in context of whole time line}
-Four displays of the metadata and quasi-metadata
 
@@
 
-The greater the number of emails between correspondents, the wider is their connecting spoke.
-Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.
+The width of the connecting spoke between Clinton and a correspondent is proportional to the number of emails sent between them, scaled relative to the maximum number of correspondences for all connections and adjusted by a constant minimum width for visibility.
+As well, correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.
+To determine these shortened spokes, the correspondents are arranged in decreasing order by number of correspondences with Clinton.
+The greatest difference between adjacent correspondence counts is determined, and all individuals which appear before the lower correspondent in this difference are given shortened spokes.
 As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
-\TODO{Chris, can you be more precise about the algorithm?}
 \subsection{Email volume}
@@
 These can be checked against world events to generate hypotheses about what might be occurring to cause such spikes.
-For example, the largest spike occurs on August 21, 2011, the beginning day of the battle for Tripoli and on which it was reported that two of Qaddafi's sons had been captured \cite{battleTripoli}.
+For example, the largest spike occurs on August 21, 2011, the beginning day of the battle for Tripoli and the day on which it was reported that two of Qaddafi's sons had been captured \cite{battleTripoli}.
 It is also the day on which the famous ``tick tock Libya'' email is composed by Jake Sullivan providing a timeline crediting Secretary Clinton with leading the U.S.
@@
 As with all other displays, the FOIA barplot will update in reaction to all filters making it a useful tool to identify how codes co-occur over any selection of emails.
-Conversely, filtering on FOIA codes can be interesting to see who sees correspondence of varying sensitivities, especially B1 and B5.
+Conversely, filtering on FOIA codes can be interesting to see who is privy to correspondence of varying sensitivities, especially B1 and B5.
 
@@
 The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display.
-\TODO{Chris, check that I have described these two measures correctly}
 The top 20 shown in Figure~\ref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.
@@
 
-Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).
+Recall from Section~\ref{sect:Displays:volume} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).
 Another series of the next largest spikes appear at the far left of this time window, at the beginning of 2011.
@@
 \end{figure}
-The left most spike is the rightmost spike of approximately 80 emails from Figure~{EmailVolumeB1JanApril2011}.   
+The left most spike is the rightmost spike of approximately 80 emails near the centre of Figure~\ref{fig:early2011B1}.   
 Of course all other displays have been updating every time the time slider is adjusted.
@@
 
-What other restrictions, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
+What other restrictions are present in this period, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
 \begin{figure}[h]
@@
 Huma Abedin is next, though she is using two emails, one a government account the other a personal \texttt{clintonemail.com} account.
-Melanie Verveer has the least amount of traffic (judging by the lightness of her spoke colour).
+Melanie Verveer has the least email traffic (judging by the lightness of her spoke colour).
 Sidney Blumenthal, a non-government employee, is in this inner circle of B1 exempted emails.
@@
 One might ask what is happening around this time period?
-By far and away the most significant event to the U.S.
+By far the most significant event to the U.S.
 administration was the attack on the U.S.
@@
 administration and Congress, not to mention the ensuing media frenzy.
+
+More specifically, the date of the U.S.
+Presidential Election in 2012 was November 6.
+The period between October 30 and November 9 also saw the discovery of controversial surveillance information in the diplomatic compound ruins \cite{BenghaziTroublingSurveillance} and a call by 53 members of the U.S.
+House of Representatives for official responses by the White House and the State Department to questions of oversight, addressed to Obama and Clinton personally.
+\cite{BenghaziLetter} It seems highly implausible that Clinton and her aides within the State Department would have nothing to email about in this period given these events.
+This makes the email gap present in the data all the more intriguing.
+
+Of course, the nature of the communications during this period and the reason for such a gap cannot be addressed by exploration within these visual analytic tools.
+Indeed, even web searches can only identify events of interest within this period, providing no definitive answer to the question begged by the existence of a period of such low communication.
+Without these visual tools, however, identifying this gap would have been difficult and unintuitive.
+In contrast, the identification of such a striking pattern arises naturally out of simple exploration using these tools.
+
 \subsubsection{Are there other zeros?}

### 2017-03-31 VAST_submission/vgtc_journal_latex/paper.tex
Updated

--- 2017-03-31 VAST_submission/vgtc_journal_latex/paper.tex
@@
 
-\subsection{No email sent by Clinton}
+\subsection{An email gap?}
+\label{sect:emailGap}
 Towards the very right of Figure~\ref{fig:lastTwoYears} there appears to be an anomaly in both time series.
@@
 One might ask what is happening around this time period?
-By far and away the most significant event to the U.S.
-administration was the attack on the U.S.
-diplomatic compound in Benghazi on September 11, 2012.
-The aftermath of this, particularly in a U.S.
-Presidential Election year (November, 2012), consumed much of the energy of the U.S.
-administration and Congress, not to mention the ensuing media frenzy.
-\subsubsection{Are there other zeros?}
-Using 1 month window slider and looking for flat spots at zero
+In a word, Benghazi.
+Five days after the assault, The U.S.
+Ambassador to the United Nations, Susan Rice, made a series of television appearances explaining that the attack on the U.S.
+Benghazi mission was a spontaneous response a  ``hateful video''  which had hours earlier caused violent protests in Cairo against the U.S.
+Embassy \cite{faceTheNationSept16}.
+Four years later, it will be revealed as part of the House Select Committee on Benghazi report \cite{BenghaziReport} that Secretary Clinton had as early as September 12, 2012, communicated to Egyptian Prime Minister Kandil that the attack was known to be planned by Al Qaeda affiliates, it was not a protest, and that it had nothing to do with the film \cite{ClintonAdmitsAttackSept12}.
+By the end of September the U.S.
+administration was calling it a terrorist attack \cite{factCheckerBenghazi}.
+A month after the attack the administration was defending itself against charges that it deliberately downplayed a terrorist attack in Libya for political reasons in a presidential election year \cite{susanRiceWashPost2012}.
+In October, focus shifted to the security at Benghazi and whether it had been reduced prior to the attack  \cite{cbsSecurity}.
+The House Committee on Oversight \& Government Reform began looking into the security aspect \cite{cbsNewsHouseOversight2012} in earnest and found evidence that security issues were known and that requests from the U.S.
+Embassy in Libya for additional security personnel had been turned down by the State Department \cite{BenghaziSecurityTestimony}.
+On October 26, CIA operatives who had defended the Benghazi mission against the terrorist attacks came forward with their story of what went on, on the ground.
+They maintain that it was an organized attack and tell reporters that they were told to ``stand down'' \cite{standDownOrder}.
+From October 26 to November 9 2012, no email appears in the archive as being sent from Secretary Clinton.
+
+Moving the time sliders to cover either side of the email gap, October 22 to November 17, 2012, results in  the term frequency displays of Figure~\ref{fig:TermFreqOct22Nov17_2012}.
+\begin{figure}[h]
+\begin{center}
+\begin{tabular}{ccc}
+\includegraphics[width=0.4\linewidth]{TFIDFOct22Nov17_2012} &
+$~~~~$ &
+\includegraphics[width=0.4\linewidth]{TermFreqOct22Nov17_2012} 
+\end{tabular}
+\caption{Around the gap: All emails, Oct. 22 to Nov.17, 2012}
+\label{fig:TermFreqOct22Nov17_2012}
+\end{center}
+\end{figure}
+It would seem that emails were very much related to the unfolding political crisis of Benghazi.  Even the name of the terrorist group, Ansar al-sharia, is picked up in the tf-idf terms.
+
+Whatever the reason for the gap seen in the archive, one thing is certain.
+The story of Benghazi and its aftermath, with its piecemeal and contentious disclosures, and unfolding as it did immediately before a U.S.
+Presidential election, would consume much of the attention and energy of the U.S.
+administration, Congress, and the U.S.
+electorate.
+It was an event that would haunt Secretary Clinton for months and years  to come \cite{benghaziTimelineWikipedia}.
+
+\subsection{A hunt for no email}
+\label{sect:emailZeros}
+The discovery of the obvious email gap in November 2012 raises the possibility that there may be other, less obvious, gaps in the email archive.
+
+To explore this possibility,  the time slider was adjusted to the relatively small span of thirty days.
+This would allow us to detect gaps in email of smaller sizes.
+To focus attention on Secretary Clinton's behaviour, only those emails sent from Secretary Clinton were displayed.
+Moving the slider, from left to right, from early 2009 to early 2013, would allow us to move a one month wide magnifying glass across the whole of the archive.
+A few contiguous days with no email from Clinton should appear visually as a small horizontal line at zero in the email volume (magenta coloured and lower) time series.
+Any such horizontal patch will be identified by its date.
+
+We found ten more patches, in addition to the original obvious email gap of Section~\ref{sect:emailGap}.
+So as to get some idea of the context of the patch, and to illustrate the advantage of coupling a visual analytic tool with an internet search engine, for every patch we found we performed a coarse internet search to see what might turn up, if anything to explain it.
+If, for example, we found a patch of no emails from Clinton in March 2009, then a Google search as shown in Figure~\ref{fig:googleSearch}
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.75\linewidth]{googleSearch}
+\caption{A coarse Google search}
+\label{fig:googleSearch}
+\end{center}
+\end{figure}
+would be tried to see what might be found in the news that could relate to that date.
+
+\subsubsection{You have no mail \ldots}
+The first, and largest, patch occurs at the very beginning of Clinton's tenure as Secretary of State.
+It is  obvious even from the full time series of Figure~\ref{fig:VolumeAll}.
+Essentially no email from Secretary Clinton, and little from anyone else appears until mid to late April 2009.
+
+In time order, the patches and what we learned from each Google search was as follows.
+
+We found ten more places in the archive which we identified as gaps to be explored.
+
+where there were zero emails from Clinton
+
 Dates:

### 2017-03-31 app.R
Changed titles to include date ranges

@@ -196,7 +196,7 @@ ui <- fluidPage(
             fluidRow(selectInput("ToFromFilter", "Show Emails: ", 
                                  c("From Clinton", "To Clinton", "All Emails"),
                                  selected = "All Emails", multiple = FALSE)),
-            fluidRow(selectInput("ClassFilter", "FOIA Codes: ",
+            fluidRow(selectInput("ClassFilter", "Emails Including Any of the FOIA Codes: ",
                                  c("B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", "B9", "None"),
                                  selected = c("B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", 
                                               "B9", "None"),
@@ -275,7 +275,7 @@ server <- function(input, output) {
      plot(x = as.chron(floor(AsSec$Date[AsSec$ID %in% inter()$selIDs])),
           y = timevalues, xlab = xlab,
           ylab = paste("Time", inter()$toFromLab), yaxt = 'n', xaxt = 'n', pch = 19,
-          main = main,
+          main = paste(main, " (", inter()$DateRange, ")", sep = ""),
           col = adjustcolor(pal[as.numeric(AsSec$Redacted[AsSec$ID %in% inter()$selIDs])+1], 
                             alpha.f = 0.5),
           cex = 0.25, ylim = c(1440,0) + c(0.05,-0.05)*1440, 
@@ -291,7 +291,7 @@ server <- function(input, output) {
    output$DaySum <- renderPlot({
      plot(x = inter()$selDays, y = inter()$selCounts, type = 'l', xlab = 'Date (dd/mm/yy)',
           xaxt = "n", ylab = 'Number of Emails', pch = 19,
-          main = "Number of Emails by Date",
+          main = paste("Number of Emails by Date (", inter()$DateRange, ")", sep = ""),
           ylim = extendrange(AScounts), col = adjustcolor("black", alpha.f = 0.6))
      # add the schedule if it has been selected
      if (inter()$dispSched) {
@@ -319,7 +319,8 @@ server <- function(input, output) {
      spiralNetPlot2(wgtTbl = sort(table(c(as.character(ClintonCom$To.name[ClintonCom$ID %in% inter()$selIDs]),
                                 as.character(ClintonCom$From.name[ClintonCom$ID %in% inter()$selIDs]))),
                         decreasing = TRUE)[-1],
-                   title = "Inner Circle by Volume of Communication")
+                   title = paste("Inner Circle by Volume of Communication (", 
[…]

### 2017-03-31 VAST_submission/vgtc_journal_latex/paper.tex
Submission

@@
 
-\abstract{We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 Hillary Clinton's emails as provided by Wikileaks.
+\abstract{We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 of Hillary Clinton's emails as provided by Wikileaks.
 
-The visualization focuses on the meta-data of each email, including its senders, receivers, and the timestamp the email appeared on the Clinton server.
+The visualization focuses on the meta-data of each email, including its senders, receivers, and the timestamp the email appeared on the Clinton server (from the Wikileaks source).
 An interactive time range slider filters all email and all displays automatically update to changes in the slider.
@@
 
-The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} which discusses the source of the data, the extraction and cleaning process in Subsection \ref{sect:data:extrclean}, the details of the metadata in Subsection \ref{sect:data:metadata}, the use of redaction codes in Subsection \ref{sect:data:redact} and content data used in Subsection \ref{sect:data:contentinfo}.
-The displays of the data and their patterns over the entire unfiltered data are then described in Section \ref{sect:Displays}.
-Within this section, Subsection \ref{sect:Displays:circle} describes the spoked network plot, Subsection \ref{sect:Displays:volume} addresses the email volume time series plot, Subsection \ref{sect:Displays:times} discusses the scatterplot of daily sending times, Subsection \ref{sect:Displays:FOIA} outlines the exemption code barplot, and Subsection \ref{sect:Displays:TFIDF} explains the content tf-idf and frequency information.
-Interactive filters which are implemented in the application are discussed in Section \ref{sect:Filters}.
-Finally, brief analysis of some interesting findings is included in Section \ref{sect:Analysis}, alongside a summary of the timeline of the Clinton email imbroglio in Section \ref{sect:Timeline}.
+The paper is organized as follows.
+We provide a brief timeline on the private email server to provide sufficient context for the work.
+Readers more familiar with the background may wish to skip to the next three sections where the implementation is described.
+Section \ref{sect:data} discusses the source of the data, its extraction and cleaning,  details of the metadata,  redaction codes as quasi-metadata,  and our limited use of the actual email content.
+Section \ref{sect:Displays} describes the displays used and illustrates them on the data.
+These include a spoked network plot, an email volume time series plot, a scatterplot of daily sending times, a barplot of exemption codes, and term frequency displays.
+Section \ref{sect:Filters} presents the interactive filters that can be applied to change the displays.
+With the implementation described, some interactive analyses of the Clinton email archive are conducted to give a quick overview of the utility of the service.
+We find some interesting, and potentially contentious, patterns in the email correspondence.
+The section closes with  caveats on over interpretation of these analyses.
+Some closing thoughts are given in the final section.
 
@@
 \label{sect:Timeline}
-While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to the private email server and its contents.
+While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to Clinton's private email server and its contents.
 It also introduces some of the principal characters involved.
@@
 
-Public knowledge that a private email server being used by Secretary Clinton and others for State Department and personal communications did not surface until March 2015 \cite{NewYorkTimes2015} during the course of a U.S.
+Public knowledge that a private email server was being used by Secretary Clinton and others for State Department and personal communications did not surface until March 2015 \cite{NewYorkTimes2015} during the course of a U.S.
 Congressional investigation \cite{BenghaziReport} of the September 11, 2012 attack by militants on U.S.
@@
 
-Combetta, Mills, and Samuelson will later be granted partial immunity by the Justice Department during the FBI investigations into the private email server, as were two others: Bryan Pagliano (original server manager) and  John Bentel (former director of Information Resources Management for the State Department's Executive Secretariat) \cite{immunityPolitico, immunityDailyCaller, immunityIT}.
+Combetta, Mills, and Samuelson were later granted partial immunity by the Justice Department during the FBI investigations into the private email server, as were two others: Bryan Pagliano (original server manager) and  John Bentel (former director of Information Resources Management for the State Department's Executive Secretariat) \cite{immunityPolitico, immunityDailyCaller, immunityIT}.
 
@@
 Presidency.
-On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce finding classified information in the emails and that the information they found was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact.
+On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce that they found classified information in the emails and that the information was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact.
 On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.
 
-Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton on use of private email server \cite{nochargeFBI}.
-On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin have been found on his laptop.
+Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton for her use of a private email server \cite{nochargeFBI}.
+On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin had been found on his laptop.
 Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.
@@
 \label{sect:data:metadata}
-For each email, from the HTML we extract as best we can the identity of the persons sending and receiving the email as well as the date and time at which the email was processed by the server.
+From the HTML of each email we extract as best we can the identity of the persons sending and receiving the email as well as the date and time at which the email was processed by the server.
 The fields used were the address to, address from, contact name to, contact name from, subject line, and time.
@@
 Moreover, because of redaction (typically FOIA exemption B6 ~\cite{FOIA}),  sender and receiver emails may have truncated domains,  contain only the person's name, or be missing altogether.
-In cases where both \texttt{From} and \texttt{To} are entirely missing, but there is an email chain within the message, we do not impute values  (e.g.
+In cases where both \texttt{From} and \texttt{To} are entirely missing we do not impute values  (e.g.
 see \texttt{https://wikileaks.org/clinton-emails/emailid/31599}).
@@
 \label{sect:Displays}
-\TODO{Each of these four displays is described and commented on in context of whole time line}
-Four displays of the metadata and quasi-metadata
-
 \subsection{Inner circle}
@@
 Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.
-Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided.
+Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided \cite{WikileaksPull}.
 
-The greater the number of emails between correspondents, the wider is their connecting spoke.
+The width of each spoke is proportional to volume of emails between correspondents, the wider is their connecting spoke.
 Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.
 As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
-\TODO{Chris, can you be more precise about the algorithm?}
 \subsection{Email volume}
@@
 One might begin an interactive analysis by looking at the email patterns in the last two years of Clinton's tenure as Secretary of State.  
-This is done by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.
 Figure~\ref{fig:lastTwoYears} 
@@
 \end{figure}
-shows the resulting email volume time series over this period.
+shows the email volume time series over this period. 
+This is accomplished by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.
 Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.
@@
 Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).
-Another series of the next largest spikes appear at the far left of this time window, at the beginning of 2011.
+The next largest spikes appear at the far left of this time window, at the beginning of 2011.
 We might choose to compress the time window more, to focus on this section of the emails.
@@
 At this time, international forces are gathering to consider Libya and its leader Muammar Gaddafi.
-Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on this, we move the sliders in to focus on the time range from February 25 when Secretary Clinton announced the suspension of the Libyan embassy in Washington, to March 18, the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes.
+Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on the Libyan crisis, we move the sliders in to focus on the time range from February 25 when Secretary Clinton announced the suspension of the Libyan embassy in Washington, to March 18, the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes.
 1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}.
@@
 \end{figure}
-The left most spike is the rightmost spike of approximately 80 emails from Figure~{EmailVolumeB1JanApril2011}.   
+The left most spike is the rightmost spike of approximately 80 emails from Figure~\ref{fig:early2011B1}.   
 Of course all other displays have been updating every time the time slider is adjusted.
@@
 
-What other restrictions, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
+Other restrictions, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
 \begin{figure}[h]
@@
 Huma Abedin is next, though she is using two emails, one a government account the other a personal \texttt{clintonemail.com} account.
-Melanie Verveer has the least amount of traffic (judging by the lightness of her spoke colour).
 Sidney Blumenthal, a non-government employee, is in this inner circle of B1 exempted emails.
@@
 Those closest to Clinton in terms of volume of correspondence are Clinton aides Jake Sullivan, Cheryl Mills, and Monica Hanley.
-Monica Hanley is also an account whose identity as government or not could not be determined.
+Monica Hanley also appears with an account whose identity as government or not could not be determined.
 Tony Blair and Sidney Blumenthal both appear in the inner circle, as does Wendy Sherman (though the account could not be identified as government or not).
@@
 In a word, Benghazi.
-Five days after the assault, The U.S.
+Five days after the assault, the U.S.
 Ambassador to the United Nations, Susan Rice, made a series of television appearances explaining that the attack on the U.S.
@@
 Embassy in Libya for additional security personnel had been turned down by the State Department \cite{BenghaziSecurityTestimony}.
-On October 26, CIA operatives who had defended the Benghazi mission against the terrorist attacks came forward with their story of what went on, on the ground.
+On October 26, CIA operatives who had defended the Benghazi mission against the terrorist attacks came forward with their story of what happened, on the ground.
 They maintain that it was an organized attack and tell reporters that they were told to ``stand down'' \cite{standDownOrder}.
@@
 
-To explore this possibility,  the time slider was adjusted to the relatively small span of thirty days.
-This would allow us to detect gaps in email of smaller sizes.
-To focus attention on Secretary Clinton's behaviour, only those emails sent from Secretary Clinton were displayed.
-Moving the slider, from left to right, from early 2009 to early 2013, would allow us to move a one month wide magnifying glass across the whole of the archive.
-A few contiguous days with no email from Clinton should appear visually as a small horizontal line at zero in the email volume (magenta coloured and lower) time series.
-Any such horizontal patch will be identified by its date.
+To explore this possibility,  the time slider can be adjusted to a relatively small span, say of thirty days.
+This would allow gaps in email of smaller sizes to be detected.
+To focus attention on Secretary Clinton's behaviour, only those emails sent from Secretary Clinton would be displayed.
+Moving the slider, from left to right, from early 2009 to early 2013, allows one to move a one month wide magnifying glass across the whole of the archive.
+A few contiguous days with no email from Clinton will appear as a small horizontal line at zero in the email volume (magenta coloured and lower) time series.
+Once found, the sliders can be further narrowed to identify the dates of the patch.
 
-We found ten more patches, in addition to the original obvious email gap of Section~\ref{sect:emailGap}.
-So as to get some idea of the context of the patch, and to illustrate the advantage of coupling a visual analytic tool with an internet search engine, for every patch we found we performed a coarse internet search to see what might turn up, if anything to explain it.
-If, for example, we found a patch of no emails from Clinton in March 2009, then a Google search as shown in Figure~\ref{fig:googleSearch}
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=0.75\linewidth]{googleSearch}
-\caption{A coarse Google search}
-\label{fig:googleSearch}
-\end{center}
-\end{figure}
-would be tried to see what might be found in the news that could relate to that date.
-
+Following this approach, ten more patches can be found (in addition to that of Section~\ref{sect:emailGap}).
+With the dates in hand, an internet search can be conducted to see what, if anything, might turn up to explain the patch.
 \subsubsection{You have no mail \ldots}
+\label{sect:youHaveNoMail}
 The first, and largest, patch occurs at the very beginning of Clinton's tenure as Secretary of State.
@@
 
+This is the time when Secretary Clinton's email habits first run afoul of the State Department as Clinton insists on using her personal Blackberry for all of her email communications.
+The issue came to a head in February when National Security officials and intelligence specialists explain the risks to Cheryl Mills.
+The officials were apparently not aware  of the private email server (e.g.
+see \cite{TakingRootWashPost}).
 In time order, the patches and what we learned from each Google search was as follows.
 
-We found ten more places in the archive which we identified as gaps to be explored.
+Three of the patches appear to be easily interpreted.
+June 17 to June 23, 2009 coincides with Clinton's surgery for an elbow fracture after a fall \cite{ClintonFracture2009}.
+A month later, from July 17 to July 23 another patch appears.
+This coincides with Secretary Clinton's State visits to India and Thailand \cite{ForeignSched}.
+Similarly, the absence of email from Clinton from October 11 to  October 13, 2009 corresponds to a State visit to Russia as shown in  Figure~\ref{fig:ClintonZerosOctober2009AndForeignSchedule} \cite{ForeignSched, visitRussia}.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.95\linewidth]{ClintonZerosOctober2009AndForeignSchedule}
+\caption{Oct 3 -- Oct 27, 2009:  Selected Emails are from Clinton; Scheduled foreign travel marked in blue.}
+\label{fig:ClintonZerosOctober2009AndForeignSchedule}
+\end{center}
+\end{figure}
 
-where there were zero emails from Clinton
+The next patch chronologically comes in 2011, from April 11 to April 15.
+On the last two days of the patch, Secretary Clinton met with other NATO Foreign Ministers in Germany \cite{ForeignSched}.
+Notably, in an April 10, 2011 email, Huma Abedin forwards Clinton the ``Stevens Update'' from Timmy Davis.
+The email describes Ambassador Stevens's concerns about the situation in Ajdabiyah (less than 150 km from Benghazi) where he is considering leaving Benghazi \cite{StevensUpdate}.
 
-Dates:
-\begin{itemize}
-\item None until mid April 2009
-\item June 8 or 9, June 15 to about the 20th, 2009
-\item July 17-20,  around 27, 2009 
-\item Day or 2 around Oct 13, 14, 2009
-\item April 10-15 or so? 2011
-\item one or two days late August, 2011. (no mail)
-\item end of march first half of April 2012
-\item August 27-30  2012 
-\item low first week of September, 2012
-\item 2 weeks end of october, up to Nov 10 approx 2012
-\item Dec 7- 17? 2012,  Again about Dec 25 2012 to Jan 1, 2013
-\end{itemize}
+From August 27 to August 29, and again in September 1, 2011 no emails are sent from Clinton.
+No foreign travel appears on the schedule.
+Notable events include ``Russian hackers'' tried to gain access to Clinton's email accounts through phishing \cite{russianHackers2}; a significant Clinton Foundation donor, Rajiv K.
+Fernando, was added by Cheryl Mills to the International Security Advisory Board  \cite{clintonDonorSecurity}; State Department executive secretary Stephen Mull is rebuffed by Huma Abedin when he suggests replacing Clinton's personal BlackBerry by a ``department-issued'' one \cite{earlyEmails}.
 
-Contentious issues:
-\begin{itemize}
-\item private email server
-\item classified documents (outside of state)
-\item Sidney Blumenthal
-\item Benghazi spin handling of media (Susan Rice)
+The next patch appears at the end of March, the beginning of April, 2012.
+The exact dates were March 27 and 29, the range from March 31 to April 3, and then again on April 10.
+Few, but not zero, emails were sent between these dates.
+Google search does not turn up much other than the fact that Clinton Staffers seem to be vetting email \cite{tightReinRecords, staffVetEmail}, and that Clinton had received an email which in part contained classified material \cite{classifiedInPart}.
 
-\item scrubbing of her server (bleachbit)
-\item missing email from online email State department
-\end{itemize}
+No emails are found sent from Secretary Clinton's email from August 27 to August 30 and September 1, 2, and 6, 2012, all just before the Benghazi attack.
+Much has already been written about this time period.
+Here we add that in a cable to ``SECSTATE'' dated August 8 (via Hanna Draper) and entitled ``The Guns of August: security in eastern Libya'' from Ambassador Stevens was the comment ``What we have seen are not random crimes of opportunity but rather targetted and discriminate attacks'' \cite{gunsOfAugust}.
+According to the Benghazi Report,  during August the number of State Department security agents assigned to the embassy in Tripoli dropped from 34 to 6 \cite{BenghaziReport}.
 
-Learn:
-\begin{itemize}
-\item inner circle over time (state or not)
-\item spike in email around Libyan revolution
-\item gap in the email
-\item daily email patterns, server behaviour (daylight savings time)
-\end{itemize}
+The last patch appears shortly after the email gap from Section~\ref{sect:emailGap}.  Figure~\ref{fig:ZeroEmailsDec2012}
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.95\linewidth]{ZeroEmailsDec2012}
+\caption{Dec 3 -- Jan 14, 2012:  Selected Emails are from Clinton; Scheduled foreign travel marked in blue.}
+\label{fig:ZeroEmailsDec2012}
+\end{center}
+\end{figure}
+shows two closely related gaps in December 2012 which, for the purposes of discourse only,  we are counting as only one of our ten patches.
+Similar to June 2009, there is an easy explanation at hand.
+Clinton had suffered a concussion ``\ldots in a fall brought on by an illness'' \cite{concussion} and was recovering at home by December 15.
+She had `already been forced to cancel a planned trip to the Middle East and north Africa'' \cite{concussion} that week; this is not the foreign travel shown in Figure~\ref{fig:ZeroEmailsDec2012} \cite{ForeignSched}.
 
- Filter:
-\begin{itemize}
-\item time
-\item redacted or not
-\item FOIA exemption codes
-\end{itemize}
+A Google search,  however, produces other news items of possible relevance.
+On  December 19, four State Department employees associated with Benghazi resigned.
+Three are top State officials: Eric Boswell,  the assistant secretary of state for diplomatic security;  Charlene Lamb, the deputy assistant responsible for embassy security; and Raymond Maxwell, the deputy assistant secretary for overseeing the Mahgreb nations of Libya, Algeria, Tunisia, and Morocco \cite{securityResigns, fourResign, tripoliPostResign}.
+The fourth is not identified in \cite{fourResign, tripoliPostResign}.
+Raymond Maxwell appears again in the news nearly two years later when, feeling scapegoated, he breaks a story to investigative reporter Sharyl Attkisson.
+Maxwell alleges that in October 2012 he witnessed a Sunday afternoon document sorting session in the basement of the State Department after Congress had called for documents related to Benghazi \cite{AttkissonBenghaziBombshell, attkissonTimeline}.
+Its purpose would appear to be the removal of potentially embarrassing documents.
+Jake Sullivan,  Cheryl Mill, and a State Department office director ``close to Clinton's top advisors''  were said to be present (see\cite{AttkissonBenghaziBombshell}).
 
- Content:
-\begin{itemize}
-\item Term frequency, TFIDF
-\end{itemize}
+Also in December 2012, Cheryl Mills and Heather Samuelson were made aware of an FOIA request from ```Citizens for Responsibility and Ethics in Washington'' (or CREW) that had been filed on December 6 \cite{CREWFOIA, FOIAcall}.
+The U.S.
+Ambassador to the U.N.
+Susan Rice withdraws her name from consideration as a Secretary of State to replace Clinton \cite{RiceWithdraws}.
+December 2012 was a busy month.
 
- Tool: 
-\begin{itemize}
-\item Web-based, interactive filter and display tool
-\end{itemize}
+\subsection{Caveats}
+\label{sect:caveats}
+As the interactive analyses above shows, there is considerable information in the metadata, especially when it can be linked to other sources of information.
+However some care must be taken when conducting such an analysis.
 
-Discovery from visualization \& connecting discovery sources
+It is quite possible that we will only see what we want to see in the data.
+The human tendency to seek out information that confirms our preconceptions has been well known since Francis Bacon first wrote in the \emph{Novum Organum},1620
+\begin{quotation}
+``The human understanding, once it has adopted opinions, either because they were already accepted and believed, or because it likes them, draws everything else to support and agree with them.
+\end{quotation}
+This is now called confirmation bias \cite{confBias} and it plagues much modern discussion, particularly on the internet and social media.
 
-- Nothing on this ... Preparation for Benghazi (security considerations)
+By building an interactive filter and display tool for Secretary Clinton's emails we have uncovered some unusual patterns.
+Many time periods show no email from Clinton.
+One large period shows no mail at all.
+Whether records are missing by accident or by intention cannot be determined by these tools.
+Instead, hypotheses have been generated which might be explored by others more knowledgeable.
 
-- House Oversight and Government Reform Committee (standing committee) 
-  Darrell Issa, Chair
-     Jason Chaffetz, Chair (Gowdy a member)
-  (discovered email server)
-    ... 
-- House Select Committee on Benghazi (Summer 2014) 
-   Trey Gowdy, Chair
+It is particularly important to try to avoid confirmation bias when performing internet search.
+For every patch where Clinton sent no email, we deliberately used a very coarse query to the Google search engine.
+Figure~\ref{fig:googleSearch}
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.75\linewidth]{googleSearch}
+\caption{The coarse Google search: for any patch found in March 2009}
+\label{fig:googleSearch}
+\end{center}
+\end{figure}
+shows the search query for a March 2009 patch.  We followed the same style of query for all patches that we uncovered visually.  Only the first page of hits returned from the search was examined.
+
+As seen in Section~\ref{sect:youHaveNoMail} this was often sufficient to turn up information that explained the lack of email as naturally occurring due to injury or travel.
+In other cases, the information turned up was possibly contentious and could be interpreted to suggest that the email was deliberately missing.
+We do not, and cannot, suggest that this is the case.
+Of course, the way in which the emails were selected to be handed over to the State Department and the subsequent erasure of all back up copies does little to dissuade anyone who might hold this view.
+
+We were surprised to see how often something contentious showed up with the Google search for the identified patches.
+To check whether such a search would nearly always turn up something contentious, we randomly selected 10 combinations of months and years over the time period of the archive (a total of 60 possible combinations).
+One of the combinations we turned up matched a patch we had identified ("Dec 2012"), the remaining did not.
+The complete sample (and what we found for each) was as follows:  ``June 2011''  (pro forma cable from SECSTATE that State Dept.
+employees change their passwords), ``Nov 2010''  (U.S.
+Embassy Cables appear on Wikileaks; Clinton rejects suggestion by Huma Abedin to start using  a government email account),   ``Dec 2012'' (documented in Section~ref{sect:youHaveNoMail}), ``March 2009'' (Clinton hits reset button with Russia), ``May 2011'' (Clinton delivers remarks on GlobalFood Security), ``March 2011'' (Clinton rejects talks with Gaddafi's sons; P.J.
+Crowley forced to resign), ``May 2012'' (State Dept.
+investigating computer shipments to Iran), ``Dec 2011'' (Putin accuses Clinton of encouraging Russian protests), ``June 2012'' (Geneva I conference on Syria; Bill Clinton checked with State Dept.
+on paid speech to group with Tehran ties; State Dept.
+Trafficking in Persons report), and  ``Oct 2011'' (Iranian plot to kill Saudi ambassador, State Dept.
+cannot confirm Gaddafi's death).
+None but ``Dec 2012'' have a connection to the Benghazi controversy though two mention email.
+
+ We need to also point out that the extraction of data is always an opportunity for error to occur.
+For example, we have already mentioned difficulties with processing email addresses that have been partially or entirely redacted.
+There are no doubt other errors as well.
+
+ The process we have followed is an automated one.  One conducted by hand, by examining each pdf, would likely produce better data but not necessarily for more than 30,000 records.
+
 \section{Concluding Remarks}
@@
 
-Finally, it is important that users realize that the value of metadata is not just in itself, where one might expect to easily discern general day-to-day habits such as one's circle of correspondents.
+Third, it is important that users realize that the value of metadata is not just in itself, where one might expect to easily discern general day-to-day habits such as one's circle of correspondents.
 Rather it is that when coupled with other sources, which are abundantly and publicly available for Senator Clinton, much more can be learned than from any one source alone.
-Those who have knowledge of, or access to, other sources may filter metadata to test previously held hypothesis; conversely, exploration of metadata could uncover patterns that generated hypotheses to be tested elsewhere.
+Those who have knowledge of, or access to, other sources may filter metadata to test previously held hypotheses; conversely, exploration of metadata could uncover patterns that generated hypotheses to be tested elsewhere.
+
+Finally, this is a cautionary tale.
+The collection and storage of metadata from any individual in our society should be of concern to all of us.
+While it is possible to discern patterns from several sources, it is also far too easy to construct a false narrative, particularly one that fits an already held point of view.
+As analysts, we fall prey to our cognitive biases \cite{gilgovichBook}.
+Interactive filter and display of metadata from a large corpus of communications add another tool to an already powerful analytic arsenal.
+As with any other powerful tool it needs to be used with caution.
 
@@
 
-\newpage
-$~~~$
+%\newpage
+%$~~~$
 \newpage

### 2017-04-01 app.R
Added timeline and a tentative/possible implementation of links to major players

@@ -151,6 +151,15 @@ Vgrepl <- function(patterns, strings) {
   sapply(strings, function(str) any(sapply(patterns, function(pat) grepl(pat, str))))
 }
 
+# define links for important individuals
+createLink <- function(val) {
+  paste("<a href='https://en.wikipedia.org/wiki/",val,"'>",
+        str_replace(val, "_", " "),"</a>", sep = "")
+}
+linknames <- c("Hillary_Clinton", "Cheryl_Mills", "Huma_Abedin", "Sidney_Blumenthal",
+               "Jake_Sullivan", "Tony_Blair")
+links <- createLink(linknames)
+
 ### App #################################################################################
 # define some javascript code for extracting user window size data
 jscode <-'
@@ -186,7 +195,10 @@ ui <- fluidPage(
      column(width = 4,
             wellPanel(id = "WriteUp", 
                       style = "overflow-y:scroll; max-height: 800px", 
-                      h3("Write up Goes Here"))),
+                      h3("A Brief Overview"),
+                      h5("Christopher D. Salahub and R. Wayne Oldford"),
+                      h3("Wikipedia Articles of Key Players"),
+                      dataTableOutput("links"))),
 
      # central interaction panel with a slider input for number of bins
      column(width = 2,
@@ -342,6 +354,10 @@ server <- function(input, output) {
    )
    # record the user window size (still cannot use in ui)
    # output$winHeight <- reactive(input$height)
+   # add server link processing
+   output$links <- renderDataTable({inter <- as.matrix(links, ncol = 1)
+   colnames(inter) <- "Individual"
+   return(inter)}, escape = FALSE)
 }
 
 # Run the application

### 2017-04-01 app.R
Added timeline text

@@ -197,6 +197,86 @@ ui <- fluidPage(
                       style = "overflow-y:scroll; max-height: 800px", 
                       h3("A Brief Overview"),
                       h5("Christopher D. Salahub and R. Wayne Oldford"),
+                      h3("The Email Timeline"),
+                      p("While the following timeline is somewhat abbreviated, it should serve
+                        to raise some of the major issues and concerns related to Clinton’s
+                        private email server and its contents. It also introduces some of the
+                        principal characters involved. More complete and in depth timelines
+                        are readily available elsewhere (e.g. ", 
+                        a("Sharyl Attkisson",  
+                          href = "https://sharylattkisson.com/hillary-clintons-email-the-definitive-timeline/"),
+                        " or ", 
+                        a("The Washington Post",
+                          href = "https://www.washingtonpost.com/news/fact-checker/wp/2015/03/10/hillary-clintons-emails-a-timeline-of-actions-and-regulations/?utm_term=.dec3139a0542"),
+                        ") On November 21, 2008, the New York Times reported that Hillary
+                        Clinton had decided to accept the position of U.S. Secretary of State.
+                        On January 13, 2009 the internet domain name clintonemail.com
+                        was ",
+                        a("registered", href = "https://www.whois.net/"), 
+                        " and eight days later Senator Clinton was confirmed as
+                        Secretary of State."),
+                      p("Public knowledge that a private email server was being used by
+                        Secretary Clinton and others for State Department and personal communications
+                        did not surface until ",
+                        a("March 2015", href = "https://www.nytimes.com/2015/03/03/us/politics/hillary-clintons-use-of-private-email-at-state-department-raises-flags.html"),
+                        " during the course of a ", 
+                        a("U.S. Congressional investigation", href = "http://benghazi.house.gov/"), 
+                        " of the September 11, 2012 attack by militants on U.S. compounds in 
+                        Benghazi Libya. The State Department had ",
+                        a("difficulty fulfilling", href = "https://www.washingtonpost.com/investigations/how-clintons-email-scandal-took-root/2016/03/27/ee301168-e162-11e5-846c-10191d1fc4ec_story.html?utm_term=.95919beefdc0"),
+                        " public FOIA and House Benghazi Committee requests for Secretary Clinton’s
+                        government emails because she had exclusively used the private server
+                        for all her email. On March 10, 2015, Clinton told reporters that she
+                        turned over 30,490 emails to the State Department and deleted 31,830
[…]

### 2017-04-01 app.R
Expanding the write up

@@ -278,7 +278,23 @@ ui <- fluidPage(
                         "In both cases, Comey and the FBI are criticized by pundits
                        from different political parties."),
                       h3("Wikipedia Articles of Key Players"),
-                      dataTableOutput("links"))),
+                      dataTableOutput("links"),
+                      h3("The Service"),
+                      p("This application provides the ability to interactively filter emails sent
+                       during Hillary Clinton's tenure as the United States Secretary of State and
+                       display features of the selected subset.The data is extracted from
+                       HTML representations of the ",
+                        a("official State Department release", href = "https://foia.state.gov/Search/Results.aspx?collection=Clinton_Email"),
+                        "provided in a ",
+                        a("Wikileaks data base.", href = "https://wikileaks.org/clinton-emails/")),
+                      p("This service is not meant to provide stand-alone means of analyzing this
+                       controversial data set. It is most powerful when used simultaneously with
+                       both internet searches and the Wikileaks data base or official State 
+                       Department site. The latter two services provide indispensable context
+                       and precision respectively; two services which the primarily 
+                       metadata-driven displays cannot provide. Rather, the intended use of this
+                       application is the exploration of patterns present in the data to
+                       generate and explore different hypotheses."))),
 
      # central interaction panel with a slider input for number of bins
      column(width = 2,

### 2017-04-02 VAST_submission/vgtc_journal_latex/paper.tex
Introduced subsections, changed ten patches to nine

@@
 
-Following this approach, ten more patches can be found (in addition to that of Section~\ref{sect:emailGap}).
+Following this approach, nine %ten 
+more patches can be found (in addition to that of Section~\ref{sect:emailGap}).
 With the dates in hand, an internet search can be conducted to see what, if anything, might turn up to explain the patch.
-\subsubsection{You have no mail \ldots}
+%\subsubsection{You have no mail \ldots}
+\subsubsection{Early days}
 \label{sect:youHaveNoMail}
@@
 see \cite{TakingRootWashPost}).
-In time order, the patches and what we learned from each Google search was as follows.
+%In time order, the patches and what we learned from each Google search was as follows.
 
+\subsubsection{Three patches in 2009}
 Three of the patches appear to be easily interpreted.
@@
 
-The next patch chronologically comes in 2011, from April 11 to April 15.
-On the last two days of the patch, Secretary Clinton met with other NATO Foreign Ministers in Germany \cite{ForeignSched}.
+\subsubsection{April 11 to April 15, 2011}
+%The next patch chronologically comes in 2011, from April 11 to April 15.   
+On the last two days of %the 
+this patch, Secretary Clinton met with other NATO Foreign Ministers in Germany \cite{ForeignSched}.
 Notably, in an April 10, 2011 email, Huma Abedin forwards Clinton the ``Stevens Update'' from Timmy Davis.
@@
 
-From August 27 to August 29, and again in September 1, 2011 no emails are sent from Clinton.
+\subsubsection{August 27 to August 29, September 1, 2011, 2011}
+%From August 27 to August 29, and again in September 1, 2011 no emails are sent from Clinton.   
 No foreign travel appears on the schedule.
@@
 
-The next patch appears at the end of March, the beginning of April, 2012.
+\subsubsection{End of March,  beginning of April, 2012}
+%The next patch appears at the end of March, the beginning of April, 2012.  
 The exact dates were March 27 and 29, the range from March 31 to April 3, and then again on April 10.
@@
 
+\subsubsection{End of August,  beginning of September, 2012}
 No emails are found sent from Secretary Clinton's email from August 27 to August 30 and September 1, 2, and 6, 2012, all just before the Benghazi attack.
@@
 
-The last patch appears shortly after the email gap from Section~\ref{sect:emailGap}.  Figure~\ref{fig:ZeroEmailsDec2012}
+\subsubsection{December 2012}
+The last patch appears 
+shortly after the email gap from Section~\ref{sect:emailGap}.  Figure~\ref{fig:ZeroEmailsDec2012}
 \begin{figure}[h]
@@
 \end{figure}
-shows two closely related gaps in December 2012 which, for the purposes of discourse only,  we are counting as only one of our ten patches.
+shows two closely related gaps in December 2012 which, for the purposes of discourse only,  we are counting as only one of our 
+%ten 
+nine
+patches.
 Similar to June 2009, there is an easy explanation at hand.

### 2017-04-02 VAST_submission/vgtc_journal_latex/paper.tex
changed back to ten patches; Dec 2012 treated as two (as originally)

@@
 
-Following this approach, nine %ten 
+Following this approach, ten 
 more patches can be found (in addition to that of Section~\ref{sect:emailGap}).
@@
 
-\subsubsection{December 2012}
-The last patch appears 
-shortly after the email gap from Section~\ref{sect:emailGap}.  Figure~\ref{fig:ZeroEmailsDec2012}
+\subsubsection{Two patches in December 2012}
+The last two patches appear 
+shortly after the email gap from Section~\ref{sect:emailGap}.  As seen in Figure~\ref{fig:ZeroEmailsDec2012}
 \begin{figure}[h]
@@
 \end{figure}
-shows two closely related gaps in December 2012 which, for the purposes of discourse only,  we are counting as only one of our 
-%ten 
-nine
-patches.
-Similar to June 2009, there is an easy explanation at hand.
+these are the two closely related gaps in December 2012 which,
+% for the purposes of discourse only,  we are counting as only one of our 
+%ten patches
+might be counted as one large gap.
+Similar to June 2009, there is an easy explanation at hand, at least for the first of these.
 Clinton had suffered a concussion ``\ldots in a fall brought on by an illness'' \cite{concussion} and was recovering at home by December 15.

### 2017-04-02 app.R
Formatting and write up updates

@@ -178,7 +178,8 @@ ui <- fluidPage(
    tags$head(tags$script(jscode)),
 
    # Application title
-   titlePanel("Secretary Clinton's Email (Source: https://wikileaks.org/clinton-emails/)"),
+   titlePanel(HTML("Secretary Clinton's Email (Source: <a href = 'https://wikileaks.org/clinton-emails/'>Wikileaks</a>)")),
+   fluidRow(column(h4("Christopher D. Salahub and R. Wayne Oldford"), offset = 0.2, width = 12)),
 
    # the slider right below the title to make it as long as possible
    fluidRow(column(width = 2, offset = 0.5, h4("Date Range:"))),
@@ -194,107 +195,62 @@ ui <- fluidPage(
      # well panel with write up
      column(width = 4,
             wellPanel(id = "WriteUp", 
-                      style = "overflow-y:scroll; max-height: 800px", 
-                      h3("A Brief Overview"),
-                      h5("Christopher D. Salahub and R. Wayne Oldford"),
-                      h3("The Email Timeline"),
-                      p("While the following timeline is somewhat abbreviated, it should serve
-                        to raise some of the major issues and concerns related to Clinton’s
-                        private email server and its contents. It also introduces some of the
-                        principal characters involved. More complete and in depth timelines
-                        are readily available elsewhere (e.g. ", 
-                        a("Sharyl Attkisson",  
-                          href = "https://sharylattkisson.com/hillary-clintons-email-the-definitive-timeline/"),
-                        " or ", 
-                        a("The Washington Post",
-                          href = "https://www.washingtonpost.com/news/fact-checker/wp/2015/03/10/hillary-clintons-emails-a-timeline-of-actions-and-regulations/?utm_term=.dec3139a0542"),
-                        ") On November 21, 2008, the New York Times reported that Hillary
-                        Clinton had decided to accept the position of U.S. Secretary of State.
-                        On January 13, 2009 the internet domain name clintonemail.com
-                        was ",
-                        a("registered", href = "https://www.whois.net/"), 
-                        " and eight days later Senator Clinton was confirmed as
-                        Secretary of State."),
-                      p("Public knowledge that a private email server was being used by
-                        Secretary Clinton and others for State Department and personal communications
-                        did not surface until ",
-                        a("March 2015", href = "https://www.nytimes.com/2015/03/03/us/politics/hillary-clintons-use-of-private-email-at-state-department-raises-flags.html"),
[…]

### 2017-04-03 VAST_submission/vgtc_journal_latex/paper.tex
Added Bacon reference

@@
 It is quite possible that we will only see what we want to see in the data.
-The human tendency to seek out information that confirms our preconceptions has been well known since Francis Bacon first wrote in the \emph{Novum Organum},1620
+The human tendency to seek out information that confirms our preconceptions has been well known since 1620 when Francis Bacon first wrote
 \begin{quotation}
-``The human understanding, once it has adopted opinions, either because they were already accepted and believed, or because it likes them, draws everything else to support and agree with them.
+``The human understanding, once it has adopted opinions, either because they were already accepted and believed, or because it likes them, draws everything else to support and agree with them.''
+
+\noindent
+Francis Bacon, \emph{Novum Organum}, Book 1, Aphorism 46 \cite{Bacon1620}
+
 \end{quotation}

### 2017-04-03 VAST_submission/vgtc_journal_latex/paper.tex
Added Wikileaks transcription error on b9

@@
 For example, we have already mentioned difficulties with processing email addresses that have been partially or entirely redacted.
+The Wikileaks transcription from pdf to HTML is also imperfect.
+For example, the string ``b9''  appears in one email record (Doc ID: 10257) in the Wikileaks transcription\cite{errorWikileaks} as a transcription error of the day ``09'' from the original pdf.
+This might be grabbed as a B9  redaction code, if one is not careful.
+The same email record also has ``To'' and ``From'' header which is empty except for a date.
+This header does not appear in the original pdf; rather the pdf begins with the next one inside the HTML.
 There are no doubt other errors as well.
 
- The process we have followed is an automated one.  One conducted by hand, by examining each pdf, would likely produce better data but not necessarily for more than 30,000 records.
+ The process we have followed is an automated one.
+One conducted by hand, by examining each pdf, would likely produce better data but not necessarily for more than 30,000 records.
+At least a check by hand of all HTML email records with blank to and froms at the beginning could be of some value.
 

### 2017-04-03 app.R
Cleaned up the errors when no emails are selected

@@ -70,10 +70,10 @@ for (name in allnames) {
 }
 
 # modify the spiral network plot function, improve it
-spiralNetPlot2 <- function(centralNode = "Hillary Clinton", wgtTbl = NA,
+spiralNetPlot2 <- function(centralNode = "Hillary Clinton", wgtTbl = integer(0),
                            levelNum = 2, nodeNum = 20, title = "Title") {
   # check the wgtTbl status
-  if (!all(is.na(wgtTbl))) {
+  if (!(length(wgtTbl) == 0)) {
     # trim wgtTbl
     if (length(wgtTbl) > nodeNum) wgtTbl <- wgtTbl[1:nodeNum]
     # start by generating state mail colouring
@@ -137,6 +137,18 @@ spiralNetPlot2 <- function(centralNode = "Hillary Clinton", wgtTbl = NA,
     grid.circle(x = 0.5, y = 0.5, r = 0.01,
                 gp = gpar(col = adjustcolor(col = "firebrick", alpha.f = 1),
                           fill = adjustcolor(col = "firebrick", alpha.f = 1)))
+    # add a title
+    grid.text(title, x = 0, y = 0.98, just = "left",
+              gp = gpar(face = 2))
+    # add a legend
+    grid.text(".gov", x = 0, y = 0.02, just = "left", 
+              gp = gpar(face = 2, col = "steelblue"))
+    grid.text(".mil", x = 0.33, y = 0.02, just = "center",
+              gp = gpar(face = 2, col = "black"))
+    grid.text("Not .gov", x = 0.66, y = 0.02, just = "center", 
+              gp = gpar(face = 2, col = "firebrick"))
+    grid.text("Unidentifiable", x = 1, y = 0.02, just = "right", 
+              gp = gpar(face = 2, col = "darkorange"))
   }
 }
 
@@ -220,7 +232,7 @@ ui <- fluidPage(
                                a("Clinton's tenure as Secretary of State",
                                  href = "https://en.wikipedia.org/wiki/Hillary_Clinton%27s_tenure_as_Secretary_of_State"),
                                " | ",
-                               a("2011 Benghazi attack",
+                               a("2012 Benghazi attack",
                                  href = "https://en.wikipedia.org/wiki/Timeline_of_the_investigation_into_the_2012_Benghazi_attack#October_2012")),
                       fluidRow(id = "App", h3("The Service"),
                                p(HTML("<a href='#Contents'>Back to table of contents</a>")),
[…]

### 2017-04-04 app.R
Improved the write up

@@ -262,7 +262,51 @@ ui <- fluidPage(
                                  "over the filtered region selected.")),
                       fluidRow(id = "Analysis", 
                                h3("Some Interesting Analyses"),
-                               p(HTML("<a href='#Contents'>Back to table of contents</a>"))))),
+                               p(HTML("<a href='#Contents'>Back to table of contents</a>")),
+                               h4("Peak Email"),
+                               p("Focusing on the email volume plot, an obvious peak can be seen
+                                 near the centre of the time series. Using the date selection
+                                 slider, the day of highest email volume can be identified as
+                                 August 21, 2011, the beginning of the ",
+                                 a("Battle of Tripoli", href = "https://en.wikipedia.org/wiki/Battle_of_Tripoli_(2011)"),
+                                 "in the Libyan Civil War. Inspecting the term frequency and tf-idf
+                                 for this day reveals a host of terms related to this conflict. The ",
+                                 a("United States", href = "https://en.wikipedia.org/wiki/American_involvement_in_the_2011_Libyan_Civil_War"),
+                                 "and ",
+                                 a("NATO", href = "http://www.nato.int/cps/en/natohq/topics_71652.htm"),
+                                 "were both heavily involved in this conflict, so this peak makes
+                                 perfect sense."),
+                               h4("Email Gaps"),
+                               p("There are a number of conspicuous time periods where no emails
+                                 are recorded in this data set. The most obvious of these occurs
+                                 in early November 2012. This time period marks the beginning of 
+                                 much of the ",
+                                 a("increased controversy", href = "https://en.wikipedia.org/wiki/Timeline_of_the_investigation_into_the_2012_Benghazi_attack"),
+                                 "surrounding the ",
+                                 a("2012 attack", href = "https://en.wikipedia.org/wiki/2012_Benghazi_attack"),
+                                 "on the US Diplomatic compound in Benghazi, and also includes the 
+                                 2012 US Presidential Election."),
+                               p("Using the time slider to select a period surrounding this gap
+                                 from October 2 to November 20, mentions of terms related to 
+                                 this attack, such as Benghazi and Ansar al-Sharia, and to the
+                                 election, such as Romney, can be seen. The network plot also 
+                                 reveals one of the contentious points of interest in Clinton's 
+                                 emails, the nature and frequency of her contact with ",
+                                 a("Sidney Blumenthal", href = "https://en.wikipedia.org/wiki/Sidney_Blumenthal#Relationship_to_Hillary_Clinton_and_post.E2.80.932007_employment"),
[…]

### 2017-04-04 app.R
More write up additions

@@ -169,7 +169,7 @@ createLink <- function(val) {
         str_replace(val, "_", " "),"</a>", sep = "")
 }
 linknames <- c("Hillary_Clinton", "Cheryl_Mills", "Huma_Abedin", "Sidney_Blumenthal",
-               "Jake_Sullivan", "Tony_Blair")
+               "Jake_Sullivan", "Tony_Blair", "Philippe_Reines")
 links <- createLink(linknames)
 
 ### App #################################################################################
@@ -263,7 +263,13 @@ ui <- fluidPage(
                       fluidRow(id = "Analysis", 
                                h3("Some Interesting Analyses"),
                                p(HTML("<a href='#Contents'>Back to table of contents</a>")),
+                               p(HTML("<a href='#Peak'>Peak Email</a>")),
+                               p(HTML("<a href='#Gaps'>Email Gaps</a>")),
+                               p(HTML("<a href='#EmTimes'>Email Times</a>"))),
+                      fluidRow(id = "Peak",
                                h4("Peak Email"),
+                               p(HTML("<a href='#Contents'>Back to table of contents</a>")),
+                               p(HTML("<a href='#Analysis'>Back to analysis links</a>")),
                                p("Focusing on the email volume plot, an obvious peak can be seen
                                  near the centre of the time series. Using the date selection
                                  slider, the day of highest email volume can be identified as
@@ -275,8 +281,14 @@ ui <- fluidPage(
                                  "and ",
                                  a("NATO", href = "http://www.nato.int/cps/en/natohq/topics_71652.htm"),
                                  "were both heavily involved in this conflict, so this peak makes
-                                 perfect sense."),
+                                 perfect sense. In fact, many other local maxima correspond to 
+                                 events related to the ",
+                                 a("Arab Spring", href = "https://en.wikipedia.org/wiki/Arab_Spring"),
+                                 "and the countries affected by this revolutionary wave.")),
+                      fluidRow(id = "Gaps",
                                h4("Email Gaps"),
+                               p(HTML("<a href='#Contents'>Back to table of contents</a>")),
+                               p(HTML("<a href='#Analysis'>Back to analysis links</a>")),
                                p("There are a number of conspicuous time periods where no emails
                                  are recorded in this data set. The most obvious of these occurs
                                  in early November 2012. This time period marks the beginning of 
@@ -286,18 +298,47 @@ ui <- fluidPage(
                                  a("2012 attack", href = "https://en.wikipedia.org/wiki/2012_Benghazi_attack"),
[…]

### 2017-04-04 app.R
Some edits to the write-up

@@ -250,7 +250,16 @@ ui <- fluidPage(
                                  and precision; two services which the primarily 
                                  metadata-driven displays cannot provide. Rather, the intended use of this
                                  application is the exploration of patterns present in the data to
-                                 generate and explore different hypotheses.")),
+                                 generate and explore different hypotheses."),
+                               p("It is hoped that this service will not simply provide you
+                                 with a means of exploring this particular data set, but will
+                                 demonstrate how much can be discovered about an individual
+                                 using visual analytic tools of uninformative metadata to
+                                 motivate searches of public data. The dissemination of data in
+                                 the modern world is a topic of heated discussion, and hopefully
+                                 experience firsthand exploring the way data can be leveraged
+                                 will prove informative to you and help you to inform your own
+                                 opinion on the subject.")),
                       fluidRow(id = "Data", h3("The Data"),
                                p(HTML("<a href='#Contents'>Back to table of contents</a>")),
                                p("This application allows the user to view the senders, recipients, times, ",
@@ -329,11 +338,11 @@ ui <- fluidPage(
                                  "of four State Department officials due to the results of the
                                  Benghazi investigation."),
                                p("A number of other gaps of possible interest are not discussed
-                                 here, and users are encouraged to investigate any period of
-                                 interest they notice for themselves. However, users should
+                                 here, and you are encouraged to investigate any period of
+                                 interest you notice for yourself. However, you should
                                  always be mindful of the tendency for all of us to seek
                                  information which confirms preconceptions, and attempt as
-                                 much as possible to be honest and unbiased in these 
+                                 much as possible to be honest and unbiased in your 
                                  investigations.")),
                       fluidRow(id = "EmTimes",
                                h4("Email Times"),

### 2017-04-05 app.R
Updated the Albright and Blair links

@@ -169,7 +169,7 @@ createLink <- function(val) {
         str_replace(val, "_", " "),"</a>", sep = "")
 }
 linknames <- c("Hillary_Clinton", "Cheryl_Mills", "Huma_Abedin", "Sidney_Blumenthal",
-               "Jake_Sullivan", "Tony_Blair", "Philippe_Reines")
+               "Jake_Sullivan", "Philippe_Reines")
 links <- createLink(linknames)
 
 ### App #################################################################################
@@ -218,7 +218,10 @@ ui <- fluidPage(
                       fluidRow(id = "KeyPlayers", 
                                h3("Wikipedia Articles of Key Players"),
                                p(HTML("<a href='#Contents'>Back to table of contents</a>")),
-                               p(HTML(paste(links, collapse = " | ")))),
+                               p(HTML(paste(paste(links, collapse = " | "),
+                                            "<a href='https://en.wikipedia.org/wiki/Madeleine_Albright'>Madeline Albright ('pathfinder')</a>",
+                                            "<a href='https://en.wikipedia.org/wiki/Tony_Blair'>Tony Blair ('aclb')</a>",
+                                            sep = " | ")))),
                       fluidRow(id = "Timeline", h3("The Email Timeline"),
                                p(HTML("<a href='#Contents'>Back to table of contents</a>")),
                                a("Sharyl Attkisson Timeline",  

### 2017-04-05 app.R
Changed a typo

@@ -219,7 +219,7 @@ ui <- fluidPage(
                                h3("Wikipedia Articles of Key Players"),
                                p(HTML("<a href='#Contents'>Back to table of contents</a>")),
                                p(HTML(paste(paste(links, collapse = " | "),
-                                            "<a href='https://en.wikipedia.org/wiki/Madeleine_Albright'>Madeline Albright ('pathfinder')</a>",
+                                            "<a href='https://en.wikipedia.org/wiki/Madeleine_Albright'>Madeleine Albright ('pathfinder')</a>",
                                             "<a href='https://en.wikipedia.org/wiki/Tony_Blair'>Tony Blair ('aclb')</a>",
                                             sep = " | ")))),
                       fluidRow(id = "Timeline", h3("The Email Timeline"),

### 2017-04-10 VAST_submission/vgtc_journal_latex/paper.tex
Cleaning up text

@@
 also provided a searchable web archive of the documents, released in several instalments from May 2015 to March 2017\cite{StateDeptFOIA}.
-Both sites provide a  useful tool for anyone searching for particular terms in the documents.
+Both sites provide a  useful tool for anyone searching for particular terms in these documents.
 The Wikileaks site was put to much use by investigative reporters, and others, to search for topical news items.
@@
 What is critically missing from either site is a facility to learn summary, or statistical, features across all, or groups of,  emails.
-A separate site where visual displays of summary features of the emails, which can be filtered by user selection, will go a long way towards filling this need.
+A separate site where visual displays of summary features of the emails, which can be filtered by user selection, would go a long way towards filling this need.
 The site would be \emph{complementary} to the more typical ``content search and record display'' site as provided by either online archive.
@@
 All displays are reactive and update simultaneously in response to user interaction with the data filters.
-Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in the emails.
+Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in, the emails.
 
@@
 With the implementation described, some interactive analyses of the Clinton email archive are conducted to give a quick overview of the utility of the service.
-We find some interesting, and potentially contentious, patterns in the email correspondence.
+We find some interesting, and potentially contentious, patterns in the email correspondence and check these against information found from Google searches for the dates in which the patterns occur.
 The section closes with caveats on over interpretation of these analyses.
@@
 
-On April 12, 2015 Clinton announces that she is running for the U.S.
+On April 12, 2015 Clinton announced that she would be running for the U.S.
 Presidency.
-On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce that they found classified information in the emails and that the information was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact.
-On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.
+On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announced that they had found classified information in the emails and that the information would have been considered classified at the time it was sent \cite{serverClassified}.
+The Clinton presidential campaign declared that the emails must have been classified after the fact.
+On August 19, 2015, Clinton described the allegation of her mishandling classified information as simply a ``disagreement between agencies'' \cite{clintonDenialGuardian}.
 
 Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton for her use of a private email server \cite{nochargeFBI}.
-On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin had been found on his laptop.
+On October 28, 2016, Comey revealed that, in a separate investigation into former Congressman Anthony Weiner, emails belonging to his wife Huma Abedin had been found on his laptop.
 Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.
-In both cases, Comey and the FBI are criticized by pundits from different political parties.
+In the first case Comey and the FBI were criticized by republican pundits and in the second case by democratic pundits.
 
@@
 As of March 3, 2017, a total of 32,795 available emails, either to or from Hillary Clinton, have been made publicly available in pdf form as a searchable database \cite{StateDeptFOIA}.
-Many of these have been redacted according to the FOIA exemption codes \cite{FOIA}.
+Many of these have been redacted according to the Freedom of Information Act (FOIA) exemption codes \cite{FOIA}.
 
@@
 The other red account is that of Sidney Blumenthal.
-This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to provide Clinton advice throughout her tenure.
+This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to have provided Secretary Clinton advice throughout her tenure.
 Wendy Sherman appears as orange.
@@
 Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.
-Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided \cite{WikileaksPull}.
+Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided by the WikiLeaks HTML \cite{WikileaksPull} (e.g.
+occasionally, a blank header appeared in the WikiLeaks HTML that was not present in the pdf; we did not correct for these manually).
 
-The width of each spoke is proportional to volume of emails between correspondents, the wider is their connecting spoke.
+The width of each spoke is proportional to the volume of emails between correspondents.
 Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.
@@
 presents two time series showing the volume of emails for each day.
-In grey at top is the daily total; in magenta at bottom is the total for the subset chosen by the filters.
-In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails sent from Clinton to others.
+In grey at top is the daily total; in magenta at bottom is the total for the subset selected by the filters.
+In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails that were sent \emph{from} Clinton.
 
@@
 shows all emails sent by Clinton over the four year period.
-Each email appears as a point determined by the time shown that it was sent.
+Each email appears as a point determined by the time that it was sent.
 The calendar date determines the position on the horizontal axis,  the (24hr) time of day determines the vertical.
@@
 If these are the actual times Secretary Clinton composed and sent email, it suggests that much of that activity occurred in the middle of the night.
-To reflect the low email activity period as actual ``downtime'' between ``days'' (now interpreted as separable 24 hour email cycles),  we offer the possibility of choosing 6 PM as the ``end'' of a day rather than midnight.
-This places the least active email period as a ``quiet'' boundary between days.
-Unfortunately, though it better represents the rhythm of correspondence, it can be confusing since it means that the ``next day''  begins just after 6 PM.
+We might alternatively choose the low email activity period to mark the actual ``downtime'' between ``days'' (now interpreted as low-activity separated 24 hour email cycles);   to this end, it is possible to select 6 PM as the ``end'' of a 24 hour email cycle rather than midnight.
+Though it better represents the rhythm of correspondence, this definition of the daily email cycle is unnatural to most people and so could be confusing to some users (hence by default the break occurs at midnight).
 
@@
 This regularity is suggestive of some automated schedule for the server which causes the email to stack up before being recorded at either 2 or 3 AM.
-The location of these lines switches exactly whenever daylight savings time switches in North America.
+The switch between these two times matches exactly the dates where daylight savings time switches in North America.
 
@@
 U.S.
-Freedom of Information Act (FOIA) exemption codes B1 through B9 are used in each place in an email where information has been redacted.
-For the authoritative definition the act itself should be consulted \cite{FOIA}.
-Briefly, they mark each redaction as B1 for national security and foreign policy matters,  B2 for personnel practices, B3 for statutory exemptions,  B4 for trade secrets or financial information obtained in confidence,  B5 for inter- or intra-agency memorandums,  B6 for personal privacy,  B7 for records compiled for law enforcement,  B8 for records prepared in relation to financial monitoring institutions, and B9 for geophysical and geological information concerning oil and gas wells.
+Freedom of Information Act (FOIA) exemption codes B1 through B9 are used anywhere in an email where information has been redacted.
+For the authoritative definition of the codes the act itself should be consulted \cite{FOIA}.
+Briefly, a redaction is marked as B1 for national security and foreign policy matters,  B2 for personnel practices, B3 for statutory exemptions,  B4 for trade secrets or financial information obtained in confidence,  B5 for inter- or intra-agency memoranda,  B6 for personal privacy,  B7 for records compiled for law enforcement,  B8 for records prepared in relation to financial monitoring institutions, and B9 for geophysical and geological information concerning oil and gas wells.
 Each email can contain redactions for any number of exemption codes.
@@
 is a barplot of all FOIA exemption codes found in all of the emails.
-As can be seen, B1 redactions made for national security and foreign policy reasons account for a small portion of all redactions made though affect more than 2,000 emails in the collection.
+As can be seen, B1 redactions made for national security and foreign policy reasons account for a small portion of all redactions made,  though this amounts to more than 2,000 emails in the collection.
 Nearly half of the emails contain B6 exemptions, those made to protect personal privacy.
@@
 
-As with all other displays, the FOIA barplot will update in reaction to all filters making it a useful tool to identify how codes co-occur over any selection of emails.
-Conversely, filtering on FOIA codes can be interesting to see who sees correspondence of varying sensitivities, especially B1 and B5.
+As with all other displays, the FOIA barplot will update in reaction to all filters and so provides a useful tool to identify co-occurrence of codes over any selection of emails.
+Conversely, filtering on FOIA codes can be interesting to see who is privy to correspondence of varying sensitivities, especially B1 and B5.
 
@@
 As seen in Figure~\ref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.
-One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies which could magnify the within email frequency for some terms.
+One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies; this could magnify the within email frequency for some terms.
 
-As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is also challenged by this messy and non-standard data.
+As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is challenged by this messy and non-standard data.
 The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'.
@@
 \end{figure}
-is used to select those emails which Clinton sent, or received, or either sent or received.
+is used to select those emails which Clinton sent, or received, or either (i.e.
+sent or received).
 The default is ``All Emails'', which is the same as no filter being applied for sent or received.
@@
 This might be used, for example, to explore whether the inner circle of correspondents changes depending on whether Secretary Clinton is emailing them, or they are emailing her.
+
+Note that if neither the ``From:''  nor ``To:''  were captured from the HTML, then such emails can appear only when ``All Emails'' is selected.
 
@@
 Any email having one or more of these codes will appear in the displays.
-To not have emails display with a B1 exemption code, for example, then the user selects and deletes the B1 from the list.
-Or to have only the emails that  contain B1 redactions displayed, all other codes can deleted from the list.
+To not have emails display with a B1 exemption code, for example, then the user selects and deletes the B1 from the list (multiple selection is enabled).
+Or to have only the emails which  contain B1 redactions displayed, all other codes can deleted from the list.
+
 Codes are returned to active display at any time by clicking in the display box and selecting them again from a drop down menu of the excluded codes.
@@
 When checked, light blue vertical bars appear on the  email volume time series display.
-Appearing underneath the series, these mark time periods when Secretary Clinton was travelling outside the U.S.
-on State Department business.
+Appearing behind the series, these mark time periods when Secretary Clinton was travelling outside the U.S.
+on State Department business according to the published schedule.
 
@@
 \label{sect:Analysis}
-Each of the displays and filters described in the previous sections gives a slightly different and informative view of summary features of the email archive.
-However, their analytic power is considerably amplified when used together.
+Each of the displays and filters described in the previous sections gives a slightly different informative view of summary features of the email archive.
+Their analytic power is considerably amplified when used together.
+
 Rather complex queries can be formed involving date, classification code, and whether Clinton is sending or receiving email.
@@
 In this section, we explore a few avenues of enquiry that might naturally arise.
-At best this only hints at the analytic power that interactive filtering and display can bring to bear on an email archive.
-To get a better appreciation, the user is encouraged to try it themselves at \texttt{rshiny.math.uwaterloo.ca/clinton}.
-The tool is useful both for generating and for testing hypotheses.
+At best these only hint at the analytic power that interactive filtering and display can bring to bear on an email archive.
+The user is encouraged to try it themselves at \texttt{rshiny.math.uwaterloo.ca/clinton}to get a fuller appreciation of the power of the tool.
+Those who are exploring the corpus will find it useful for generating hypotheses that might be tested against other sources.
+This is the approach taken here.
+Conversely, those more knowledgeable about the contexts surrounding Secretary Clinton and her private email server might use the visual analytic tool more as a means to test hypotheses already held.
 
 \subsection{The last two years}
-One might begin an interactive analysis by looking at the email patterns in the last two years of Clinton's tenure as Secretary of State.  
+We begin our interactive analysis by looking at  email patterns in the last two years of Clinton's tenure as Secretary of State.
+
 Figure~\ref{fig:lastTwoYears} 
@@
 The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.
-As can be seen, there are relatively few of these.
+As can be seen, there are \emph{relatively} few of these.
 
@@
 The next largest spikes appear at the far left of this time window, at the beginning of 2011.
-We might choose to compress the time window more, to focus on this section of the emails.
-
+We might choose to compress the time window more, to focus on the early 2011 section of the emails.
 Bringing the right end of the time slider in to April 15, 2011 produces the series of Figure~\ref{fig:early2011B1}.
@@
 \end{figure}
+
 As can be seen, there continue to be highs and lows of email activity.
-The highest peak now is at the far left.
+The highest peak now is at the left, though closer to the centre of Figure~\ref{fig:early2011B1} than it had been in Figure~\ref{fig:lastTwoYears} .
 This is January 29, 2011, the day after President Mubarak of Egypt has ordered the army into the streets of Cairo to quell protests \cite{cairoTanks2011, cairoResponse2011}.
-Dipping into the contents of the Wikileaks email archive, much of the email on this day is seen to be about managing the Egyptian Crisis.
+Dipping into the contents of the Wikileaks email archive, much of the email on this day is seen to be about managing this Egyptian crisis.
 
-Slightly to the right of this peak, there are several more.
+Just to the right of this peak, there are several more.
 At this time, international forces are gathering to consider Libya and its leader Muammar Gaddafi.
-Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on the Libyan crisis, we move the sliders in to focus on the time range from February 25 when Secretary Clinton announced the suspension of the Libyan embassy in Washington, to March 18, the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes.
-1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}.
+Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on the Libyan crisis, we move the sliders in again, this time to focus on the time range from February 25 (when Secretary Clinton announced the suspension of the Libyan embassy in Washington) to March 18 (the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes.
+1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}).
 On March 19, 2011 NATO military operations in Libya began \cite{LibyaTimelineWiki}.
@@
 \end{figure}
-The left most spike is the rightmost spike of approximately 80 emails from Figure~\ref{fig:early2011B1}.   
+The left most spike in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}  is the rightmost spike of the pair of approximately 80 emails from Figure~\ref{fig:early2011B1}.   
 Of course all other displays have been updating every time the time slider is adjusted.
@@
 Third, it is important that users realize that the value of metadata is not just in itself, where one might expect to easily discern general day-to-day habits such as one's circle of correspondents.
-Rather it is that when coupled with other sources, which are abundantly and publicly available for Senator Clinton, much more can be learned than from any one source alone.
+Rather it is that when coupled with other sources, which are abundantly and publicly available for Secretary Clinton, much more can be learned than from any one source alone.
 Those who have knowledge of, or access to, other sources may filter metadata to test previously held hypotheses; conversely, exploration of metadata could uncover patterns that generated hypotheses to be tested elsewhere.

### 2017-04-10 VAST_submission/vgtc_journal_latex/paper.tex
Cleaning up text

@@
 Not surprisingly,  Libya and its leader figure prominently in this restricted set of emails.  
-
 Perhaps more surprising is a term like ``windrush''.
@@
 Blair's electronic signature.
-Mr.
-Blair is part of this B1 selected correspondence.
+Mr. Blair is part of this B1 selected correspondence.
 
@@
 \label{sect:emailGap}
-Towards the very right of Figure~\ref{fig:lastTwoYears} there appears to be an anomaly in both time series.
-There is a noticeable flat area at zero in both series.
+Towards the very right of Figure~\ref{fig:lastTwoYears} there appears to be an anomaly in both time series.  There is a noticeable flat area at zero in both series.
+
 If we move the leftmost time sliders to the June 1, 2012, and the rightmost to the end, February 1, 2013, we have the email volumes for the last 8 months of the archive.
@@
 \end{figure}
-There appears to be no emails for a fairly large chunk of time.
-In fact, there are few or no emails in the archive from October 30 to November 9, 2012.
+There appears to be no emails for a fairly large chunk of time.  In fact, there are few or no emails in the archive from October 30 to November 9, 2012.  
 This seems curious, given that the archive is supposed to be complete in Secretary Clinton's work-related email during this time period.
+
 The block of no email appears as an even more unusual pattern in Figure~\ref{fig:teaser}(a).
@@
 
-One might ask what is happening around this time period?
-In a word, Benghazi.
+One might ask what is happening around this time period?  In a word, Benghazi.
+
 Five days after the assault, the U.S.
@@
 A month after the attack the administration was defending itself against charges that it deliberately downplayed a terrorist attack in Libya for political reasons in a presidential election year \cite{susanRiceWashPost2012}.
+
 In October, focus shifted to the security at Benghazi and whether it had been reduced prior to the attack  \cite{cbsSecurity}.
@@
 Embassy in Libya for additional security personnel had been turned down by the State Department \cite{BenghaziSecurityTestimony}.
+
 On October 26, CIA operatives who had defended the Benghazi mission against the terrorist attacks came forward with their story of what happened, on the ground.
@@
 
-To explore this possibility,  the time slider can be adjusted to a relatively small span, say of thirty days.
+To explore this possibility,  the time slider can be adjusted to a relatively small span, say of thirty days length.
 This would allow gaps in email of smaller sizes to be detected.
-To focus attention on Secretary Clinton's behaviour, only those emails sent from Secretary Clinton would be displayed.
+To focus attention on Secretary Clinton's behaviour, only those emails sent from Secretary Clinton will be displayed.
 Moving the slider, from left to right, from early 2009 to early 2013, allows one to move a one month wide magnifying glass across the whole of the archive.
@@
 This work was supported in part by
-a Discovery grant from the Natural Sciences and Engineering Research Council of Canada..}
+a Discovery grant from the Natural Sciences and Engineering Research Council of Canada.}
 

### 2017-04-10 VAST_submission/vgtc_journal_latex/paper.tex

@@
 Notably, in an April 10, 2011 email, Huma Abedin forwards Clinton the ``Stevens Update'' from Timmy Davis.
-The email describes Ambassador Stevens's concerns about the situation in Ajdabiyah (less than 150 km from Benghazi) where he is considering leaving Benghazi \cite{StevensUpdate}.
+The email describes Ambassador Stevens's concerns about the situation in Ajdabiyah (less than 150 km from Benghazi) and that Ambassador Stevens is considering leaving Benghazi \cite{StevensUpdate}.
 
@@
 Clinton had suffered a concussion ``\ldots in a fall brought on by an illness'' \cite{concussion} and was recovering at home by December 15.
-She had `already been forced to cancel a planned trip to the Middle East and north Africa'' \cite{concussion} that week; this is not the foreign travel shown in Figure~\ref{fig:ZeroEmailsDec2012} \cite{ForeignSched}.
+She had ``already been forced to cancel a planned trip to the Middle East and north Africa'' \cite{concussion} that week; this is not the foreign travel shown in Figure~\ref{fig:ZeroEmailsDec2012} \cite{ForeignSched}.
 
@@
 In other cases, the information turned up was possibly contentious and could be interpreted to suggest that the email was deliberately missing.
-We do not, and cannot, suggest that this is the case.
+We do not, and cannot, say that this is the case.
 Of course, the way in which the emails were selected to be handed over to the State Department and the subsequent erasure of all back up copies does little to dissuade anyone who might hold this view.
@@
 One of the combinations we turned up matched a patch we had identified ("Dec 2012"), the remaining did not.
-The complete sample (and what we found for each) was as follows:  ``June 2011''  (pro forma cable from SECSTATE that State Dept.
-employees change their passwords), ``Nov 2010''  (U.S.
-Embassy Cables appear on Wikileaks; Clinton rejects suggestion by Huma Abedin to start using  a government email account),   ``Dec 2012'' (documented in Section~ref{sect:youHaveNoMail}), ``March 2009'' (Clinton hits reset button with Russia), ``May 2011'' (Clinton delivers remarks on GlobalFood Security), ``March 2011'' (Clinton rejects talks with Gaddafi's sons; P.J.
+The complete sample (and what we found for each) was as follows:  ``June 2011''  (SECSTATE cables State Dept.
+employees to change their passwords), ``Nov 2010''  (U.S.
+Embassy Cables appear on WikiLeaks; Clinton rejects suggestion by Huma Abedin to start using  a government email account),   ``Dec 2012'' (documented in Section~\ref{sect:youHaveNoMail}), ``March 2009'' (Clinton hits reset button with Russia), ``May 2011'' (Clinton delivers remarks on Global Food Security), ``March 2011'' (Clinton rejects talks with Gaddafi's sons; P.J.
 Crowley forced to resign), ``May 2012'' (State Dept.
-investigating computer shipments to Iran), ``Dec 2011'' (Putin accuses Clinton of encouraging Russian protests), ``June 2012'' (Geneva I conference on Syria; Bill Clinton checked with State Dept.
+investigating computer shipments to Iran), ``Dec 2011'' (Putin accuses Clinton of encouraging Russian protests), ``June 2012'' (Geneva 1 conference on Syria; Bill Clinton checked with State Dept.
 on paid speech to group with Tehran ties; State Dept.
-Trafficking in Persons report), and  ``Oct 2011'' (Iranian plot to kill Saudi ambassador, State Dept.
+Trafficking in Persons report), and  ``Oct 2011'' (Iranian plot to kill Saudi ambassador; State Dept.
 cannot confirm Gaddafi's death).
-None but ``Dec 2012'' have a connection to the Benghazi controversy though two mention email.
+None but ``Dec 2012'' have a connection to the Benghazi controversy, though two mention email.
 
@@
 This might be grabbed as a B9  redaction code, if one is not careful.
-The same email record also has ``To'' and ``From'' header which is empty except for a date.
+The same email record also has ``To:'' and ``From:'' header which is empty except for a date.
 This header does not appear in the original pdf; rather the pdf begins with the next one inside the HTML.
@@
 
-First, metadata is often more reliable, regular, and, being easily collected, more generally available.
+First, metadata are often more reliable, regular, and, being easily collected, more generally available.
 For Secretary Clinton's emails,  some metadata is lost, either because their source is a printed form or because it has been redacted.

### 2017-04-11 app.R
Added link to paper

@@ -226,6 +226,9 @@ ui <- fluidPage(
                                a("Wikipedia", href = "https://en.wikipedia.org/wiki/Hillary_Clinton_email_controversy")),
                       fluidRow(id = "Context", h3("Useful Context"),
                                p(HTML("<a href='#Contents'>Back to table of contents</a>")),
+                               a("Interactive Filter and Display of Hillary Clinton's Emails: A Cautionary Tale of Metadata",
+                                 href = "https://www.researchgate.net/publication/315876309_Interactive_Filter_and_Display_of_Hillary_Clinton%27s_Emails_A_Cautionary_Tale_of_Metadata"),
+                               " | ",
                                a("Clinton's tenure as Secretary of State",
                                  href = "https://en.wikipedia.org/wiki/Hillary_Clinton%27s_tenure_as_Secretary_of_State"),
                                " | ",

### 2017-04-11 VAST_submission/vgtc_journal_latex/paper.tex

@@
 
-On March 16, 2016 Wikileaks  published a searchable archive \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton on her private server.
+On March 16, 2016 Wikileaks published a \href{https://wikileaks.org/clinton-emails/}{\color{blue}searchable archive} \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton on her private server.
 The documents were provided as pdfs by the U.S.
@@
 The State Dept.
-also provided a searchable web archive of the documents, released in several instalments from May 2015 to March 2017\cite{StateDeptFOIA}.
+also provided a \href{https://foia.state.gov/Search/Results.aspx?collection=Clinton_Email}{\color{blue}searchable web archive} of the documents, released in several instalments from May 2015 to March 2017\cite{StateDeptFOIA}.
 Both sites provide a  useful tool for anyone searching for particular terms in these documents.
@@
 One imagines, for example, having three browsers open simultaneously, one for the data base search (e.g.
-Wikileaks), one for statistical visualization and exploration (proposed here), and one for internet search (to provide important context for the other two services).
+\href{https://wikileaks.org/clinton-emails/}{\color{blue}WikiLeaks}), one for statistical visualization and exploration (\href{https://rshiny.math.uwaterloo.ca/clinton}{\color{blue} proposed here}), and one for \href{https://google.com}{\color {blue} internet search} (to provide important context for the other two services).
 Each fills a need not met by the other two; the synergy of all three provide the user a powerful set of investigative tools.
 
-In this paper, we describe an implementation of such a visual analytic service and illustrate it on the Secretary Clinton's e-mail archive.
-The implementation is located at  \texttt{rshiny.math.uwaterloo.ca/clinton} where any visitor may interactively explore summary features  of the entire corpus of emails.
+In this paper, we describe an implementation of such a visual analytic service and illustrate it on the Secretary Clinton's e-mail archive.  The implementation is located at
+\href{https://rshiny.math.uwaterloo.ca/clinton}{\color{blue} \texttt{rshiny.math.uwaterloo.ca/clinton} }
+where any visitor may interactively explore summary features  of the entire corpus of emails.
 Different visual displays present different salient features from all selected emails; emails can be selected via a variety of interactive filters.
@@
 Readers more familiar with the background may wish to skip to the next three sections where the implementation is described.
-Section \ref{sect:data} discusses the source of the data, its extraction and cleaning,  details of the metadata,  redaction codes as quasi-metadata,  and our limited use of the actual email content.
-Section \ref{sect:Displays} describes the displays used and illustrates them on the data.
+\autoref{sect:data} discusses the source of the data, its extraction and cleaning,  details of the metadata,  redaction codes as quasi-metadata,  and our limited use of the actual email content.
+Section \autoref{sect:Displays} describes the displays used and illustrates them on the data.
 These include a spoked network plot, an email volume time series plot, a scatterplot of daily sending times, a barplot of exemption codes, and term frequency displays.
-Section \ref{sect:Filters} presents the interactive filters that can be applied to change the displays.
+Section \autoref{sect:Filters} presents the interactive filters that can be applied to change the displays.
 With the implementation described, some interactive analyses of the Clinton email archive are conducted to give a quick overview of the utility of the service.
@@
 Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.
-In the first case Comey and the FBI were criticized by republican pundits and in the second case by democratic pundits.
+In the first case Comey and the FBI were criticized by Republican pundits and in the second case by Democrat pundits.
 
@@
 In cases where both \texttt{From} and \texttt{To} are entirely missing we do not impute values  (e.g.
-see \texttt{https://wikileaks.org/clinton-emails/emailid/31599}).
+see \href{https://wikileaks.org/clinton-emails/emailid/31599}{\color {blue} \texttt{https://wikileaks.org/clinton-emails/emailid/31599}}).
 
@@
 \end{figure}
-Figure~\ref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
+\autoref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
 Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.
@@
 Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.
-As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
+As \autoref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
 \subsection{Email volume}
 \label{sect:Displays:volume}
-Figure~\ref{fig:VolumeAll} 
+\autoref{fig:VolumeAll} 
 \begin{figure}[h]
@@
 In grey at top is the daily total; in magenta at bottom is the total for the subset selected by the filters.
-In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails that were sent \emph{from} Clinton.
+In \autoref{fig:VolumeAll}, the bottom series is the total daily emails that were sent \emph{from} Clinton.
 
-Over the four year scale shown in Figure~\ref{fig:VolumeAll}, the display is quite busy.
+Over the four year scale shown in \autoref{fig:VolumeAll}, the display is quite busy.
 Even so, a few features stand out.
@@
 \label{sect:Displays:times}
-Figure~\ref{fig:TimesAll}  
+\autoref{fig:TimesAll}  
 \begin{figure}[h]
@@
 
-Figure \ref{fig:FOIAcodes}
+\autoref{fig:FOIAcodes}
 \begin{figure}[h]
@@
 The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display.
-The top 20 shown in Figure~\ref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.
+The top 20 shown in \autoref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.
 
 The terms give some limited insight into the contents of the email.
-As seen in Figure~\ref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.
+As seen in \autoref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.
 One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies; this could magnify the within email frequency for some terms.
 
-As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is challenged by this messy and non-standard data.
+As mentioned in \autoref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is challenged by this messy and non-standard data.
 The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'.
@@
 \end{figure}
-Figure~\ref{fig:selectFOIA} displays the codes that are selected for email display.
+\autoref{fig:selectFOIA} displays the codes that are selected for email display.
 Any email having one or more of these codes will appear in the displays.
@@
 At best these only hint at the analytic power that interactive filtering and display can bring to bear on an email archive.
-The user is encouraged to try it themselves at \texttt{rshiny.math.uwaterloo.ca/clinton}to get a fuller appreciation of the power of the tool.
+The user is encouraged to try it themselves at \href{https://rshiny.math.uwaterloo.ca/clinton}{\color{blue} \texttt{rshiny.math.uwaterloo.ca/clinton} } to get a fuller appreciation of the power of the tool.
 Those who are exploring the corpus will find it useful for generating hypotheses that might be tested against other sources.
@@
 
-Figure~\ref{fig:lastTwoYears} 
+\autoref{fig:lastTwoYears} 
 \begin{figure}[h]
@@
 shows the email volume time series over this period. 
-This is accomplished by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.
+This is accomplished by simply having the right end of the time slider of \autoref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.
 Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.
@@
 
-Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).
+Recall from \autoref{sect:Displays:circle} that the tallest spike of \autoref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).
 The next largest spikes appear at the far left of this time window, at the beginning of 2011.
 We might choose to compress the time window more, to focus on the early 2011 section of the emails.
-Bringing the right end of the time slider in to April 15, 2011 produces the series of Figure~\ref{fig:early2011B1}.
+Bringing the right end of the time slider in to April 15, 2011 produces the series of \autoref{fig:early2011B1}.
 \begin{figure}[h]
@@
 As can be seen, there continue to be highs and lows of email activity.
-The highest peak now is at the left, though closer to the centre of Figure~\ref{fig:early2011B1} than it had been in Figure~\ref{fig:lastTwoYears} .
+The highest peak now is at the left, though closer to the centre of \autoref{fig:early2011B1} than it had been in \autoref{fig:lastTwoYears} .
 This is January 29, 2011, the day after President Mubarak of Egypt has ordered the army into the streets of Cairo to quell protests \cite{cairoTanks2011, cairoResponse2011}.
@@
 
-The readjusted time series are shown in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}.
+The readjusted time series are shown in \autoref{fig:EmailVolumeB1LibyaBuildup}.
 \begin{figure}[h]
@@
 \end{figure}
-The left most spike in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}  is the rightmost spike of the pair of approximately 80 emails from Figure~\ref{fig:early2011B1}.   
+The left most spike in \autoref{fig:EmailVolumeB1LibyaBuildup}  is the rightmost spike of the pair of approximately 80 emails from \autoref{fig:early2011B1}.   
 Of course all other displays have been updating every time the time slider is adjusted.
 
-Some sense of the content of the emails can be had from the updated term frequency displays shown in Figure~\ref{fig:TermFreqB1LibyaBuildup}.
+Some sense of the content of the emails can be had from the updated term frequency displays shown in \autoref{fig:TermFreqB1LibyaBuildup}.
 \begin{figure}[h]
@@
 
-Other restrictions, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
+Other restrictions, as measured by the FOIA exemption codes, can be seen in \autoref{fig:FOIAB12011LibyaBuildup}.
 \begin{figure}[h]
@@
 
-We might now consider the composition of the inner circle for this set of emails.  Figure~\ref{fig:InnerCircle2011B1LibyaBuildup} 
+We might now consider the composition of the inner circle for this set of emails.  \autoref{fig:InnerCircle2011B1LibyaBuildup} 
 \begin{figure}[h]
@@
 Prime Minister Tony Blair; this is the unidentifiable domain account \texttt{aclb}.
-It should be noted that there could be other accounts which do not appear in the inner circle because their emails were not present in the email record.  See Section~\ref{sect:data:metadata}.
+It should be noted that there could be other accounts which do not appear in the inner circle because their emails were not present in the email record.  See \autoref{sect:data:metadata}.
 
@@
 We might also more closely connect email dates and correspondents with world events.
-Instead, we now turn our attention to another feature which appears in Figure~\ref{fig:lastTwoYears}.
+Instead, we now turn our attention to another feature which appears in \autoref{fig:lastTwoYears}.
 
@@
 \label{sect:emailGap}
-Towards the very right of Figure~\ref{fig:lastTwoYears} there appears to be an anomaly in both time series.  There is a noticeable flat area at zero in both series.
+Towards the very right of \autoref{fig:lastTwoYears} there appears to be an anomaly in both time series.  There is a noticeable flat area at zero in both series.
 
 If we move the leftmost time sliders to the June 1, 2012, and the rightmost to the end, February 1, 2013, we have the email volumes for the last 8 months of the archive.
-The resulting series are shown in Figure~\ref{fig:EmailVolumeB1Jun1On}.
+The resulting series are shown in \autoref{fig:EmailVolumeB1Jun1On}.
 \begin{figure}[h]
@@
 
-The block of no email appears as an even more unusual pattern in Figure~\ref{fig:teaser}(a).
+The block of no email appears as an even more unusual pattern in \autoref{fig:teaser}(a).
 There the block is conspicuous by the absence amongst the emails from September 11 to November 23, 2012.
-The email on which the displays of Figure~\ref{fig:teaser} are based, is no longer filtered by FOIA code.
+The email on which the displays of \autoref{fig:teaser} are based, is no longer filtered by FOIA code.
 It contains all email in the archive in this time period.
-Direct examination of the archives at Wikileaks and the State Department show no mail in this empty block beyond that seen in Figure~\ref{fig:teaser}(a).
+Direct examination of the archives at Wikileaks and the State Department show no mail in this empty block beyond that seen in \autoref{fig:teaser}(a).
 
-The inner circle for all emails from September 11 to November 23, 2012 is shown in Figure~\ref{fig:teaser}(b).
+The inner circle for all emails from September 11 to November 23, 2012 is shown in \autoref{fig:teaser}(b).
 Those closest to Clinton in terms of volume of correspondence are Clinton aides Jake Sullivan, Cheryl Mills, and Monica Hanley.
@@
 
-Moving the time sliders to cover either side of the email gap, October 22 to November 17, 2012, results in  the term frequency displays of Figure~\ref{fig:TermFreqOct22Nov17_2012}.
+Moving the time sliders to cover either side of the email gap, October 22 to November 17, 2012, results in  the term frequency displays of \autoref{fig:TermFreqOct22Nov17_2012}.
 \begin{figure}[h]
@@
 Following this approach, ten 
-more patches can be found (in addition to that of Section~\ref{sect:emailGap}).
-With the dates in hand, an internet search can be conducted to see what, if anything, might turn up to explain the patch.
+more patches can be found (in addition to that of \autoref{sect:emailGap}).  With the dates in hand, an internet search can be conducted to see what, if anything, might turn up to explain the patch.
 %\subsubsection{You have no mail \ldots}
@@
 The first, and largest, patch occurs at the very beginning of Clinton's tenure as Secretary of State.
-It is  obvious even from the full time series of Figure~\ref{fig:VolumeAll}.
+It is  obvious even from the full time series of \autoref{fig:VolumeAll}.
 Essentially no email from Secretary Clinton, and little from anyone else appears until mid to late April 2009.
@@
 This coincides with Secretary Clinton's State visits to India and Thailand \cite{ForeignSched}.
-Similarly, the absence of email from Clinton from October 11 to  October 13, 2009 corresponds to a State visit to Russia as shown in  Figure~\ref{fig:ClintonZerosOctober2009AndForeignSchedule} \cite{ForeignSched, visitRussia}.
+Similarly, the absence of email from Clinton from October 11 to  October 13, 2009 corresponds to a State visit to Russia as shown in  \autoref{fig:ClintonZerosOctober2009AndForeignSchedule} \cite{ForeignSched, visitRussia}.
 \begin{figure}[h]
@@
 The last two patches appear 
-shortly after the email gap from Section~\ref{sect:emailGap}.  As seen in Figure~\ref{fig:ZeroEmailsDec2012}
+shortly after the email gap from \autoref{sect:emailGap}.  As seen in \autoref{fig:ZeroEmailsDec2012}
 \begin{figure}[h]
@@
 Clinton had suffered a concussion ``\ldots in a fall brought on by an illness'' \cite{concussion} and was recovering at home by December 15.
-She had ``already been forced to cancel a planned trip to the Middle East and north Africa'' \cite{concussion} that week; this is not the foreign travel shown in Figure~\ref{fig:ZeroEmailsDec2012} \cite{ForeignSched}.
+She had ``already been forced to cancel a planned trip to the Middle East and north Africa'' \cite{concussion} that week; this is not the foreign travel shown in \autoref{fig:ZeroEmailsDec2012} \cite{ForeignSched}.
 
@@
 For every patch where Clinton sent no email, we deliberately used a very coarse query to the Google search engine.
-Figure~\ref{fig:googleSearch}
+\autoref{fig:googleSearch}
 \begin{figure}[h]
@@
 
-As seen in Section~\ref{sect:youHaveNoMail} this was often sufficient to turn up information that explained the lack of email as naturally occurring due to injury or travel.
+As seen in \autoref{sect:youHaveNoMail} this was often sufficient to turn up information that explained the lack of email as naturally occurring due to injury or travel.
 In other cases, the information turned up was possibly contentious and could be interpreted to suggest that the email was deliberately missing.
@@
 employees to change their passwords), ``Nov 2010''  (U.S.
-Embassy Cables appear on WikiLeaks; Clinton rejects suggestion by Huma Abedin to start using  a government email account),   ``Dec 2012'' (documented in Section~\ref{sect:youHaveNoMail}), ``March 2009'' (Clinton hits reset button with Russia), ``May 2011'' (Clinton delivers remarks on Global Food Security), ``March 2011'' (Clinton rejects talks with Gaddafi's sons; P.J.
+Embassy Cables appear on WikiLeaks; Clinton rejects suggestion by Huma Abedin to start using  a government email account),   ``Dec 2012'' (documented in \autoref{sect:youHaveNoMail}), ``March 2009'' (Clinton hits reset button with Russia), ``May 2011'' (Clinton delivers remarks on Global Food Security), ``March 2011'' (Clinton rejects talks with Gaddafi's sons; P.J.
 Crowley forced to resign), ``May 2012'' (State Dept.
@@
 
-\bibliographystyle{abbrv}
+%\bibliographystyle{abbrv}
 %\bibliographystyle{abbrv-doi}
@@
 %\bibliographystyle{abbrv-doi-hyperref}
-%\bibliographystyle{abbrv-doi-hyperref-narrow}
+\bibliographystyle{abbrv-doi-hyperref-narrow}
 

### 2017-04-12 app.R
Changed ordering of links

@@ -202,6 +202,8 @@ ui <- fluidPage(
      column(width = 4,
             wellPanel(id = "WriteUp", 
                       style = "overflow-y:scroll; max-height: 800px",
+                      fluidRow(a("Interactive Filter and Display of Hillary Clinton's Emails: A Cautionary Tale of Metadata",
+                                 href = "https://www.researchgate.net/publication/315876309_Interactive_Filter_and_Display_of_Hillary_Clinton%27s_Emails_A_Cautionary_Tale_of_Metadata")),
                       fluidRow(id = "Contents", h3("Table of Contents"),
                                p(HTML("<a href='#KeyPlayers'>Wikipedia Articles of Key Players</a>")),
                                p(HTML("<a href='#Timeline'>Timeline Links</a>")),
@@ -226,9 +228,6 @@ ui <- fluidPage(
                                a("Wikipedia", href = "https://en.wikipedia.org/wiki/Hillary_Clinton_email_controversy")),
                       fluidRow(id = "Context", h3("Useful Context"),
                                p(HTML("<a href='#Contents'>Back to table of contents</a>")),
-                               a("Interactive Filter and Display of Hillary Clinton's Emails: A Cautionary Tale of Metadata",
-                                 href = "https://www.researchgate.net/publication/315876309_Interactive_Filter_and_Display_of_Hillary_Clinton%27s_Emails_A_Cautionary_Tale_of_Metadata"),
-                               " | ",
                                a("Clinton's tenure as Secretary of State",
                                  href = "https://en.wikipedia.org/wiki/Hillary_Clinton%27s_tenure_as_Secretary_of_State"),
                                " | ",

### 2017-04-13 app.R
Changed layout

@@ -185,7 +185,9 @@ ui <- fluidPage(
 
    # Application title
    titlePanel(HTML("Secretary Clinton's Email (Source: <a href = 'https://wikileaks.org/clinton-emails/'>Wikileaks</a>)")),
-   fluidRow(column(h4("Christopher D. Salahub and R. Wayne Oldford"), offset = 0.2, width = 12)),
+   fluidRow(column(h4("Christopher D. Salahub and R. Wayne Oldford:", a("Interactive Filter and Display of Hillary Clinton's Emails: A Cautionary Tale of Metadata",
+                                                                        href = "https://www.researchgate.net/publication/315876309_Interactive_Filter_and_Display_of_Hillary_Clinton%27s_Emails_A_Cautionary_Tale_of_Metadata")),
+                   offset = 0.2, width = 12)),
 
    # the slider right below the title to make it as long as possible
    fluidRow(column(width = 2, offset = 0.5, h4("Date Range:"))),
@@ -202,46 +204,39 @@ ui <- fluidPage(
      column(width = 4,
             wellPanel(id = "WriteUp", 
                       style = "overflow-y:scroll; max-height: 800px",
-                      fluidRow(a("Interactive Filter and Display of Hillary Clinton's Emails: A Cautionary Tale of Metadata",
-                                 href = "https://www.researchgate.net/publication/315876309_Interactive_Filter_and_Display_of_Hillary_Clinton%27s_Emails_A_Cautionary_Tale_of_Metadata")),
-                      fluidRow(id = "Contents", h3("Table of Contents"),
-                               p(HTML("<a href='#KeyPlayers'>Wikipedia Articles of Key Players</a>")),
-                               p(HTML("<a href='#Timeline'>Timeline Links</a>")),
-                               p(HTML("<a href='#Context'>Context Links</a>")),
-                               p(HTML("<a href='#App'>Service Description</a>")),
-                               p(HTML("<a href='#Data'>Data</a>")),
-                               p(HTML("<a href='#Analysis'>Some Interesting Analyses</a>"))),
-                      fluidRow(id = "KeyPlayers", 
-                               h3("Wikipedia Articles of Key Players"),
-                               p(HTML("<a href='#Contents'>Back to table of contents</a>")),
-                               p(HTML(paste(paste(links, collapse = " | "),
-                                            "<a href='https://en.wikipedia.org/wiki/Madeleine_Albright'>Madeleine Albright ('pathfinder')</a>",
[…]

### 2017-04-16 app.R
Added scaling feature, investigated TFIDF

@@ -338,6 +338,7 @@ ui <- fluidPage(
      column(width = 2,
             fluidRow(selectInput("AdjVar", "Daily Reference Point: ", c("0:00", "18:00"),
                                  selected = "0:00")),
+            fluidRow(checkboxInput("SelScale", "Scale by selected")),
             fluidRow(checkboxInput("Schedule", "Display Foreign Travel Schedule")),
             fluidRow(selectInput("ToFromFilter", "Show Emails: ", 
                                  c("From Clinton", "To Clinton", "All Emails"),
@@ -348,7 +349,7 @@ ui <- fluidPage(
                                  selected = c("B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", 
                                               "B9", "None"),
                                  multiple = TRUE)),
-            fluidRow(h4("20 Highest", a("TF_IDF", href = "https://en.wikipedia.org/wiki/Tf%E2%80%93idf"),
+            fluidRow(h4("20 Highest", a("TF-IDF", href = "https://en.wikipedia.org/wiki/Tf%E2%80%93idf"),
                         "Terms")),
             fluidRow(textOutput("tfidf")),
             fluidRow(h4("20 Highest Frequency Terms")),
@@ -398,6 +399,7 @@ server <- function(input, output) {
                                           StartDate < input$range[2]) |
                                (EndDate < input$range[2] & EndDate >= input$range[1]))
     intermed$dispSched <- input$Schedule
+    intermed$ScaleSel <- input$SelScale
     intermed$adjVar <- input$AdjVar == "18:00"
     intermed$ZeroSel <- length(intermed$selIDs) == 0
     return(intermed)
@@ -438,10 +440,12 @@ server <- function(input, output) {
    })
    # next display the time series of emails sent by day
    output$DaySum <- renderPlot({
+     if (inter()$ScaleSel) yrange <- extendrange(inter()$selCounts)
+     else yrange <- extendrange(AScounts)
      plot(x = inter()$selDays, y = inter()$selCounts, type = 'l', xlab = 'Date (dd/mm/yy)',
           xaxt = "n", ylab = 'Number of Emails', pch = 19,
           main = "Number of Emails by Date", sub = inter()$DateRange,
-          ylim = extendrange(AScounts), col = adjustcolor("black", alpha.f = 0.6))
+          ylim = yrange, col = adjustcolor("black", alpha.f = 0.6))
      # add the schedule if it has been selected
      if (inter()$dispSched) {
        apply(inter()$Sched, 1,

### 2017-04-17 app.R
Updated links

@@ -231,6 +231,8 @@ ui <- fluidPage(
                                  HTML(paste(paste(links, collapse = " | "),
                                             "<a href='https://en.wikipedia.org/wiki/Madeleine_Albright'>Madeleine Albright ('pathfinder')</a>",
                                             "<a href='https://en.wikipedia.org/wiki/Tony_Blair'>Tony Blair ('aclb')</a>",
+                                            "<a href='https://www.google.com/search?q=Monica+Hanley'>Monica Hanley</a>",
+                                            "<a href='https://www.google.com/search?q=Lauren+Jiloty'>Lauren Jiloty</a>",
                                             sep = " | ")))),
                       fluidRow(id = "Analysis", 
                                h3("Analysis"),

### 2017-04-17 VAST_submission/vgtc_journal_latex/paper.tex
updated the figures to correspond to the varying spoke lengths

@@
   \begin{tabular}{cc}
-  \includegraphics[height=0.22\textheight]{EmailSentRecdSept11toNov23} &
-  \includegraphics[height=0.22\textheight]{InnerCircleSept11ToNov23} \\
+  \includegraphics[height=0.2\textheight]{EmailSentRecdSept11toNov23} &
+  \includegraphics[height=0.2\textheight]{InnerCircleSept11ToNov23} \\
   &\\
@@
 The width of each spoke is proportional to the volume of emails between correspondents.  
-Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.
+Similarly, the length of each spoke is inversely proportional to the volume of email so 
+%Whenever the difference in volume is great enough,  correspondents are separated into two groups: 
+those correspondents having the greatest volume will 
+%with the greatest correspondence 
+have shorter spokes visually placing them ``closer'' to Clinton.  
 As \autoref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
@@
 We begin our interactive analysis by looking at  email patterns in the last two years of Clinton's tenure as Secretary of State.  
-
 \autoref{fig:lastTwoYears} 
@@
 shows the email volume time series over this period.
+
 This is accomplished by simply having the right end of the time slider of \autoref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.

### 2017-04-19 EmailFunctions.R
Added function calls

@@ -193,6 +193,9 @@ get_Clin_emails <- function(ids = 1:32795, # give possible emails for selection
   return(FullData)
 }
 
+# the function call
+# get_Clin_emails(contAtErr = TRUE)
+
 ### Pre-loaded data #####################################################################
 EmailData <- read.csv('ClintonEmailData.csv')
 
